---
title: "Characteristic-function goodness-of-fit tests"
output: pdf_document
date: "2026-07-19"
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)

library(tensormodels)
```


```{r}
cftnorm <- function(freq, mu, sigmas) {
  num_freq <- n_draws(freq)
  
  res <- complex(num_freq)
  
  for(i in 1:num_freq) {
    # pull_draw() keeps the tensor dimensions even when there is only one
    # frequency.
    curr_freq <- pull_draw(freq, i)
    
    second_term <- curr_freq
  
    for(d in 1:length(dim(curr_freq))) {
      second_term <- n_prod(second_term, sigmas[[d]], d)
    }
  
    first_term <- 1i * sum(curr_freq * mu)
  
    res[i] <- exp(first_term - 1/2 * sum(curr_freq * second_term))
  }
  res
}
```


```{r}
tensor_ecf <- function(draws, freq) {
  num_draws <- n_draws(draws)
  num_freq <- n_draws(freq)
  nstar <- prod(draw_shape(draws))

  #each row is a vectorized tensor obs
  draw_matrix <- matrix(unclass(draws),
                        nrow = num_draws,
                        ncol = nstar)

  empirical_cf <- complex(num_freq)

  for (j in 1:num_freq) { #get empirical cf for each freq
    # pull_draw() works consistently for one or many frequencies.
    curr_freq <- pull_draw(freq, j)

    #compute 1 inner prod for each obs
    phases <- drop(draw_matrix %*% c(curr_freq))

    #average cf
    empirical_cf[j] <- mean(cos(phases)) + 1i * mean(sin(phases))
  }

  empirical_cf
}
```


```{r}
norm_draws <- rtnorm(n = 1e3, mu = array(1:24, dim = c(2, 3, 4)))

mle_norm <- tensor_mle(norm_draws, model = "normal")

freq <- tensor(data = rnorm(48), dim = c(2, 3, 4), n = 2)

cftnorm(freq = freq,
        mu = mle_norm$mu, 
        sigmas = mle_norm$sigmas)

tensor_ecf(norm_draws, freq)
```


```{r}
set.seed(123)

make_reference_freq <- function(tensor_dim,
                                num_per_radius = 50,
                                seed = 123) {

  set.seed(seed)

  p <- prod(tensor_dim)

  radii <- c(0.5, 1, 2, 3)

  number_frequencies <-
    length(radii) * num_per_radius

  frequency_values <-
    numeric(number_frequencies * p)

  frequency_number <- 0

  for (radius in radii) {

    for (j in 1:num_per_radius) {

      frequency_number <-
        frequency_number + 1

      # Generate a direction
      direction <- rnorm(p)

      # Make its length exactly one
      direction <-
        direction /
        sqrt(sum(direction^2))

      # Give it the desired distance from zero
      current_frequency <-
        radius * direction

      # Store this frequency
      first_position <-
        (frequency_number - 1) * p + 1

      last_position <-
        frequency_number * p

      frequency_values[
        first_position:last_position
      ] <- current_frequency
    }
  }

  tensor(
    data = frequency_values,
    dim = tensor_dim,
    n = number_frequencies
  )
}
```


```{r raw-scale-cf-statistic}
# Generate frequencies on a reference scale. These have norms
# 0.5, 1, 2, and 3, regardless of the dimensions of a tensor draw.
reference_freq <- make_reference_freq(
  tensor_dim = draw_shape(norm_draws),
  num_per_radius = 50,
  seed = 123
)

# Keep the observations on their original scale. Instead, transform only the
# frequencies so that they account for the fitted mode covariance matrices.
freq <- reference_freq

for (d in seq_along(mle_norm$sigmas)) {
  frequency_transform <- solve(chol(mle_norm$sigmas[[d]]))

  freq <- n_prod(
    freq,
    frequency_transform,
    d
  )
}

# Compare the theoretical and empirical CFs at the transformed frequencies.
cf_tibble <- tibble::tibble(
  freq = freq,
  cf_theory = cftnorm(
    freq,
    mu = mle_norm$mu,
    sigmas = mle_norm$sigmas
  ),
  cf_emp = tensor_ecf(
    norm_draws,
    freq
  ),
  diff = Mod(cf_theory - cf_emp)
)

# The test statistic is n times the mean SQUARED CF difference.
stat <- n_draws(norm_draws) * mean(cf_tibble$diff^2)

stat
```


```{r}
L <- n_draws(freq)

# Each frequency contributes one cosine feature and one sine feature.
number_features <- 2 * L

# The means of the cosine features are Re(phi), and the means of the sine
# features are Im(phi).
phi <- cf_tibble$cf_theory
feature_mean <- c(Re(phi), Im(phi))

# This matrix will contain E[feature_a * feature_b].
second_moment <- matrix(
  0,
  nrow = number_features,
  ncol = number_features
)

for (j in 1:L) {
  freq_j <- pull_draw(freq, j)

  for (k in 1:L) {
    freq_k <- pull_draw(freq, k)

    freq_plus <- freq_j + freq_k
    freq_j_minus_k <- freq_j - freq_k
    freq_k_minus_j <- freq_k - freq_j

    # The product-to-sum formulas require the fitted CF at the sum and
    # differences of each pair of frequencies.
    phi_plus <- cftnorm(
      freq_plus,
      mu = mle_norm$mu,
      sigmas = mle_norm$sigmas
    )
    phi_j_minus_k <- cftnorm(
      freq_j_minus_k,
      mu = mle_norm$mu,
      sigmas = mle_norm$sigmas
    )
    phi_k_minus_j <- cftnorm(
      freq_k_minus_j,
      mu = mle_norm$mu,
      sigmas = mle_norm$sigmas
    )

    # E[cos(t_j'X) cos(t_k'X)]
    second_moment[j, k] <-
      Re(phi_j_minus_k + phi_plus) / 2

    # E[sin(t_j'X) sin(t_k'X)]
    second_moment[L + j, L + k] <-
      Re(phi_j_minus_k - phi_plus) / 2

    # E[cos(t_j'X) sin(t_k'X)]
    second_moment[j, L + k] <-
      Im(phi_plus + phi_k_minus_j) / 2

    # The second-moment matrix is symmetric.
    second_moment[L + k, j] <-
      second_moment[j, L + k]
  }
}

# Convert second moments to covariances:
# Cov(A, B) = E[A * B] - E[A] * E[B].
feature_covariance <- matrix(
  0,
  nrow = number_features,
  ncol = number_features
)

for (a in 1:number_features) {
  for (b in 1:number_features) {
    feature_covariance[a, b] <-
      second_moment[a, b] -
      feature_mean[a] * feature_mean[b]
  }
}

# Remove tiny numerical asymmetry before calling eigen().
feature_covariance <-
  (feature_covariance + t(feature_covariance)) / 2

eigen_result <- eigen(
  feature_covariance / L,
  symmetric = TRUE,
  only.values = TRUE
)

all_lambda <- eigen_result$values

cutoff <- max(all_lambda) * 1e-10

lambda <- all_lambda[
  all_lambda > cutoff
]
```


```{r vectorized-normal-feature-covariance}
# This is a faster, vectorized version of the preceding covariance calculation.
# It avoids constructing t_j + t_k and t_j - t_k with nested R loops.

L <- n_draws(freq)
p <- prod(draw_shape(freq))

# Rows are frequencies and columns are tensor entries.
reference_matrix <- matrix(
  unclass(reference_freq),
  nrow = L,
  ncol = p
)

raw_frequency_matrix <- matrix(
  unclass(freq),
  nrow = L,
  ncol = p
)

# Squared norm of each reference frequency.
squared_norm <- rowSums(reference_matrix^2)

# Entry (j, k) is the inner product of reference frequencies j and k.
frequency_inner_products <- tcrossprod(reference_matrix)

# Entry (j, k) is ||s_j||^2 + ||s_k||^2.
pair_squared_norm <- outer(
  squared_norm,
  squared_norm,
  "+"
)

# Calculate ||s_j + s_k||^2 and ||s_j - s_k||^2 for every pair.
quadratic_plus <-
  pair_squared_norm +
  2 * frequency_inner_products

quadratic_minus <-
  pair_squared_norm -
  2 * frequency_inner_products

# Remove tiny negative values caused by floating-point rounding.
quadratic_minus <- pmax(quadratic_minus, 0)

# The raw-scale normal CF also contains t' * mu.
mu_vector <- c(mle_norm$mu)

location <- drop(
  raw_frequency_matrix %*% mu_vector
)

location_plus <- outer(
  location,
  location,
  "+"
)

location_minus <- outer(
  location,
  location,
  "-"
)

# Evaluate phi(t_j + t_k) and phi(t_j - t_k) for every pair at once.
phi_plus <- exp(
  1i * location_plus -
  0.5 * quadratic_plus
)

phi_minus <- exp(
  1i * location_minus -
  0.5 * quadratic_minus
)

# Use the product-to-sum identities to form all second moments.
cos_cos_second_moment <-
  (Re(phi_minus) + Re(phi_plus)) / 2

sin_sin_second_moment <-
  (Re(phi_minus) - Re(phi_plus)) / 2

# phi(t_k - t_j) is Conj(phi(t_j - t_k)), which gives the minus sign.
cos_sin_second_moment <-
  (Im(phi_plus) - Im(phi_minus)) / 2

second_moment_vectorized <- rbind(
  cbind(
    cos_cos_second_moment,
    cos_sin_second_moment
  ),
  cbind(
    t(cos_sin_second_moment),
    sin_sin_second_moment
  )
)

# Means of the original cosine and sine features.
phi_original <- exp(
  1i * location -
  0.5 * squared_norm
)

feature_mean_vectorized <- c(
  Re(phi_original),
  Im(phi_original)
)

# Cov(A, B) = E[A * B] - E[A] * E[B].
feature_covariance_vectorized <-
  second_moment_vectorized -
  tcrossprod(feature_mean_vectorized)

feature_covariance_vectorized <-
  (
    feature_covariance_vectorized +
    t(feature_covariance_vectorized)
  ) / 2

eigen_result_vectorized <- eigen(
  feature_covariance_vectorized / L,
  symmetric = TRUE,
  only.values = TRUE
)

all_lambda_vectorized <-
  eigen_result_vectorized$values

cutoff_vectorized <-
  max(all_lambda_vectorized) * 1e-10

lambda_vectorized <- all_lambda_vectorized[
  all_lambda_vectorized > cutoff_vectorized
]

# These should be close to zero when the loop and vectorized calculations agree.
max_covariance_difference <- max(
  abs(
    feature_covariance_vectorized -
    feature_covariance
  )
)

max_eigenvalue_difference <- max(
  abs(
    all_lambda_vectorized -
    all_lambda
  )
)

c(
  max_covariance_difference = max_covariance_difference,
  max_eigenvalue_difference = max_eigenvalue_difference
)

# Use the faster result in the null-distribution calculation below.
feature_covariance <- feature_covariance_vectorized
lambda <- lambda_vectorized
```


```{r}
set.seed(123)

number_simulations <- 10000

null_statistics <- numeric(
  number_simulations
)

for (j in 1:length(lambda)) {

  null_statistics <-
    null_statistics +
    lambda[j] *
    rchisq(
      number_simulations,
      df = 1
    )
}

ggplot() +
  geom_histogram(aes(x = null_statistics), fill = "pink", color = "black", bins = 80) +
  geom_vline(xintercept = stat, lty = 2, color = "red") +
  theme_minimal()

mean(stat <= null_statistics)
```


```{r}
eigen(mle_norm$sigmas[[1]])
```


```{r}
standardize <- function(draws) {
  mle_draws <- tensor_mle(draws, model = "normal")
  
  #standardize draws with mean and covar
  standard_draws <- draws - mle_draws$mu
  
  num_sigmas <- length(draw_shape(draws))
  
  for(i in 1:num_sigmas) {
    sigma_i <- mle_draws$sigmas[[i]]
    inv_sigma <- solve(t(chol(sigma_i)))
    
    standard_draws <- n_prod(standard_draws, inv_sigma, i)
  }
  standard_draws
}
```


```{r}
n <- n_draws(standard_draws)
p <- prod(draw_shape(standard_draws))

# Rows are draws; columns are tensor entries.
draw_matrix <- matrix(
  unclass(standard_draws),
  nrow = n,
  ncol = p
)

# Choose one frequency tensor.
set.seed(123)

frequency <- array(
  rnorm(p, sd = 1 / sqrt(p)),
  dim = draw_shape(standard_draws)
)

frequency_vector <- as.numeric(frequency)

# One tensor inner product for each draw.
phases <- drop(draw_matrix %*% frequency_vector)

# Empirical characteristic function at this frequency.
empirical_cf <-
  mean(cos(phases)) +
  1i * mean(sin(phases))

empirical_cf
```


```{r}
theoretical_cf <- complex(num_freq)

for (j in seq_len(num_freq)) {
  frequency <- frequencies[j, ]

  phases <- drop(draw_matrix %*% frequency)

  empirical_cf[j] <-
    mean(cos(phases)) +
    1i * mean(sin(phases))

  theoretical_cf[j] <-
    exp(-sum(frequency^2) / 2)
}
```


```{r}
n_frequencies <- 1e3

frequencies <- matrix(
  rnorm(n_frequencies * nstar, sd = 1 / sqrt(nstar)),
  nrow = n_frequencies
)


#empirical CF
phases <- standard_draws %*% frequency
empirical_cf <- mean(exp(1i * phases))

#theoretical CF
```

# A simple tensor characteristic-function goodness-of-fit test

The function below compares the empirical characteristic function with the
characteristic function of a fitted tensor-variate normal or tensor-variate
skewed $t$ model. Its statistic is

\[
T_n = \frac{n}{L}\sum_{j=1}^L
\left|\widehat\varphi_n(t_j)-\varphi(t_j;\widehat\theta)\right|^2,
\]

where $L$ is the number of randomly selected frequency vectors. The p-value
is computed with a parametric bootstrap. Each bootstrap sample is re-fitted,
which accounts for using the same data to estimate the model and assess its
fit. A small `B` is useful while experimenting; use at least `B = 999` for a
reported analysis.

```{r simple-gof-test}
# Return matrices W_k for which W_k Sigma_k W_k' = I.
# Applying these matrices mode by mode standardizes the covariance without
# constructing a potentially enormous Kronecker covariance matrix.
.gof_whitening_matrices <- function(sigmas) {
  whitening <- vector("list", length(sigmas))

  for (k in seq_along(sigmas)) {
    sigma_k <- as.matrix(sigmas[[k]])
    sigma_k <- (sigma_k + t(sigma_k)) / 2
    chol_k <- chol(sigma_k)
    whitening[[k]] <- solve(t(chol_k))
  }

  whitening
}

# Center and whiten every tensor draw, then place the vectorized draws in an
# n by p matrix. Here p is the number of entries in one tensor draw.
.gof_standardize_draws <- function(draws, fit) {
  n <- n_draws(draws)
  p <- prod(draw_shape(draws))
  whitening <- .gof_whitening_matrices(fit$sigmas)
  standardized <- matrix(0, nrow = n, ncol = p)

  for (i in seq_len(n)) {
    current_draw <- pull_draw(draws, i) - fit$mu

    for (k in seq_along(whitening)) {
      current_draw <- n_prod(current_draw, whitening[[k]], k)
    }

    standardized[i, ] <- as.numeric(unclass(current_draw))
  }

  standardized
}

# Characteristic function of the standardized skewed t distribution.
# If W ~ InvGamma(nu/2, nu/2), then the standardized model is
# Z = W * a + sqrt(W) * epsilon, where epsilon ~ N(0, I).
.gof_skewt_cf <- function(frequencies, whitened_skew, nu) {
  number_frequencies <- nrow(frequencies)
  cf_values <- complex(number_frequencies)
  a <- nu / 2
  b <- nu / 2

  for (j in seq_len(number_frequencies)) {
    frequency <- frequencies[j, ]
    q <- sum(frequency^2)
    r <- sum(frequency * whitened_skew)
    s <- q / 2 - 1i * r

    # The CF equals one at the zero frequency.
    if (Mod(s) < .Machine$double.eps^0.5) {
      cf_values[j] <- 1 + 0i
    } else {
      # E[exp(-s W)] for an inverse-gamma W. The exponentially scaled
      # Bessel function is used to reduce numerical underflow.
      bessel_argument <- 2 * sqrt(b * s)
      log_cf <- log(2) +
        (a / 2) * (log(b) + log(s)) -
        lgamma(a) +
        log(Bessel::BesselK(
          bessel_argument,
          nu = a,
          expon.scaled = TRUE
        )) -
        bessel_argument

      cf_values[j] <- exp(log_cf)
    }
  }

  cf_values
}

# Create a function that evaluates the fitted standardized model CF. For the
# skewed t model, the skew tensor is whitened once when this function is made.
.gof_make_model_cf <- function(model, fit) {
  if (model == "normal") {
    function(frequencies) {
      number_frequencies <- nrow(frequencies)
      cf_values <- complex(number_frequencies)

      for (j in seq_len(number_frequencies)) {
        cf_values[j] <- exp(-sum(frequencies[j, ]^2) / 2)
      }

      cf_values
    }
  } else {
    whitening <- .gof_whitening_matrices(fit$sigmas)
    whitened_skew <- fit$skew

    for (k in seq_along(whitening)) {
      whitened_skew <- n_prod(whitened_skew, whitening[[k]], k)
    }

    whitened_skew <- as.numeric(whitened_skew)

    function(frequencies) {
      .gof_skewt_cf(
        frequencies = frequencies,
        whitened_skew = whitened_skew,
        nu = fit$nu
      )
    }
  }
}

# Compute T_n for one data set and an already fitted model.
.gof_cf_statistic <- function(draws, model, fit, frequencies) {
  standardized <- .gof_standardize_draws(draws, fit)
  number_frequencies <- nrow(frequencies)
  empirical_cf <- complex(number_frequencies)

  # Compute the empirical CF separately at each frequency.
  for (j in seq_len(number_frequencies)) {
    phases <- standardized %*% frequencies[j, ]
    empirical_cf[j] <- mean(exp(1i * phases))
  }

  model_cf <- .gof_make_model_cf(model, fit)
  theoretical_cf <- model_cf(frequencies)

  differences <- empirical_cf - theoretical_cf
  statistic <- n_draws(draws) * mean(Mod(differences)^2)

  list(
    statistic = statistic,
    empirical_cf = empirical_cf,
    theoretical_cf = theoretical_cf
  )
}

# Compute the known-parameter covariance matrix of the cosine and sine CF
# features. Product-to-sum identities let us calculate every entry using CF
# values at sums and differences of the original frequencies.
.gof_feature_covariance <- function(frequencies, model, fit) {
  number_frequencies <- nrow(frequencies)
  number_features <- 2 * number_frequencies
  model_cf <- .gof_make_model_cf(model, fit)

  phi <- model_cf(frequencies)
  feature_mean <- c(Re(phi), Im(phi))
  second_moment <- matrix(
    0,
    nrow = number_features,
    ncol = number_features
  )

  for (j in seq_len(number_frequencies)) {
    for (k in seq_len(number_frequencies)) {
      frequency_plus <- frequencies[j, ] + frequencies[k, ]
      frequency_j_minus_k <- frequencies[j, ] - frequencies[k, ]
      frequency_k_minus_j <- -frequency_j_minus_k

      pair_frequencies <- rbind(
        frequency_plus,
        frequency_j_minus_k,
        frequency_k_minus_j
      )
      pair_cf <- model_cf(pair_frequencies)

      phi_plus <- pair_cf[1]
      phi_j_minus_k <- pair_cf[2]
      phi_k_minus_j <- pair_cf[3]

      # cos(A) cos(B) = {cos(A-B) + cos(A+B)} / 2
      second_moment[j, k] <-
        Re(phi_j_minus_k + phi_plus) / 2

      # sin(A) sin(B) = {cos(A-B) - cos(A+B)} / 2
      second_moment[number_frequencies + j,
                    number_frequencies + k] <-
        Re(phi_j_minus_k - phi_plus) / 2

      # cos(A) sin(B) = {sin(A+B) + sin(B-A)} / 2
      second_moment[j, number_frequencies + k] <-
        Im(phi_plus + phi_k_minus_j) / 2
      second_moment[number_frequencies + k, j] <-
        second_moment[j, number_frequencies + k]
    }
  }

  feature_covariance <- second_moment - tcrossprod(feature_mean)

  # Remove tiny asymmetry caused by floating-point calculations.
  (feature_covariance + t(feature_covariance)) / 2
}

#' Characteristic-function goodness-of-fit test for tensor data
#'
#' @param draws A tensor object containing IID tensor draws.
#' @param model Either "normal" or "skewt".
#' @param B Number of parametric bootstrap samples. Use 999 or more for a
#'   reported analysis.
#' @param n_frequencies Number of characteristic-function frequencies.
#' @param frequency_sd Standard deviation used to generate each frequency
#'   coordinate. The default keeps frequency norms near one as p increases.
#' @param max_iter,tol,quiet,restrict Arguments passed to tensor_mle().
#' @param seed Optional random-number seed for reproducibility.
#'
#' @return An htest object containing the statistic and bootstrap p-value.
gof_test <- function(draws,
                     model = c("normal", "skewt"),
                     B = 199L,
                     n_frequencies = 20L,
                     frequency_sd = NULL,
                     max_iter = 1000L,
                     tol = 1e-4,
                     quiet = TRUE,
                     restrict = NULL,
                     seed = NULL) {
  model <- match.arg(model)
  data_name <- deparse(substitute(draws))

  if (!inherits(draws, "tensor")) {
    stop("`draws` must be a tensor object.", call. = FALSE)
  }

  if (n_draws(draws) < 2L) {
    stop("At least two tensor draws are required.", call. = FALSE)
  }

  if (length(B) != 1L || !is.finite(B) || B < 1 || B != floor(B)) {
    stop("`B` must be one positive integer.", call. = FALSE)
  }

  if (length(n_frequencies) != 1L || !is.finite(n_frequencies) ||
      n_frequencies < 1 || n_frequencies != floor(n_frequencies)) {
    stop("`n_frequencies` must be one positive integer.", call. = FALSE)
  }

  B <- as.integer(B)
  n_frequencies <- as.integer(n_frequencies)

  if (!is.null(seed)) {
    set.seed(seed)
  }

  n <- n_draws(draws)
  p <- prod(draw_shape(draws))

  if (is.null(frequency_sd)) {
    frequency_sd <- 1 / sqrt(p)
  }

  # Use one fixed frequency matrix for the observed statistic and every
  # bootstrap statistic. This ensures that they are directly comparable.
  frequencies <- matrix(
    rnorm(n_frequencies * p, sd = frequency_sd),
    nrow = n_frequencies,
    ncol = p
  )

  # Fit the requested null model to the observed draws.
  observed_fit <- tensor_mle(
    draws,
    model = model,
    max_iter = max_iter,
    tol = tol,
    quiet = quiet,
    restrict = restrict
  )

  observed <- .gof_cf_statistic(
    draws = draws,
    model = model,
    fit = observed_fit,
    frequencies = frequencies
  )

  bootstrap_statistics <- numeric(B)

  # Generate samples from the fitted model. Crucially, each bootstrap sample
  # is re-fitted before its statistic is computed.
  for (b in seq_len(B)) {
    if (model == "normal") {
      bootstrap_draws <- rtnorm(
        n = n,
        mu = observed_fit$mu,
        sigmas = observed_fit$sigmas
      )
    } else {
      bootstrap_draws <- rtskewt(
        n = n,
        mu = observed_fit$mu,
        sigmas = observed_fit$sigmas,
        skew = observed_fit$skew,
        nu = observed_fit$nu
      )
    }

    bootstrap_fit <- tensor_mle(
      bootstrap_draws,
      model = model,
      max_iter = max_iter,
      tol = tol,
      quiet = quiet,
      restrict = restrict
    )

    bootstrap_result <- .gof_cf_statistic(
      draws = bootstrap_draws,
      model = model,
      fit = bootstrap_fit,
      frequencies = frequencies
    )

    bootstrap_statistics[b] <- bootstrap_result$statistic
  }

  # The add-one formula prevents a bootstrap p-value of exactly zero.
  p_value <- (1 + sum(bootstrap_statistics >= observed$statistic)) /
    (B + 1)

  method_name <- if (model == "normal") {
    "Bootstrap CF goodness-of-fit test for the tensor-variate normal"
  } else {
    "Bootstrap CF goodness-of-fit test for the tensor-variate skewed t"
  }

  structure(
    list(
      statistic = c(T = observed$statistic),
      parameter = c(B = B),
      p.value = p_value,
      method = method_name,
      data.name = data_name,
      fit = observed_fit,
      frequencies = frequencies,
      empirical_cf = observed$empirical_cf,
      theoretical_cf = observed$theoretical_cf,
      bootstrap_statistics = bootstrap_statistics
    ),
    class = "htest"
  )
}

#' Eigenvalue CF goodness-of-fit test for tensor data
#'
#' This faster version treats the fitted parameters as if they were known. It
#' does not generate tensor bootstrap samples or re-fit the model. Instead, it
#' approximates the null distribution by a weighted sum of chi-squares:
#'
#'     lambda_1 * chi-square_1 + ... + lambda_r * chi-square_1.
#'
#' @param draws A tensor object containing IID tensor draws.
#' @param model Either "normal" or "skewt".
#' @param n_frequencies Number of characteristic-function frequencies.
#' @param n_mixture Number of draws from the weighted chi-squared limit used
#'   to calculate the p-value. These are inexpensive scalar simulations.
#' @param frequency_sd Standard deviation used to generate each frequency
#'   coordinate. The default keeps frequency norms near one as p increases.
#' @param eigen_tolerance Relative cutoff for retaining positive eigenvalues.
#' @param max_iter,tol,quiet,restrict Arguments passed to tensor_mle().
#' @param seed Optional random-number seed for reproducibility.
#'
#' @return An htest object containing the statistic, eigenvalues, and p-value.
gof_test_eigenvalues <- function(draws,
                                 model = c("normal", "skewt"),
                                 n_frequencies = 20L,
                                 n_mixture = 100000L,
                                 frequency_sd = NULL,
                                 eigen_tolerance = 1e-10,
                                 max_iter = 1000L,
                                 tol = 1e-4,
                                 quiet = TRUE,
                                 restrict = NULL,
                                 seed = NULL) {
  model <- match.arg(model)
  data_name <- deparse(substitute(draws))

  if (!inherits(draws, "tensor")) {
    stop("`draws` must be a tensor object.", call. = FALSE)
  }

  if (n_draws(draws) < 2L) {
    stop("At least two tensor draws are required.", call. = FALSE)
  }

  if (length(n_frequencies) != 1L || !is.finite(n_frequencies) ||
      n_frequencies < 1 || n_frequencies != floor(n_frequencies)) {
    stop("`n_frequencies` must be one positive integer.", call. = FALSE)
  }

  if (length(n_mixture) != 1L || !is.finite(n_mixture) ||
      n_mixture < 1 || n_mixture != floor(n_mixture)) {
    stop("`n_mixture` must be one positive integer.", call. = FALSE)
  }

  if (length(eigen_tolerance) != 1L || !is.finite(eigen_tolerance) ||
      eigen_tolerance <= 0) {
    stop("`eigen_tolerance` must be one positive number.", call. = FALSE)
  }

  n_frequencies <- as.integer(n_frequencies)
  n_mixture <- as.integer(n_mixture)

  if (!is.null(seed)) {
    set.seed(seed)
  }

  p <- prod(draw_shape(draws))

  if (is.null(frequency_sd)) {
    frequency_sd <- 1 / sqrt(p)
  }

  frequencies <- matrix(
    rnorm(n_frequencies * p, sd = frequency_sd),
    nrow = n_frequencies,
    ncol = p
  )

  # This is the only model fit required by the eigenvalue method.
  fit <- tensor_mle(
    draws,
    model = model,
    max_iter = max_iter,
    tol = tol,
    quiet = quiet,
    restrict = restrict
  )

  observed <- .gof_cf_statistic(
    draws = draws,
    model = model,
    fit = fit,
    frequencies = frequencies
  )

  feature_covariance <- .gof_feature_covariance(
    frequencies = frequencies,
    model = model,
    fit = fit
  )

  # Because T_n divides the squared CF differences by L, its limiting
  # chi-squared weights are the eigenvalues of V / L.
  eigen_result <- eigen(
    feature_covariance / n_frequencies,
    symmetric = TRUE,
    only.values = TRUE
  )
  all_eigenvalues <- eigen_result$values
  largest_eigenvalue <- max(all_eigenvalues, 0)
  cutoff <- largest_eigenvalue * eigen_tolerance
  eigenvalues <- all_eigenvalues[all_eigenvalues > cutoff]

  if (length(eigenvalues) == 0L) {
    stop("The feature covariance has numerical rank zero.", call. = FALSE)
  }

  # This simulation uses scalar chi-squares only. It does not simulate tensor
  # data and does not call tensor_mle() again.
  mixture_statistics <- numeric(n_mixture)

  for (j in seq_along(eigenvalues)) {
    mixture_statistics <- mixture_statistics +
      eigenvalues[j] * rchisq(n_mixture, df = 1)
  }

  # The add-one formula prevents a simulated p-value of exactly zero.
  p_value <- (1 + sum(mixture_statistics >= observed$statistic)) /
    (n_mixture + 1)

  method_name <- if (model == "normal") {
    paste(
      "Eigenvalue CF goodness-of-fit test for the tensor-variate normal",
      "(fitted parameters treated as known)"
    )
  } else {
    paste(
      "Eigenvalue CF goodness-of-fit test for the tensor-variate skewed t",
      "(fitted parameters treated as known)"
    )
  }

  structure(
    list(
      statistic = c(T = observed$statistic),
      parameter = c(rank = length(eigenvalues)),
      p.value = p_value,
      method = method_name,
      data.name = data_name,
      fit = fit,
      frequencies = frequencies,
      empirical_cf = observed$empirical_cf,
      theoretical_cf = observed$theoretical_cf,
      feature_covariance = feature_covariance,
      eigenvalues = eigenvalues,
      all_eigenvalues = all_eigenvalues,
      n_mixture = n_mixture,
      parameters_treated_as_known = TRUE
    ),
    class = "htest"
  )
}

# Example:
normal_result <- gof_test(
  draws = normal_draws,
  model = "normal",
  B = 199,
  seed = 123
)

normal_result

skewt_draws <- rtskewt(n = 1e3, mu = normal_mu,
                       sigmas = normal_sigmas,
                       skew = matrix(rep(1, 4), nrow = 2, ncol = 2))

# Fast eigenvalue version (no tensor bootstrap or repeated model fits):
fast_normal_result <- gof_test_eigenvalues(
  draws = normal_draws,
  model = "normal",
  seed = 123
)

fast_normal_result
```

# Simple simulation study

This simulation checks three basic behaviors of the eigenvalue test:

1. Normal data should usually not be rejected by the normal model.
2. Skewed (t) data should often be rejected by the normal model.
3. Skewed (t) data should usually not be rejected by the skewed (t) model.

The first and third scenarios estimate the null rejection rate, which should be
near `alpha`. The second scenario estimates power and should have a larger
rejection rate. Because the fitted parameters are currently treated as known,
the null rejection rates are particularly important to inspect.

```{r simple-eigenvalue-simulation}
# Run a small simulation study for the eigenvalue GOF test.
simple_gof_simulation <- function(number_simulations = 25L,
                                  sample_size = 200L,
                                  alpha = 0.05,
                                  n_frequencies = 10L,
                                  n_mixture = 20000L,
                                  seed = 20260721,
                                  show_progress = TRUE) {
  set.seed(seed)

  # Parameters shared by the normal and skewed t distributions.
  mu <- matrix(c(-0.5, 0, 0.5, 1), nrow = 2, ncol = 2)
  sigmas <- list(
    matrix(c(1, 0.25, 0.25, 1.25), nrow = 2, ncol = 2),
    matrix(c(0.8, -0.15, -0.15, 1.1), nrow = 2, ncol = 2)
  )

  # These values create a moderately skewed, heavy-tailed alternative to the
  # tensor normal. Change them to investigate easier or harder alternatives.
  skew <- matrix(c(0.7, -0.35, 0.45, 0.2), nrow = 2, ncol = 2)
  nu <- 6

  scenario_names <- c(
    "normal data: test normal",
    "skewt data: test normal",
    "skewt data: test skewt"
  )

  # There are three test results for each simulation repetition.
  results <- data.frame(
    simulation = rep(seq_len(number_simulations), each = 3L),
    scenario = rep(scenario_names, times = number_simulations),
    p_value = NA_real_,
    reject = NA,
    expected_reject = rep(c(FALSE, TRUE, FALSE),
                          times = number_simulations),
    correct_decision = NA
  )

  row_number <- 1L

  for (simulation in seq_len(number_simulations)) {
    if (show_progress) {
      message(
        "Simulation ", simulation, " of ", number_simulations
      )
    }

    normal_draws <- rtnorm(
      n = sample_size,
      mu = mu,
      sigmas = sigmas
    )

    skewt_draws <- rtskewt(
      n = sample_size,
      mu = mu,
      sigmas = sigmas,
      skew = skew,
      nu = nu
    )

    # Normal model applied to data generated from the normal model.
    normal_null <- tryCatch(
      gof_test_eigenvalues(
        draws = normal_draws,
        model = "normal",
        n_frequencies = n_frequencies,
        n_mixture = n_mixture,
        quiet = TRUE
      ),
      error = function(error) NULL
    )

    if (!is.null(normal_null)) {
      results$p_value[row_number] <- normal_null$p.value
    }
    row_number <- row_number + 1L

    # Normal model applied to skewed t data. Rejection is the desired outcome.
    normal_alternative <- tryCatch(
      gof_test_eigenvalues(
        draws = skewt_draws,
        model = "normal",
        n_frequencies = n_frequencies,
        n_mixture = n_mixture,
        quiet = TRUE
      ),
      error = function(error) NULL
    )

    if (!is.null(normal_alternative)) {
      results$p_value[row_number] <- normal_alternative$p.value
    }
    row_number <- row_number + 1L

    # Skewed t model applied to data generated from the skewed t model.
    skewt_null <- tryCatch(
      gof_test_eigenvalues(
        draws = skewt_draws,
        model = "skewt",
        n_frequencies = n_frequencies,
        n_mixture = n_mixture,
        quiet = TRUE
      ),
      error = function(error) NULL
    )

    if (!is.null(skewt_null)) {
      results$p_value[row_number] <- skewt_null$p.value
    }
    row_number <- row_number + 1L
  }

  results$reject <- results$p_value < alpha
  results$correct_decision <-
    results$reject == results$expected_reject

  # Summarize each scenario with explicit loops to keep the calculation easy
  # to inspect.
  summary <- data.frame(
    scenario = scenario_names,
    successful_fits = integer(length(scenario_names)),
    rejection_rate = numeric(length(scenario_names)),
    correct_decision_rate = numeric(length(scenario_names))
  )

  for (j in seq_along(scenario_names)) {
    rows <- results$scenario == scenario_names[j]
    successful <- rows & !is.na(results$p_value)

    summary$successful_fits[j] <- sum(successful)

    if (sum(successful) > 0L) {
      summary$rejection_rate[j] <- mean(results$reject[successful])
      summary$correct_decision_rate[j] <-
        mean(results$correct_decision[successful])
    } else {
      summary$rejection_rate[j] <- NA_real_
      summary$correct_decision_rate[j] <- NA_real_
    }
  }

  list(
    summary = summary,
    results = results,
    settings = list(
      number_simulations = number_simulations,
      sample_size = sample_size,
      alpha = alpha,
      n_frequencies = n_frequencies,
      n_mixture = n_mixture,
      mu = mu,
      sigmas = sigmas,
      skew = skew,
      nu = nu
    )
  )
}
```

Start with a small run to make sure everything behaves as expected. Increase
`number_simulations` to at least 500 when estimating a (5\%\) rejection rate
for a formal study.

```{r run-simple-eigenvalue-simulation, eval=FALSE}
simulation_result <- simple_gof_simulation(
  number_simulations = 25,
  sample_size = 200,
  seed = 123
)

simulation_result$summary

# Inspect all individual p-values and decisions if needed.
head(simulation_result$results)
```

For the two correctly specified scenarios, do not look only at the
`correct_decision_rate`: a test that never rejects would score highly by that
measure. Also check that the `rejection_rate` is reasonably close to `alpha`
and that the null p-values are spread across ([0,1]). Rejection rates near zero
together with p-values concentrated near one indicate that treating fitted
parameters as known is making the test conservative.


```{r}
univar_norm <- rnorm(1e3, mean = 2, sd = 4)

mean_est <- mean(univar_norm)
sd_est <- sd(univar_norm)

norm_cf <- function(t, mean, sd) {
  exp(1i * mean * t - 1/2 * sd^2 * t^2)
}

empirical_cf <- function(draws, t) {
  z <- complex(length(t))

  for (j in seq_along(t)) {
    z[j] <- mean(exp(1i * draws * t[j]))
  }

  z
}
```


```{r}
library(ggplot2)

t <- seq(-10, 10, length.out = 500)

all_vals <-
  tibble(t,
         z_theory = norm_cf(t, mean_est, sd_est),
         real_theory = Re(z_theory),
         img_theory = Im(z_theory),
         z_emp = empirical_cf(univar_norm, t),
         real_emp = Re(z_emp),
         img_emp = Im(z_emp))  
  
all_vals |>   
  select(t, real_theory, img_theory, real_emp, img_emp) |>
  pivot_longer(cols = -t, names_to = c("component", "source"), 
               names_sep = "_", values_to = "value") |> 
  ggplot() +
  geom_line(aes(x = t, y = value)) +
  facet_grid(source ~ component) +
  theme_minimal()

all_vals
```


```{r}
tau <- 5
#all_vals_diff <- 
  all_vals |>
  mutate(
    weight = exp(-0.5 * (t / tau)^2),

    z_diff = z_emp - z_theory,

    real_diff = real_emp - real_theory,
    img_diff  = img_emp - img_theory,

    weighted_real_diff = weight * real_diff,
    weighted_img_diff  = weight * img_diff,

    abs_diff = Mod(z_diff),
    weighted_sq_error = weight * abs_diff^2
  ) |> 
    
ggplot() +
    geom_histogram(aes(x = weighted_sq_error), 
                   fill = "pink", color = "black", bins = 50) +
    theme_minimal()
```


```{r}
set.seed(123)

B <- 2000
n <- length(univar_norm)

# Observed statistic
mean_est <- mean(univar_norm)
sd_est <- sd(univar_norm)

z_emp <- empirical_cf(univar_norm, t)
z_theory <- norm_cf(t, mean_est, sd_est)

# Use the same weight function as defined above
weight <- exp(-0.5 * t^2)

dt <- diff(t)[1]

T_obs <- n * sum(
  weight * Mod(z_emp - z_theory)^2
) * dt

# Bootstrap null distribution
T_boot <- numeric(B)

for (b in seq_len(B)) {
  x_boot <- rnorm(
    n = n,
    mean = mean_est,
    sd = sd_est
  )

  # Re-estimate parameters within each bootstrap sample
  mean_boot <- mean(x_boot)
  sd_boot <- sd(x_boot)

  z_emp_boot <- empirical_cf(x_boot, t)
  z_theory_boot <- norm_cf(t, mean_boot, sd_boot)

  T_boot[b] <- n * sum(
    weight * Mod(z_emp_boot - z_theory_boot)^2
  ) * dt
}

#bootstrap p-value
mean(T_boot >= T_obs)
```


```{r}
set.seed(123)

R <- 500
alpha <- 0.05
reject <- logical(R)

for (r in seq_len(R)) {
  # Example alternative: standardized t distribution
  x_alt <- as.numeric(scale(rt(n, df = 3)))

  mean_alt <- mean(x_alt)
  sd_alt <- sd(x_alt)

  z_emp_alt <- empirical_cf(x_alt, t)
  z_theory_alt <- norm_cf(t, mean_alt, sd_alt)

  T_alt <- n * sum(
    weight * Mod(z_emp_alt - z_theory_alt)^2
  ) * dt

  # Bootstrap null fitted to the alternative sample
  T_null <- numeric(B)

  for (b in seq_len(B)) {
    x_boot <- rnorm(n, mean_alt, sd_alt)

    z_emp_boot <- empirical_cf(x_boot, t)
    z_theory_boot <- norm_cf(
      t, mean(x_boot), sd(x_boot)
    )

    T_null[b] <- n * sum(
      weight * Mod(z_emp_boot - z_theory_boot)^2
    ) * dt
  }

  p_alt <- mean(T_null >= T_alt)
  reject[r] <- p_alt < alpha
}

mean(reject)
```


# Goal

Let \(X_1,\ldots,X_n\) be independent tensor observations. At a frequency
vector \(t\), the empirical characteristic function (CF) is

\[
\widehat\varphi_n(t)
=
\frac{1}{n}\sum_{i=1}^n \exp\{i t^\top X_i\}.
\]

Its real and imaginary parts are sample averages of \(\cos(t^\top X_i)\)
and \(\sin(t^\top X_i)\). The fitted distribution supplies their theoretical
means and their joint covariance. Consequently, at a fixed set of \(L\)
frequencies, the CF differences can be treated like a vector of correlated
sample-mean differences.

Let \(d\) contain the real and imaginary CF differences at all frequencies,
and let \(V\) be their covariance under the null model. The test statistic is

\[
Q=n d^\top V^+ d,
\]

where \(V^+\) is a numerical pseudoinverse. When the model parameters are
fixed, the asymptotic null distribution is

\[
Q\ \overset{\cdot}{\sim}\ \chi^2_{\operatorname{rank}(V)}.
\]

The code below implements this test for the tensor-variate normal (TVN) and
the tensor-variate skewed \(t\). It avoids a full Kronecker covariance matrix
by whitening each tensor mode separately.

# Shared helpers

```{r cf-helpers}
# Apply mode-specific matrices to one tensor.
.mode_transform <- function(x, matrices) {
  for (k in seq_along(matrices)) {
    x <- n_prod(x, matrices[[k]], k)
  }
  x
}

# Return matrices W_k satisfying W_k Sigma_k W_k' = I.
.whitening_matrices <- function(sigmas) {
  lapply(sigmas, function(S) {
    S <- as.matrix(S)
    S <- (S + t(S)) / 2
    solve(t(chol(S)))
  })
}

# Center and modewise-whiten every tensor, returning an n by p matrix.
.standardize_tensor_sample <- function(draws, mu, sigmas) {
  n <- n_draws(draws)
  dims <- draw_shape(draws)
  p <- prod(dims)
  whitening <- .whitening_matrices(sigmas)
  z <- matrix(NA_real_, nrow = n, ncol = p)

  for (i in seq_len(n)) {
    zi <- pull_draw(draws, i) - mu
    zi <- .mode_transform(zi, whitening)
    z[i, ] <- as.numeric(zi)
  }

  z
}

# The cosine/sine observations whose column means form the empirical CF.
.cf_features <- function(z, frequencies) {
  phase <- z %*% t(frequencies)
  cbind(cos(phase), sin(phase))
}

# Given a CF evaluator, calculate the exact mean vector and covariance matrix
# of [cos(t_1'X),...,cos(t_L'X),sin(t_1'X),...,sin(t_L'X)].
#
# `cf` must accept a matrix with one frequency per row and return one complex
# CF value per row.
.cf_feature_moments <- function(frequencies, cf) {
  L <- nrow(frequencies)
  phi <- cf(frequencies)

  if (length(phi) != L || any(!is.finite(phi))) {
    stop("The CF evaluator returned invalid values.", call. = FALSE)
  }

  feature_mean <- c(Re(phi), Im(phi))
  feature_second <- matrix(0, nrow = 2 * L, ncol = 2 * L)

  for (j in seq_len(L)) {
    for (k in seq_len(L)) {
      t_plus <- frequencies[j, ] + frequencies[k, ]
      t_j_minus_k <- frequencies[j, ] - frequencies[k, ]
      t_k_minus_j <- -t_j_minus_k

      phi_plus <- cf(matrix(t_plus, nrow = 1L))
      phi_j_minus_k <- cf(matrix(t_j_minus_k, nrow = 1L))
      phi_k_minus_j <- cf(matrix(t_k_minus_j, nrow = 1L))

      # Product-to-sum identities:
      # cos(A)cos(B) = {cos(A-B) + cos(A+B)}/2
      # sin(A)sin(B) = {cos(A-B) - cos(A+B)}/2
      # cos(A)sin(B) = {sin(B+A) + sin(B-A)}/2
      feature_second[j, k] <-
        0.5 * Re(phi_j_minus_k + phi_plus)

      feature_second[L + j, L + k] <-
        0.5 * Re(phi_j_minus_k - phi_plus)

      feature_second[j, L + k] <-
        0.5 * Im(phi_plus + phi_k_minus_j)

      feature_second[L + k, j] <- feature_second[j, L + k]
    }
  }

  feature_cov <- feature_second - tcrossprod(feature_mean)
  feature_cov <- (feature_cov + t(feature_cov)) / 2

  list(mean = feature_mean, covariance = feature_cov)
}

# Calculate d' V^+ d using only numerically positive eigenvalues of V.
.quadratic_form_pinv <- function(d, V, relative_tolerance = 1e-8) {
  eig <- eigen(V, symmetric = TRUE)
  cutoff <- max(eig$values, 0) * relative_tolerance
  keep <- eig$values > cutoff

  if (!any(keep)) {
    stop("The CF feature covariance has numerical rank zero.", call. = FALSE)
  }

  projected <- crossprod(eig$vectors[, keep, drop = FALSE], d)
  value <- sum(as.numeric(projected)^2 / eig$values[keep])

  list(value = value, rank = sum(keep), eigenvalues = eig$values)
}
```

# Model characteristic functions

After modewise whitening, the TVN target distribution is \(N_p(0,I_p)\), so

\[
\varphi_{N_p(0,I)}(t)=\exp(-\lVert t\rVert^2/2).
\]

For the package's skewed \(t\), the whitened representation is

\[
Z = W a + \sqrt{W}\,\varepsilon,
\qquad
W\sim\operatorname{InvGamma}(\nu/2,\nu/2),
\qquad
\varepsilon\sim N_p(0,I_p),
\]

where \(a\) is the modewise-whitened skew tensor. Conditional on \(W\),

\[
E\{\exp(i t^\top Z)\mid W\}
=
\exp\left\{W\left(i t^\top a-\frac{1}{2}\lVert t\rVert^2\right)\right\}.
\]

The skewed-\(t\) CF below integrates out \(W\) using a large, reusable set of
inverse-gamma quadrature draws. These are numerical integration values, not
estimates of the unobserved \(W_i\)'s that generated the data.

```{r model-cfs}
.standard_normal_cf <- function(frequencies) {
  values <- exp(-0.5 * rowSums(frequencies^2))
  as.complex(values)
}

.make_standard_skewt_cf <- function(whitened_skew, nu,
                                    n_mixing = 20000L) {
  if (!is.numeric(nu) || length(nu) != 1L || !is.finite(nu) || nu <= 0) {
    stop("`nu` must be one positive finite number.", call. = FALSE)
  }

  w_quadrature <- invgamma::rinvgamma(
    n = n_mixing,
    shape = nu / 2,
    rate = nu / 2
  )

  whitened_skew <- as.numeric(whitened_skew)

  function(frequencies) {
    frequencies <- as.matrix(frequencies)
    q <- rowSums(frequencies^2)
    ta <- drop(frequencies %*% whitened_skew)

    vapply(seq_len(nrow(frequencies)), function(j) {
      if (q[j] == 0 && ta[j] == 0) {
        return(1 + 0i)
      }

      mean(exp(w_quadrature * (1i * ta[j] - 0.5 * q[j])))
    }, FUN.VALUE = complex(1))
  }
}
```

# The test function

The default frequencies have independent \(N(0,1/p)\) entries. Thus their
expected squared norm is approximately one even when the vectorized tensor
dimension \(p\) is large. This avoids the near-zero kernel/CF behavior that
can occur when frequencies are too large in high dimension.

```{r test-function}
tensor_cf_wald_test <- function(draws,
                                model = c("normal", "skewt"),
                                fit = NULL,
                                frequencies = NULL,
                                n_frequencies = 10L,
                                frequency_sd = NULL,
                                n_mixing = 20000L,
                                max_iter = 1000L,
                                tol = 1e-6,
                                quiet = TRUE,
                                restrict = NULL) {
  model <- match.arg(model)

  if (!inherits(draws, "tensor_samples")) {
    stop("`draws` must be a `tensor_samples` object.", call. = FALSE)
  }

  n <- n_draws(draws)
  dims <- draw_shape(draws)
  p <- prod(dims)

  if (n < 2L) {
    stop("At least two tensor observations are required.", call. = FALSE)
  }

  parameters_estimated <- is.null(fit)

  if (parameters_estimated) {
    fit <- tensor_mle(
      draws,
      model = model,
      max_iter = max_iter,
      tol = tol,
      quiet = quiet,
      restrict = restrict
    )
  }

  required <- if (model == "normal") {
    c("mu", "sigmas")
  } else {
    c("mu", "sigmas", "skew", "nu")
  }

  missing_fit <- setdiff(required, names(fit))
  if (length(missing_fit)) {
    stop(
      "`fit` is missing: ", paste(missing_fit, collapse = ", "),
      call. = FALSE
    )
  }

  if (is.null(frequencies)) {
    if (is.null(frequency_sd)) {
      frequency_sd <- 1 / sqrt(p)
    }

    frequencies <- matrix(
      stats::rnorm(n_frequencies * p, sd = frequency_sd),
      nrow = n_frequencies,
      ncol = p
    )
  } else {
    frequencies <- as.matrix(frequencies)

    if (ncol(frequencies) != p) {
      stop(
        "`frequencies` must have ", p, " columns for this tensor shape.",
        call. = FALSE
      )
    }
  }

  z <- .standardize_tensor_sample(draws, fit$mu, fit$sigmas)
  observed_features <- .cf_features(z, frequencies)
  observed_mean <- colMeans(observed_features)

  if (model == "normal") {
    cf <- .standard_normal_cf
  } else {
    whitening <- .whitening_matrices(fit$sigmas)
    whitened_skew <- .mode_transform(fit$skew, whitening)
    cf <- .make_standard_skewt_cf(
      whitened_skew = whitened_skew,
      nu = fit$nu,
      n_mixing = n_mixing
    )
  }

  null_moments <- .cf_feature_moments(frequencies, cf)
  difference <- observed_mean - null_moments$mean
  quadratic <- .quadratic_form_pinv(
    difference,
    null_moments$covariance
  )

  statistic <- n * quadratic$value
  df <- quadratic$rank
  p_value <- stats::pchisq(statistic, df = df, lower.tail = FALSE)

  method <- paste0(
    "Finite-frequency CF-Wald test for the tensor-variate ",
    if (model == "normal") "normal" else "skewed t"
  )

  if (parameters_estimated) {
    method <- paste0(method, " (plug-in parameter estimates)")
  }

  structure(
    list(
      statistic = c(Q = statistic),
      parameter = c(df = df),
      p.value = p_value,
      method = method,
      data.name = deparse(substitute(draws)),
      fit = fit,
      frequencies = frequencies,
      observed_cf = complex(
        real = observed_mean[seq_len(nrow(frequencies))],
        imaginary = observed_mean[nrow(frequencies) + seq_len(nrow(frequencies))]
      ),
      theoretical_cf = complex(
        real = null_moments$mean[seq_len(nrow(frequencies))],
        imaginary = null_moments$mean[
          nrow(frequencies) + seq_len(nrow(frequencies))
        ]
      ),
      difference = difference,
      feature_covariance = null_moments$covariance,
      covariance_eigenvalues = quadratic$eigenvalues,
      parameters_estimated = parameters_estimated
    ),
    class = "htest"
  )
}
```

# TVN example

Here the true parameters are supplied to demonstrate the fixed-parameter
version of the test. Setting `fit = NULL` instead asks the function to estimate
the TVN parameters with `tensor_mle()`.

```{r tvn-example}
set.seed(20260719)

normal_mu <- array(c(-0.5, 0, 0.5, 1), dim = c(2, 2))
normal_sigmas <- list(
  matrix(c(1, 0.3, 0.3, 1.5), 2, 2),
  matrix(c(0.8, -0.2, -0.2, 1.2), 2, 2)
)

normal_draws <- rtnorm(
  n = 1000,
  mu = normal_mu,
  sigmas = normal_sigmas
)

normal_fixed_fit <- list(
  mu = normal_mu,
  sigmas = normal_sigmas
)

set.seed(1)
normal_test <- tensor_cf_wald_test(
  normal_draws,
  model = "normal",
  fit = normal_fixed_fit,
  n_frequencies = 10
)

normal_test
```

For real data, the call is:

```{r tvn-fitted-example, eval=FALSE}
set.seed(1)
normal_test_fitted <- tensor_cf_wald_test(
  observed_tensor_draws,
  model = "normal",
  n_frequencies = 10
)
```

# Tensor skewed-t example

The skewed-\(t\) example uses the same finite-frequency statistic, but its
theoretical CF and covariance are calculated by integrating over the fitted
inverse-gamma mixing distribution.

```{r skewt-example}
set.seed(20260719)

skewt_mu <- array(c(-0.5, 0, 0.5, 1), dim = c(2, 2))
skewt_skew <- array(c(0.35, -0.15, 0.2, 0.1), dim = c(2, 2))
skewt_sigmas <- list(
  matrix(c(1, 0.2, 0.2, 1.3), 2, 2),
  matrix(c(0.9, -0.1, -0.1, 1.1), 2, 2)
)
skewt_nu <- 8

skewt_draws <- rtskewt(
  n = 1000,
  mu = skewt_mu,
  sigmas = skewt_sigmas,
  skew = skewt_skew,
  nu = skewt_nu
)

skewt_fixed_fit <- list(
  mu = skewt_mu,
  sigmas = skewt_sigmas,
  skew = skewt_skew,
  nu = skewt_nu
)

set.seed(2)
skewt_test <- tensor_cf_wald_test(
  skewt_draws,
  model = "skewt",
  fit = skewt_fixed_fit,
  n_frequencies = 10,
  n_mixing = 20000
)

skewt_test
```

For real data, estimate the parameters by omitting `fit`:

```{r skewt-fitted-example, eval=FALSE}
set.seed(2)
skewt_test_fitted <- tensor_cf_wald_test(
  observed_tensor_draws,
  model = "skewt",
  n_frequencies = 10,
  n_mixing = 20000,
  tol = 1e-5
)
```

# Inspecting the CF differences

The result retains the pointwise empirical and theoretical CF values. This is
useful because a single p-value does not show which frequencies disagree.

```{r cf-difference-plot}
cf_difference <- Mod(
  skewt_test$observed_cf - skewt_test$theoretical_cf
)

plot(
  cf_difference,
  type = "h",
  lwd = 2,
  xlab = "Frequency number",
  ylab = "Absolute CF difference",
  main = "Tensor skewed-t CF discrepancies"
)
```

# Interpretation and current limitation

Large values of \(Q\), and therefore small chi-squared p-values, indicate that
the empirical CF differs from the model CF by more than expected from ordinary
sampling variation at the chosen frequencies. Failing to reject does not prove
that the model is correct; it says that this collection of CF features did not
find a detectable discrepancy.

The chi-squared reference distribution directly applies when `fit` contains
parameters fixed independently of `draws`. When `fit = NULL`, the same data are
used both to estimate the model and to test it. Fitting makes the empirical and
theoretical CFs artificially closer. The returned p-value is therefore a
**plug-in asymptotic diagnostic**, not yet a fully calibrated composite-null
test. A final package implementation should add the likelihood-score correction
for parameter estimation and should validate its finite-sample size in a
simulation study. The `parameters_estimated` element of the result records
which case was used.

Using the same fixed frequency matrix is important when comparing models or
repeating a simulation study. For example:

```{r shared-frequencies, eval=FALSE}
set.seed(100)
p <- prod(draw_shape(observed_tensor_draws))
shared_frequencies <- matrix(
  rnorm(10 * p, sd = 1 / sqrt(p)),
  nrow = 10,
  ncol = p
)

normal_result <- tensor_cf_wald_test(
  observed_tensor_draws,
  model = "normal",
  frequencies = shared_frequencies
)

skewt_result <- tensor_cf_wald_test(
  observed_tensor_draws,
  model = "skewt",
  frequencies = shared_frequencies
)
```
