---
title: "normal_mean_variance_mixture"
format: html
editor: visual
execute:
  cache: true
---

```{r, message = FALSE, warning = FALSE, cache = FALSE}
library(tensormodels)
library(tidyverse)
library(latex2exp)
library(patchwork)
library(ggdensity)
devtools::load_all(".", quiet = TRUE)

theme_update(plot.title = element_text(hjust = 0.5))
```

## Poisson Sim for EM

Let $X_{i} \sim \text{Poisson}(\lambda_{i})$ and $Y_{i} \sim \text{Poisson}(\beta \lambda_{i})$ be independent draws. Denote the vector of observations as $\textbf{x}$ and $\textbf{y}$, along with the vector of true $\lambda_{i}$ values as $\mathbf{\lambda}$. 

If we have all of the data, then we can compute the MLEs. 

$$f(\textbf{X}, \textbf{Y}|\beta, \mathbf{\lambda}) = \left[\prod_{i=1}^{n} \frac{e^{-\beta \lambda_{i}} (\beta \lambda_{i})^{y_{i}}}{y_{i}!} \right] \left[\frac{e^{-\lambda_{i}} (\lambda_{i})^{x_{i}}}{x_{i}!}\right]$$

To compute the MLEs, it is easier to take the log likelihood. 

$$\ell(\beta, \mathbf{\lambda}| \textbf{x}, \textbf{y}) = \sum_{i=1}^{n} \bigg[(-\beta \lambda_{i}) + y_{i} \log(\beta \lambda_{i}) - \log(y_{i}!) - \lambda_{i} + x_{i} \log(\lambda_{i}) - \log(x_{i}!)\bigg]\\
 = \sum_{i=1}^{n} \bigg[-\lambda_{i}(\beta + 1) + y_{i} \log(\beta) + (x_{i} + y_{i}) \log(\lambda_{i}) - \log(y_{i}!) - \log(x_{i}!) \bigg]$$
 
Now that we have the log likelihood, we can take the derivative relative to each parameter to obtain the MLEs. 

$$\frac{\partial}{\partial \beta} \ell(\beta, \mathbf{\lambda}|\textbf{x}, \textbf{y}) = \sum_{i=1}^{n} \bigg[-\lambda_{i} + \frac{y_{i}}{\beta} \bigg] = 0 \\ \sum_{i=1}^{n} \bigg[-\beta \lambda_{i}\bigg] = \sum_{i=1}^{n} \bigg[-y_{i} \bigg] \\
\sum_{i=1}^{n} \bigg[\beta \bigg] = \sum_{i=1}^{n} \bigg[\frac{y_{i}}{\lambda_{i}} \bigg] \\ 
\frac{\partial}{\partial \mathbf{\lambda}} \ell(\beta, \mathbf{\lambda}|\textbf{x}, \textbf{y}) =\sum_{i=1}^{n} \bigg[-(\beta + 1) + \frac{(x_{i} + y_{i})}{\lambda_{i}} \bigg] = 0 \\
\sum_{i=1}^{n} \bigg[-\lambda_{i} (\beta + 1) + (x_{i} + y_{i})\bigg] = 0 \\
\sum_{i=1}^{n} \bigg[-\lambda_{i} (\beta + 1)\bigg] = -\sum_{i=1}^{n} \bigg[x_{i} + y_{i}\bigg]
$$

$$\hat{\beta} = \frac{\sum_{i=1}^{n} y_{i}}{\sum_{i=1}^{n} x_{i}}$$ 

and $$\hat{\lambda}_{i} = \frac{x_{i} + y_{i}}{\hat{\beta} + 1}.$$

```{r}
n <- 20

lambdas <- 30 * runif(n = 20)
beta <- 0.2

x <- rpois(n = n, lambda = lambdas)
y <- rpois(n = n, lambda = beta * lambdas)

beta_hat <- sum(y)/sum(x)

lambdas_hat <- (x + y)/(beta_hat + 1)

sqrt(mean((lambdas_hat - lambdas)^2))
```


Now, suppose we are missing $x_{1}$ and still want to estimate $\lambda_{1}$. The incomplete likelihood is now 

$$L(\beta, \lambda_{1}, \lambda_{2}, \dots, \lambda_{n}|y_{1}, (x_{2}, y_{2}), \dots, (x_{n}, y_{n}) = \\ \left[\prod_{i=1}^{n} \frac{e^{-\beta \lambda_{i}} (\beta \lambda_{i}^{y_{i}})}{y_{i}!} \right] \left[\prod_{i=2}^{n} \frac{e^{-\lambda_{i}} (\lambda_{i}^{x_{i}})}{_{i}!}\right]$$

```{r}
x_missing <- c(NA, x[2:n])
```

### Optim

```{r}
fit_beta_lambda <- function(x, y) {
  n <- length(x)
  
  neg_loglik <- function(par) {
    beta <- exp(par[1])
    lambda <- exp(par[-1])
    
    ll_y <- sum(dpois(y, lambda = beta * lambda, log = TRUE))
    
    ll_x <- sum(dpois(x[2:n], lambda = lambda[2:n], log = TRUE))
    
    -(ll_y + ll_x)
  }
  
  start_beta <- 1
  start_lambda <- pmax(y/x, 0.1)
  start_lambda[is.na(start_lambda)] <- 1
  start_lambda[is.infinite(start_lambda)] <- 1
  
  fit <- optim(par = log(c(start_beta, start_lambda)),
               fn = neg_loglik, method = "BFGS", hessian = TRUE)
  
  beta_hat <- exp(fit$par[1])
  lambda_hat <- exp(fit$par[-1])
  
  list(beta = beta_hat, lambda = lambda_hat)
}

fit_beta_lambda(x_missing, y)
```

### EM

```{r}
fit_beta_lambda_em <- function(x, y, max_iter = 1000, tol = 1e-8) {
  n <- length(x)
  
  beta <- 1
  lambda <- pmax(y/x, 0.1)
  lambda[is.na(lambda)] <- 1
  lambda[is.infinite(lambda)] <- 1

  loglik <- function(beta, lambda) { # incomplete loglik
    sum(dpois(y, lambda = beta * lambda, log = TRUE)) +
      sum(dpois(x[2:n], lambda = lambda[2:n], log = TRUE))
  }

  loglik_vals <- numeric(max_iter)

  for (iter in seq_len(max_iter)) {
    beta_old <- beta
    lambda_old <- lambda

    # E-step
    x1_hat <- lambda[1] # sufficient stat is lambda1

    # M-step
    beta <- sum(y) / (x1_hat + sum(x[2:n]))

    lambda[1] <- (x1_hat + y[1]) / (1 + beta)
    lambda[2:n] <- (x[2:n] + y[2:n]) / (1 + beta)

    loglik_vals[iter] <- loglik(beta, lambda)

    diff <- max(abs(c(beta, lambda) - c(beta_old, lambda_old)))

    if (diff < tol) { #s top when converge
      loglik_vals <- loglik_vals[seq_len(iter)]
      break
    }
  }

  list(beta = beta, lambda = lambda)
}

fit_beta_lambda_em(x_missing, y)
```


## Plots of distributions

### dtnorm

```{r}
mu <- matrix(c(0, 0), nrow = 2, ncol = 1)

sigmas <- list(matrix(c(1, 0.5,
                        0.5, 1), nrow = 2))

eval_points <- expand_grid(x11 = seq(-8, 8, length.out = 150),
                           x21 = seq(-8, 8, length.out = 150)) 

eval_mat <- as.matrix(eval_points)

eval_dtnorm <- .make_dtnorm_evaluator(mu, sigmas, log = TRUE)

dtnorm_density <- vector("numeric", nrow(eval_points))

for(i in 1:nrow(eval_points)) {
  dtnorm_density[[i]] <- eval_dtnorm(matrix(eval_mat[i,], nrow = 2))
}

norm_density <-
  eval_points |> 
  mutate(log_density = dtnorm_density) 

ggplot(norm_density, aes(x = x11, y = x21)) +
  geom_raster(aes(fill = log_density)) +
  geom_contour(aes(z = log_density), color = "black", linewidth = 0.4) +
  scale_fill_viridis_c(option = "D") +
  coord_fixed() +
  labs(x = "X1", y = "X2", fill = "Log Density", title = "Normal") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

### Normal Mean-Variance Mixture with $U \sim Unif(0, 1)$

```{r}
mean_mix <- function(n, mu = 0, sigmas = list(matrix(1)), skew = 1, nu = 4) {
  dims <- dim(mu)
  
  if(is.null(dims)) dims <- length(mu)

  # draw tensor variate normals
  tensor_norms <- rtnorm(n, mu = array(0, dim = dims), sigmas)

  # generate unif
  unif_draws <- runif(n = n)

  mix_unif_draws <- vector("list", n)

  for(i in 1:n) { # scale normals and skew by unif
    mix_unif_draws[[i]] <- mu + tensor_norms[[i]] * sqrt(unif_draws[i]) +
                           skew * unif_draws[i]
  }

  mix_unif_draws
}
```


### dunifmix no skew vs skew

```{r}
s1 <- matrix(c(1, 0.25, 0.25, 0.5), nrow = 2)

n <- 1e5

mix_unif_noskew <- mean_mix(n = n, mu = c(1, 2), 
                            sigmas = list(s1), skew = c(0, 0))

pull_draw <- function(i, data) {
  curr_draw <- data[[i]]
  tibble(val = c(curr_draw[1], curr_draw[2]), 
         pos = c("x1", "x2"), draw = c(i, i))
}

noskew_draws_pos <- map_dfr(.x = 1:n, .f = pull_draw, 
                            data = mix_unif_noskew)

mix_unif_skew <- mean_mix(n = n, mu = c(1, 2), 
                          sigmas = list(s1), skew = c(2, 1))

skew_draws_pos <- map_dfr(.x = 1:n, .f = pull_draw, 
                          data = mix_unif_skew)

mean_points <-
  tibble(setting = c("No Skew", "Skew = c(2, 1)"),
         x1 = c(1, 2), x2 = c(2, 2.5),
         point = c("E[X] no skew", "E[X] skew"))

bind_rows(noskew_draws_pos |> mutate(setting = "No Skew"),
          skew_draws_pos |> mutate(setting = "Skew = c(2, 1)")) |>
  pivot_wider(names_from = pos, values_from = val) |> 
  ggplot(aes(x = x1, y = x2)) + 
  geom_density_2d_filled(breaks = seq(0.1, 1, by = 0.1),
                          color = "black", linewidth = 0.2) +
  geom_point(data = mean_points,
             aes(x = x1, y = x2, color = point), size = 1.2) +
  facet_wrap(~setting) +
  scale_fill_viridis_d(option = "D") +
  coord_fixed() +
  labs(x = TeX("$\\X_{1}$"),
       y = TeX("$\\X_{2}$"),
       fill = "Density",
       color = "Mean",
       title = TeX("W $\\sim$ Unif(0, 1)")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```


### dtskewt no skew vs skew

```{r}
# 
# ggplot() +
#   geom_hdr_fun(fun = noskew_fun, xlim = c(-3, 5), ylim = c(-3, 6),
#                probs = c(0.50, 0.75, 0.95, 0.99),
#                aes(fill = factor(after_stat(probs))),
#                color = "black", linewidth = 0.4, alpha = 0.6) +
#   scale_fill_viridis_d(name = "HDR probability", direction = -1) +
#   coord_fixed() +
#   labs(x = expression(X[1]), y = expression(X[2])) +
#   theme_minimal() +
# ggplot() +
#   geom_hdr_fun(fun = skew_fun, xlim = c(-3, 5), ylim = c(-3, 6),
#                probs = c(0.50, 0.75, 0.95, 0.99),
#                aes(fill = factor(after_stat(probs))),
#                color = "black", linewidth = 0.4, alpha = 0.6) +
#   scale_fill_viridis_d(name = "HDR probability", direction = -1) +
#   coord_fixed() +
#   labs(x = expression(X[1]), y = expression(X[2])) +
#   theme_minimal()

eval_points <- 
  expand_grid(x11 = seq(-6, 8, length.out = 300),
              x21 = seq(-6, 8, length.out = 300))

multivar_noskewt <- .make_dtskewt_evaluator(mu = c(1, 2), skew = c(1e-6, 1e-6),
                                          sigmas = list(s1), nu = 20, log = TRUE)


multivar_skewt <- .make_dtskewt_evaluator(mu = c(1, 2), skew = c(2, 1),
                                          sigmas = list(s1), nu = 20, log = TRUE)

multivar_dtnoskewt <- vector("numeric", nrow(eval_points))
multivar_dtskewt <- vector("numeric", nrow(eval_points))

for(i in 1:nrow(eval_points)) {
  multivar_dtnoskewt[[i]] <- multivar_noskewt(unlist(eval_points[i,]))
  multivar_dtskewt[[i]] <- multivar_skewt(unlist(eval_points[i,]))
}

tibble(eval_points, multivar_dtnoskewt, multivar_dtskewt) |> 
  pivot_longer(cols = c(multivar_dtnoskewt, multivar_dtskewt)) |> 
  mutate(name = ifelse(name == "multivar_dtnoskewt", 
                       "No skew", "Skew = c(2, 1)")) |> 
  ggplot(aes(x = x11, y = x21)) +
  geom_raster(aes(fill = value)) +
  geom_contour(aes(z = value), color = "black", linewidth = 0.4) +
  facet_grid(~name) +
  scale_fill_viridis_c(option = "D") +
  coord_fixed() +
  labs(x = TeX("$\\X_{1}$"),
       y = TeX("$\\X_{2}$"),
       fill = "Log Density",
       title = TeX("W $\\sim$ Inv-Gamma(10, 10)")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```


### dtskewt hdr no skew vs skew

```{r}
noskew_dtskewt <- .make_dtskewt_evaluator(mu = c(1, 2), skew = c(1e-6, 1e-6), sigmas = list(s1), nu = 20)

noskew_fun <- function(x, y) {
  x_list <- Map(c, x, y)
  noskew_dtskewt(x_list)
}

skew_dtskewt <- .make_dtskewt_evaluator(mu = c(1, 2), skew = c(2, 1), sigmas = list(s1), nu = 20)

skew_fun <- function(x, y) {
  x_list <- Map(c, x, y)
  skew_dtskewt(x_list)
}

ggplot() + 
  geom_hdr_fun(fun = noskew_fun, xlim = c(-4, 6), ylim = c(-2, 6)) +
  labs(x = TeX("$\\X_{1}$"),
       y = TeX("$\\X_{2}$"),
       subtitle = "No skew") + 
  coord_fixed(xlim = c(-4, 8), ylim = c(-2, 6)) +
  theme_minimal() +
ggplot() + 
  geom_hdr_fun(fun = skew_fun, xlim = c(-4, 6), ylim = c(-2, 6)) +
  labs(x = TeX("$\\X_{1}$"),
       y = TeX("$\\X_{2}$"),
       subtitle = "Skew = c(2, 1)") +
  coord_fixed(xlim = c(-4, 8), ylim = c(-2, 6)) +
  theme_minimal()
```


### dtskewt different nu

```{r}
nu_vals <- seq(4, 26, by = 2) 

mean_diff_nus <-
  tibble(nu = nu_vals,
         x1 = 1 + (nu)/(nu-2) * 2,
         x2 = 2 + (nu)/(nu-2) * 1,
         point = as.character(nu))


eval_points <- 
  expand_grid(x11 = seq(-4, 10, length.out = 200),
              x21 = seq(-4, 10, length.out = 200))

x_list <- vector(mode = "list", length = 200*200)

for(i in 1:(200*200)) {
  x_list[[i]] <- unlist(eval_points[i,])
}

skewt_density <-
  tibble(nu = nu_vals) |> 
  rowwise() |> 
  mutate(eval_dtskewt = 
         list(.make_dtskewt_evaluator(mu = c(1, 2), sigmas = list(s1),
                                      skew = c(2, 1), nu = nu, log = TRUE))) |> 
  mutate(density_grid = list(mutate(eval_points, nu = nu, 
                                    log_density = eval_dtskewt(x_list)))) |>
  ungroup() |> 
  select(density_grid) |> 
  unnest(density_grid) |> 
  mutate(nu = factor(nu, levels = nu_vals))

ggplot(skewt_density, aes(x = x11, y = x21)) +
  geom_raster(aes(fill = log_density), interpolate = TRUE) +
  geom_contour(aes(z = log_density), color = "black", linewidth = 0.25) +
  geom_point(data = mean_diff_nus,
             aes(x = x1, y = x2), size = 1.2, color = "blue") +
  facet_wrap(~nu, ncol = 4) +
  scale_fill_viridis_c(option = "D") +
  coord_fixed() +
  labs(x = TeX("$\\X_{1}$"),
       y = TeX("$\\X_{2}$"),
       fill = "Log density",
       title = "Skew-t density across degrees of freedom",
       subtitle = TeX("$\\mu = c(1, 2), A = c(2, 1)$")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```


### dtskewt matrix variate

```{r}
mu <- matrix(c(1, 2,
               3, 4), nrow = 2)

skew <- matrix(c(2, 1,
                 1, 0.5), nrow = 2)

s1 <- matrix(c(1, 0.5, 0.5, 2), nrow = 2)
s2 <- matrix(c(2, 1, 1, 2), nrow = 2)
  
eval_points <- 
  expand_grid(x11 = seq(-2, 6, length.out = 40),
              x21 = seq(-2, 6, length.out = 40),
              x12 = 0:10, x22 = 0:10) 

eval_matrix <- as.matrix(eval_points)

dtskewt_density <- vector("numeric", nrow(eval_points))

eval_dtskewt <- .make_dtskewt_evaluator(mu, skew, sigmas = list(s1, s2), 
                                        nu = 20, log = TRUE)

for(i in 1:nrow(eval_points)) {
  dtskewt_density[[i]] <- eval_dtskewt(matrix(eval_matrix[i,], nrow = 2))
}

mat_skewt_density <-
  eval_points |> 
  mutate(log_density = dtskewt_density) 

mat_skewt_density |> 
  mutate(x12_lab = factor(round(x12, 2), 
                          levels = sort(unique(round(x12, 2)))),
         x22_lab = factor(round(x22, 2), 
                          levels = sort(unique(round(x22, 2))))) |> 
  ggplot(aes(x = x11, y = x21)) +
  geom_raster(aes(fill = log_density)) +
  geom_contour(aes(z = log_density), color = "black") +
  facet_grid(rows = vars(`X[2,2]` = x22_lab),
             cols = vars(`X[1,2]` = x12_lab), 
             labeller = label_both) +
  scale_fill_viridis_c(option = "H") +
  coord_fixed() +
  labs(x = "X[1,1]", y = "X[2,1]",
      fill = "Log Density",
      title = "Skewed T") +
  theme_minimal(base_size = 9) +
  theme(plot.title = element_text(hjust = 0.5))
```


```{r}
plot_density(x12 = seq(-4, 6, by = 1),
             x22 = seq(-4, 6, by = 1),
             x11 = seq(-6, 6, length.out = 50),
             x21 = seq(-6, 6, length.out = 50),
             mu = mu, sigmas = list(s1, s2),
             model = "skewt", skew = skew, nu = 20)
```


### dtnorm matrix variate

```{r}
plot_density(x12 = seq(0, 10, by = 1),
             x22 = seq(0, 10, by = 1),
             x11 = seq(-2, 6, length.out = 50),
             x21 = seq(-2, 6, length.out = 50),
             mu = mu, sigmas = list(s1, s2),
             model = "normal")
```


```{r}
dtnorm_density <- vector("numeric", nrow(eval_points))

eval_matnorm <- .make_dtnorm_evaluator(mu, sigmas = list(s1, s2), log = TRUE)

for(i in 1:nrow(eval_points)) {
  dtnorm_density[[i]] <- eval_matnorm(matrix(eval_matrix[i,], nrow = 2))
}

mat_normal_density <-
  eval_points |> 
  mutate(log_density = dtnorm_density) 

mat_normal_density |> 
  mutate(x12_lab = factor(round(x12, 2), 
                          levels = sort(unique(round(x12, 2)))),
         x22_lab = factor(round(x22, 2), 
                          levels = sort(unique(round(x22, 2))))) |> 
  ggplot(aes(x = x11, y = x21)) +
  geom_raster(aes(fill = log_density)) +
  geom_contour(aes(z = log_density), color = "black") +
  facet_grid(rows = vars(`X[2,2]` = x22_lab),
             cols = vars(`X[1,2]` = x12_lab), 
             labeller = label_both) +
  scale_fill_viridis_c(option = "H") +
  coord_fixed() +
  labs(x = "X[1,1]", y = "X[2,1]",
      fill = "log density",
      title = "Normal") +
  theme_minimal(base_size = 9)
```


### dtskewt multivar for different skew

```{r}
mu <- matrix(c(0.5, 0.25, 0.25, 0.5), nrow = 2)

skew_settings <- tibble(
  skew_name = c("No skew", "Mild skew", "Strong skew"),
  skew = list(matrix(c(1e-6, 1e-6, 1e-6, 1e-6), nrow = 2),
              matrix(c(1, 0.5, 0.5, 0.25), nrow = 2),
              matrix(c(2, 1, 1, 0.5), nrow = 2)))

eval_points_2d <- expand_grid(x11 = seq(-10, 10, length.out = 200),
                              x21 = seq(-10, 10, length.out = 200))

density_compare <-
  skew_settings |>
  mutate(skew_name = factor(skew_name, levels = c("No skew", "Mild skew", "Strong skew")),
         dtskewt_eval = lapply(skew, function(curr_skew) {
           .make_dtskewt_evaluator(mu = mu, skew = curr_skew, sigmas = list(s1, s2),
                                   nu = 20, log = TRUE)
           })) |>
  crossing(eval_points_2d) |>
  group_by(skew_name) |>
  mutate(log_density = {
      f <- dtskewt_eval[[1]]
      x_list <- lapply(seq_len(n()), function(i) {
        matrix(c(x11[i], x21[i],
                 mu[1, 2], mu[2, 2]), nrow = 2)
      })

      f(x_list)
    }) |>
  ungroup() |>
  select(-dtskewt_eval)

ggplot(density_compare, aes(x = x11, y = x21)) +
  geom_raster(aes(fill = log_density)) +
  geom_contour(aes(z = log_density), color = "black") +
  facet_wrap(~skew_name, nrow = 1) +
  scale_fill_viridis_c(option = "H") +
  coord_fixed() +
  labs(
    x = "X[1,1]",
    y = "X[2,1]",
    fill = "log density",
    title = "Effect of skew on a matrix-variate skew-t density",
    subtitle = "X[1,2] and X[2,2] fixed at their mean values"
  ) +
  theme_minimal()
```
