---
output: github_document
bibliography: Bibliography.bib
csl: nature.csl
header-includes: 
  - |
    \AtBeginDocument{
      \renewcommand{\r}[1]{\textcolor{red}{#1}}
      \renewcommand{\b}[1]{\textcolor{blue}{#1}}
    }
  - \usepackage{xcolor}
  - \usepackage{algorithm}
  - \usepackage{algpseudocode}
  - \usepackage{mathtools}
  - \usepackage{float}
  - \newcommand{\coleq}{\mathrel{\mathop:}=}
  - \newcommand{\ve}[1]{\textbf{#1}}
  - \newcommand{\field}[1]{\mathbb{#1}}
  - \newcommand{\defn}[1]{\emph{#1}}
  - \newcommand{\oneVect}[1]{#1}
  - \newcommand{\mat}[1]{\textbf{#1}}
  - \newcommand{\oneMat}[1]{#1}
  - \newcommand{\tensor}[1]{\mathcal{#1}}
  - \newcommand{\oneTensor}[1]{#1} 
  - \newcommand{\vs}[1]{\mathcal{#1}}
  #- \newcommand{\mn}[1]{{#1} \choose {#2}}
  #- \newcommand{\mn}[2]{\mathrel{\underset{\scriptstyle {#1}\,{#2}}{\times}}}
  #- \newcommand{\mn}[2]{\underset{\scriptsize\color{red}{#1}\,\color{blue}{#2}}{\times}}
  #- \newcommand{\mn}[2]{\mathbb{#1}}
  #- \newcommand{\mn}[2]{\sideset{_{\color{red}{#1}}}{_{\color{blue}{#2}}}{\boldsymbol{\times}}}
  - \newcommand{\mn}[2]{\sideset{_{#1}}{_{#2}}{\mathop{\boldsymbol{\times}}}}
  - \newtheorem{proposition}{Proposition}[section]
  - \newtheorem{proof}{Proof}[section]
---




```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
set.seed(1)
```

# **tensormodels**

Tensor operations, decompositions, and distributions in R. 

```{r}
#| message: false
library("tensormodels")
library("tidyverse")
library("tictoc")
old_digits <- getOption("digits")
options(digits = 3)
```







# Operations



## n-mode prod

The $n$-mode matrix product of  a tensor $\tensor{A} \in \field{k}^{I_1 \times I_2 \times ... \times I_n}$ with a matrix $\mat{U} \in \field{k}^{J \times I_n}$  
is
$$(\tensor{A} \times_n \mat{U}) \in \field{k}^{I_1 \times ... \times I_{n-1} \times J \times I_{n+1} \times ... \times I_N}$$ with entries 
$$(\tensor{A} \times_n \mat{U})_{i_1, ..., i_{n-1}, j, i_{n+1}, ... I_N} = \sum_{i_n = 1}^{I_n} a_{i_1, i_2, \dots, i_N} u_{j, i_n}.$$ 
The tensor and matrix share one mode in common, denoted here as $I_n$. This operation is also called the tensor times matrix product.

The n-mode product between a tensor and a matrix can be computed with `n_mode_prod()`{.R}.

```{r}
a <- array(1:3, dim = c(3, 1, 1))
x <- matrix(4:9, nrow = 2, ncol = 3)

n_prod(a, x, 1)
```





## nm-mode prod

The nm-mode product between a tensor and a tensor can be computed with
`nm_prod()`{.R}.

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)
x <- matrix(c(5, 6), nrow = 2)

nm_prod(A, x, 1, 1)
```




## [Tensor product](https://en.wikipedia.org/wiki/Tensor_product)

The tensor product between a tensor and a tensor can be computed with
`tensor_prod()`{.R}.

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)
x <- matrix(c(5, 6), nrow = 2)

tensor_prod(A, x)
```






# Simulating tensor variate normal draws

The density function of the multilinear normal distribution is @tensornormprop
$$f(x) = (2\pi)^{-p*/2} \left(\prod_{i=1}^{k} |\Sigma_i|^{-p*/(2\pi)}\right) \exp\left\{-\frac{1}{2} (x-\mu)' \Sigma_{1:k}^{-1} (x-\mu\right\}$$
where $\Sigma$ is positive definite, $x \in \mathbb{R}^p$, $\mu \in \mathbb{R}^p$ and $\Sigma_{1:k} \in \mathbb{R}^p.$

The function `rtnorm()`{.R} simulates random draws from the tensor variate normal with a specified mean array mu and a list of covariance matrices called list_sigmas.

Since the univariate normal and matrix variate normal are simpler cases of the
tensor variate normal, this function can simulate from them as well.  Here, we
simulate random draws from a univariate normal with mean $-2$ and variance $4$.


```{r}
univar_draws <- rtnorm(n = 1000, mu = -2, sigmas = 4)

mean(simplify2array(univar_draws))

var(simplify2array(univar_draws))
```


We can also simulate random draws from a multivariate normal distribution by specifying a mean vector and the covariance matrix. 

```{r}
S1 <- crossprod(matrix(data = c(1, 0.5, 0.5, 1), nrow = 2))

(multivar_draws <- rtnorm(n = 5, mu = c(2, 3), sigmas = list(S1)))
```


And now we simulate from the matrix variate normal of size $2 \times 3$. By
default, the covariance matrices will be the identity.

```{r}
matrix_draws <- rtnorm(n = 1e3, mu = matrix(1:6, nrow = 2, ncol = 3))

matrix_draws[[1]]
```


Below is a simulation from a tensor variate normal of size $3 \times 4 \times
2$.The covariance structure is specified with a list of matrices. 

```{r}
mu_true <- array(1:24, dim = c(3, 4, 2))

S1 <- crossprod(matrix(rnorm(9), nrow = 3))
S2 <- crossprod(matrix(rnorm(16), nrow = 4))
S3 <- crossprod(matrix(rnorm(4), nrow = 2))

tvn_draws <- rtnorm(n = 1e3, mu = mu_true,
                    sigmas = list(S1, S2, S3))

tvn_draws[[1]]
```



# Estimation

## MLE estimation for the tensor variate normal 

The package supports MLE estimation. Given an array of draws, it will return a list containing the MLE for the mean and covariance matrices. 

```{r}
(univarnorm_est <- tensor_mle(draws = univar_draws, model = "normal"))
```


Here, the mean matrix from the draws above has values 1 through 6. The
covariance matrices are the identity matrices by default.

```{r}
matrix_est <- tensor_mle(matrix_draws, model = "normal")

matrix_est$mu
matrix_est$sigmas |> lapply(round, 3)
```


This function works for tensor-valued data. 

```{r}
tensor_est <- tensor_mle(draws = tvn_draws, model = "normal")

frob_norm_diff(tensor_est$mu, mu_true)

true_sigmas <- list(S1, S2, S3)

for(i in 1:3) {
  true_scaled <- true_sigmas[[i]] / sum(diag(true_sigmas[[i]]))
  est_scaled <- tensor_est$sigmas[[i]] / sum(diag(tensor_est$sigmas[[i]]))
  
  print(frob_norm(true_scaled - est_scaled) / frob_norm(true_scaled))
}
```






# Other models

## Tensor variate normal inverse Gaussian 

The density function of the normal inverse Gaussian distribution is

$$f_{\text{TVNIG}}(\tensor{X}|\mat{V}) =
\frac{2 \exp\left\{\text{vec}(\tensor{X} - \tensor{M})^{T} \bigotimes_{d=1}^{D} \Sigma_{d}^{-1} \text{vec}(\tensor{A} + \kappa) \right\}}{(2\pi)^{\frac{n^{*}}{2}} \prod_{d=1}^{D} |\Sigma_{d}|^{\frac{n^{*}}{2n_{d}}}} \left(\frac{\delta(\tensor{X}; \tensor{M}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + 1}{\rho(\tensor{A}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + \kappa^2}\right)^{-\frac{1 + n^{*}}{4}}$$ 

$$\quad K_{- \frac{1 + n^{*}}{2}} \left(\sqrt{[\rho(\tensor{A}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}}) + \kappa^{2}] \left[\delta(\tensor{X}; \tensor{M}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + 1\right]\right)$$
where $\Sigma$ is positive definite, $x \in \mathbb{R}^p$, $\mu \in \mathbb{R}^p$ and $\Sigma_{1:k} \in \mathbb{R}^p.$

The function `rtinvgauss()`{.R} simulates random draws from the tensor variate normal inverse Gaussian with a specified mean array mu, a list of covariance matrices, a skew array, and $\kappa$ which describes the shape of the inverse Gaussian distribution

```{r}
mu_true <- array(1:12, dim = c(3, 4))
skew_true <- array(seq(0, 4, length.out = 12), dim = c(3, 4))
kappa_true <- 2

invgauss_draws <- rtinvgauss(n = 1e3, mu = mu_true, skew = skew_true, 
                             sigmas = list(S1, S2), kappa = kappa_true)
```


```{r}
mu_true <- array(1:24, dim = c(3, 4, 2))
skew_true <- array(seq(0, 4, length.out = 24), dim = c(3, 4, 2))
kappa_true <- 2

invgauss_draws <- rtinvgauss(n = 1e3, mu = mu_true, skew = skew_true, 
                             sigmas = list(S1, S2), kappa = kappa_true)
```


```{r}
invgauss_est <- tensor_mle(invgauss_draws, model = "invgauss", 
                           quiet = FALSE, tol = 1e-3)
```


For the inverse gaussian distribution, the true mean $E[X] = \tensor{M} + \tensor{A}/\kappa$. We can compare the model's estimation for the mean with the true values. 

```{r}
frob_norm_diff(with(invgauss_est, mu + skew/kappa),
               mu_true + skew_true/kappa_true)
```


```{r}
frob_norm_diff(invgauss_est$mu, mu_true)
frob_norm_diff(invgauss_est$skew, skew_true)
invgauss_est$kappa

for(i in 1:2) {
  true_scaled <- true_sigmas[[i]] / sum(diag(true_sigmas[[i]]))
  est_scaled <- invgauss_est$sigmas[[i]] / sum(diag(invgauss_est$sigmas[[i]]))
  
  print(frob_norm(true_scaled - est_scaled) / frob_norm(true_scaled))
}
```




## Tensor variate generalized hyperbolic

The density function of the generalized hyperbolic distribution is
$$f(x) = (2\pi)^{-p*/2} \left(\prod_{i=1}^{k} |\Sigma_i|^{-p*/(2\pi)}\right) \exp\left\{-\frac{1}{2} (x-\mu)' \Sigma_{1:k}^{-1} (x-\mu\right\}$$
where $\Sigma$ is positive definite, $x \in \mathbb{R}^p$, $\mu \in \mathbb{R}^p$ and $\Sigma_{1:k} \in \mathbb{R}^p.$

The function `rtgenhyper()`{.R} simulates random draws from the tensor variate generalized hyperbolic with a specified mean array mu and a list of covariance matrices called list_sigmas.

Since the univariate normal and matrix variate normal are simpler cases of the
tensor variate normal, this function can simulate from them as well.  Here, we
simulate random draws from a univariate normal with mean $-2$ and variance $4$.

```{r}
lambda_true <- 2
omega_true <- 2

genhyper_draws <- rtgenhyper(n = 1e3, mu = mu_true, skew = skew_true,
                             sigmas = list(S1, S2, S3))
```



For the generalized hyperbolic distribution, the true mean $E[X] = \tensor{M} + \frac{K_{\lambda + 1}(\omega)}{K_{\lambda}(\omega)} \tensor{A}$. We can compare the model's estimation for the mean with the true values. 

```{r}
genhyper_est <- tensor_mle(genhyper_draws, model = "genhyper", quiet = FALSE)

frob_norm_diff(
  with(genhyper_est, mu + besselK(x = omega, nu = lambda + 1)/
                          besselK(x = omega, nu = lambda) * skew),
       mu_true + besselK(x = omega_true, nu = lambda_true + 1)/
                 besselK(x = omega_true, nu = lambda_true) * skew_true)

frob_norm_diff(genhyper_est$mu, mu_true)
frob_norm_diff(genhyper_est$skew, skew_true)

for(i in 1:3) {
  true_scaled <- true_sigmas[[i]] / sum(diag(true_sigmas[[i]]))
  est_scaled <- genhyper_est$sigmas[[i]] / sum(diag(genhyper_est$sigmas[[i]]))
  
  print(frob_norm(true_scaled - est_scaled) / frob_norm(true_scaled))
}

genhyper_est$lambda
genhyper_est$omega
```


## Tensor variate variance gamma
```{r}
vargamma_draws <- rtvargamma(n = 1e3, mu = mu_true, skew = skew_true,
                             sigmas = list(S1, S2, S3))
```



```{r}
vargamma_est <- tensor_mle(vargamma_draws, model = "vargamma", quiet = FALSE)

frob_norm_diff(with(vargamma_est, mu + skew),
               mu_true + skew_true)

frob_norm_diff(vargamma_est$mu, mu_true)
frob_norm_diff(vargamma_est$skew, skew_true)

for(i in 1:3) {
  true_scaled <- true_sigmas[[i]] / sum(diag(true_sigmas[[i]]))
  est_scaled <- vargamma_est$sigmas[[i]] / sum(diag(vargamma_est$sigmas[[i]]))
  
  print(frob_norm(true_scaled - est_scaled) / frob_norm(true_scaled))
}

vargamma_est$gamma
```

## Tensor variate skewed t

```{r}
nu_true <- 20
skew_true <- array(c(
   0.6,  0.3, -0.2,  0.5,
   0.4, -0.5,  0.2, -0.3,
  -0.4,  0.7,  0.1, -0.6,

   0.5, -0.2,  0.3,  0.4,
  -0.3,  0.6, -0.4,  0.2,
   0.1, -0.5,  0.7, -0.1
), dim = c(3, 4, 2))

skewt_draws <- rtskewt(n = 1e3, mu = mu_true, skew = skew_true,
                       sigmas = list(S1, S2, S3), nu = nu_true)
```


```{r}
skewt_est <- tensor_mle(skewt_draws, model = "skewt",
                        quiet = FALSE, tol = 1e-3)

frob_norm_diff(with(skewt_est, mu + (nu)/(nu-2) * skew),
               mu_true + nu_true / (nu_true - 2) * skew_true)

for(i in 1:3) {
  true_scaled <- true_sigmas[[i]] / sum(diag(true_sigmas[[i]]))
  est_scaled <- skewt_est$sigmas[[i]] / sum(diag(skewt_est$sigmas[[i]]))
  
  print(frob_norm_diff(est_scaled, true_scaled))
}
```



## Assessing tensor variate normality

We can use the Mahalanobis distance to assess the tensor variate normality. If we compute the distances and plot them, they should be distributed as
approximately $\chi_{p}$ where $p$ is the product of all the modes of the draws.

```{r}
A <- matrix(rnorm(6^2), 6, 6)
s_vec <- crossprod(A)   # positive definite, general covariance

vec_draws <- rtnorm(n = 1e3, mu = 1:6, sigmas = list(s_vec))

matrix_nocovar_draws <- lapply(1:1e3, function(i) {
  array(vec_draws[[i]], dim = c(2, 3))
})
```


```{r}
plot_malanobis <- function(draws, title = "Dist") {
  distances <- mahalanobis_dist(draws)
  test <- mahalanobis_test(distances)
  plot_dist(distances)
  title(main = title)
  with(test, mtext(text = sprintf("D = %.3f, p = %.3f", statistic, p.value)), 
       side = 3, line = 1)
}
```


```{r}
par(mfcol = c(1, 4))
plot_malanobis(multivar_draws, title = "Multivar")
plot_malanobis(matrix_draws, title = "Mat norm")
plot_malanobis(matrix_nocovar_draws, title = "Mat no covar norm")
plot_malanobis(tvn_draws, title = "Tensor norm")
```






## Tensor Variate Skewed T


```{r}
tibble(skew25 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 0.25, nu = 6)),
       skew50 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 0.5, nu = 6)),
       skew0 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 0, nu = 6)),
       skew1 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 6)),
       skew5 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 5, nu = 6)),
       skew10 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 10, nu = 6))) |>
  pivot_longer(cols = everything()) |> 
  mutate(name = factor(name, levels = c("skew0", "skew1", 
                                     "skew5", "skew10", "skew25", "skew50"))) |> 
  ggplot() +
  geom_histogram(aes(x = value), color = "black", fill = "pink") +
  facet_wrap(~name, nrow = 2, scales = "free_x") +
  theme_minimal()

tibble(nu05 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 0.5)),
       nu1 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 1)),
       nu5 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 5)),
       nu10 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 10)),
       nu20 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 20)),
       nu50 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 50))) |>
  pivot_longer(cols = everything()) |>
  mutate(name = factor(name, levels = c("nu05", "nu1", "nu5", "nu10", "nu20", "nu50"))) |> 
  ggplot() +
  geom_histogram(aes(x = value), color = "black", fill = "pink") +
  facet_wrap(~name, nrow = 2, scales = "free_x") +
  theme_minimal()
```








## Installation

You can install the development version of tensormodels from
[GitHub](https://github.com/) with:
    
```{r}
#| eval: false
if (!requireNamespace("remotes")) install.packages("remotes")
remotes::install_github("ankan12/tensormodels")
```


Reset digits

```{r}
options(digits = old_digits)
```





