---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tensor_models
Tensor operations, decompositions, and distributions in R. 


## Installation

You can install the development version of tensormodels from [GitHub](https://github.com/) with:
    
```{r, eval = FALSE}
#install.packages("devtools")
devtools::install_github("ankan12/tensormodels")
```

```{r, echo = FALSE}
library(tensormodels)
```

# Operations

# n-mode prod

The n-mode product between a tensor and a matrix can be computed with `n_mode_prod`. 

```{r}
a <- array(1:3, dim = c(3, 1, 1))
b <- matrix(4:9, nrow = 2, ncol = 3)
n_mode_prod(a, b, 1)
```

# nm-mode prod

The nm-mode product between a tensor and a tensor can be computed with `nm-mode-prod`.

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)
b <- matrix(c(5, 6), nrow = 2)
nm_prod(A, b, 1, 1)
```

# tensor prod

The tensor product between a tensor and a tensor can be computed with `tensor_prod`. 

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)
b <- matrix(c(5, 6), nrow = 2)
tensor_prod(A, b)
```

# Simulating Tensor Variate Normal Draws

The function `rtnorm` simulates from the tensor variate normal with a specified mean array mu and a list of covariance matrices called list_sigmas. 

Since the univariate normal and matrix variate normal are simpler cases of the tensor variate normal, this function can simulate from them as well.  Here, we simulate from a univariate normal with mean $-2$ and variance $4$. 
```{r}
univar_draws <- rtnorm(n = 10000, mu = -2, list_sigmas = 4)
mean(univar_draws)
sd(univar_draws)
```


And now we simulate from the matrix variate normal of size $2 \times 3$. By default, the covariance matrices will be the identity. 
```{r}
matrix_draws <- rtnorm(n = 10000, mu = matrix(1:6, nrow = 2, ncol = 3))

matrix_draws[1, , ]
```


Below is a simulation from a tensor variate normal of size $3 \times 4 \times 2$. 
```{r}
tensor_draws <- rtnorm(n = 1e3, mu = array(1:24, dim = c(3, 4, 2)))

tensor_draws[1, , , ]
```


# MLE Estimation for the Tensor Variate Normal 

The package supports MLE estimation for the mean and list of covariance matrices. 
```{r}
univar_est <- mle_est(univar_draws)

univar_est$mu
univar_est$sigmas
```



Here, the mean matrix from the draws above has values 1 through 6. The covariance matrices are the identity matrices. 
```{r}
matrix_est <- mle_est(matrix_draws)

matrix_est$mu
matrix_est$sigmas
```


This function works on array objects of dimensions above 2. 
```{r}
tensor_est <- mle_est(tensor_draws)

tensor_est$mu
tensor_est$sigmas
```
