---
output: github_document
bibliography: Bibliography.bib
csl: nature.csl
header-includes: 
  - |
    \AtBeginDocument{
      \renewcommand{\r}[1]{\textcolor{red}{#1}}
      \renewcommand{\b}[1]{\textcolor{blue}{#1}}
    }
  - \usepackage{xcolor}
  - \usepackage{algorithm}
  - \usepackage{algpseudocode}
  - \usepackage{mathtools}
  - \usepackage{float}
  - \newcommand{\coleq}{\mathrel{\mathop:}=}
  - \newcommand{\ve}[1]{\textbf{#1}}
  - \newcommand{\field}[1]{\mathbb{#1}}
  - \newcommand{\defn}[1]{\emph{#1}}
  - \newcommand{\oneVect}[1]{#1}
  - \newcommand{\mat}[1]{\textbf{#1}}
  - \newcommand{\oneMat}[1]{#1}
  - \newcommand{\tensor}[1]{\mathcal{#1}}
  - \newcommand{\oneTensor}[1]{#1} 
  - \newcommand{\vs}[1]{\mathcal{#1}}
  #- \newcommand{\mn}[1]{{#1} \choose {#2}}
  #- \newcommand{\mn}[2]{\mathrel{\underset{\scriptstyle {#1}\,{#2}}{\times}}}
  #- \newcommand{\mn}[2]{\underset{\scriptsize\color{red}{#1}\,\color{blue}{#2}}{\times}}
  #- \newcommand{\mn}[2]{\mathbb{#1}}
  #- \newcommand{\mn}[2]{\sideset{_{\color{red}{#1}}}{_{\color{blue}{#2}}}{\boldsymbol{\times}}}
  - \newcommand{\mn}[2]{\sideset{_{#1}}{_{#2}}{\mathop{\boldsymbol{\times}}}}
  - \newtheorem{proposition}{Proposition}[section]
  - \newtheorem{proof}{Proof}[section]
---




```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
set.seed(1)
```

# **tensormodels**

Tensor operations, decompositions, and distributions in R. 

```{r}
#| message: false
library("tensormodels")
library("tidyverse")
library("tictoc")
old_digits <- getOption("digits")
options(digits = 3)
```







# Operations



## n-mode prod

The $n$-mode matrix product of an $O$th order tensor $\tensor{A} \in \field{k}^{I_1 \times I_2 \times \dots \times I_{n} \times \dots \times I_{O}}$ with a matrix $\mat{U} \in \field{k}^{J \times I_n}$  
is
$$(\tensor{A} \times_n \mat{U}) \in \field{k}^{I_1 \times \dots \times I_{n-1} \times J \times I_{n+1} \times \dots \times I_{O}}$$ with entries 
$$(\tensor{A} \times_n \mat{U})_{i_1, \dots, i_{n-1}, j, i_{n+1}, \dots i_{O}} = \sum_{i_n = 1}^{I_{N}} a_{i_{1}, i_{2}, \dots, i_{O}} u_{j, i_n}.$$ 
The tensor and matrix share one mode in common, denoted here as $I_{n}$. This operation is also called the tensor times matrix product.

The n-mode product between a tensor and a matrix can be computed with `n_mode_prod()`{.R}.

```{r}
a <- array(1:3, dim = c(3, 1, 1))
x <- matrix(4:9, nrow = 2, ncol = 3)

n_prod(a, x, 1)
```





## nm-mode prod

The $nm$-mode product between a tensor and a tensor can be computed with
`nm_prod()`{.R}. The $nm$-mode matrix product of an $O$-th order tensor $\tensor{A} \in \field{k}^{I_{1} \times I_{2} \times \dots \times I_{n} \times \dots \times I_{O}}$ with a $P$-th order tensor $\mat{U} \in \field{k}^{J_{1} \times {J_2} \times \dots \times J_{m} = I_{n} \dots \times J_{P}}$  
is
$$(\tensor{A} \mn{n}{m} \tensor{U}) \in \field{k}^{I_1 \times \dots \times I_{n-1} \times I_{n+1} \times \dots \times I_{O} \times J_{1} \times \dots \times J_{m-1} \times J_{m+1} \times \dots \times J_{P}}$$ with entries 
$$(\tensor{A} \mn{n}{m} \tensor{U})_{i_1, \dots, i_{n-1}, k, i_{n+1}, \dots I_{O}, j_{1}, \dots, j_{m-1}, j_{m+1}, \dots j_{P}} = \sum_{i_n = 1}^{I_n} a_{i_1, i_2, \dots, i_{n}, \dots i_N} u_{j_{1}, j_{2}, \dots i_n, \dots j_{P}}.$$ 

The $n$th mode of the first tensor matches the $m$th mode of the second tensor, denoted here as $I_{n} = J_{m}$. This operation is also called the tensor times tensor product.

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)
x <- matrix(c(5, 6), nrow = 2)

nm_prod(A, x, 1, 1)
```




## [Tensor product](https://en.wikipedia.org/wiki/Tensor_product)

The tensor product between a tensor and a tensor can be computed with
`tensor_prod()`{.R}. Take 2 tensors $\tensor{A} \in \field{k}^{I_{1} \times I_{2} \times \dots \times I_{O}}$ and $\tensor{U} \in \field{k}^{J_{1} \times J_{2} \times \dots \times J_{P}}$

Then the tensor product is $$(\tensor{A} \otimes \tensor{B}) \in \field{k}^{I_{1} \times I_{2} \dots \times I_{O} \times J_{1} \times \dots \times J_{P}}$$ with entries 

$$(\tensor{A} \otimes \tensor{U})_{i_1, \dots, i_{O}, j_{1}, \dots, j_{P}} = a_{i_1, i_2, \dots, i_{o}} u_{j_{1}, j_{2}, \dots \dots j_{p}}.$$ 

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)
x <- matrix(c(5, 6), nrow = 2)

tensor_prod(A, x)
```






# Simulating tensor variate normal draws

The density function of the multilinear normal distribution is @tensornormprop
$$f(x) = (2\pi)^{-p*/2} \left(\prod_{i=1}^{k} |\Sigma_i|^{-p*/(2\pi)}\right) \exp\left\{-\frac{1}{2} (x-\mu)^{T} \Sigma_{1:k}^{-1} (x-\mu)\right\}$$
where $\Sigma$ is positive definite, $x \in \mathbb{R}^p$, $\mu \in \mathbb{R}^p$ and $\Sigma_{1:k} \in \mathbb{R}^p.$

The function `rtnorm()`{.R} simulates random draws from the tensor variate normal with a specified mean array mu and a list of covariance matrices called list_sigmas.

Since the univariate normal and matrix variate normal are simpler cases of the
tensor variate normal, this function can simulate from them as well.  Here, we
simulate random draws from a univariate normal with mean $-2$ and variance $4$.


```{r}
univar_draws <- rtnorm(n = 1000, mu = -2, 
                       sigmas = list(matrix(4)))

mean(simplify2array(univar_draws))

var(simplify2array(univar_draws))
```


We can also simulate random draws from a multivariate normal distribution by specifying a mean vector and the covariance matrix. The dimensions of the draws are dependent on the $\mu$ provided and the corresponding covariance matrices $\mathbf{\Sigma}$ must conform to the $\mu$ provided.

```{r}
S1 <- crossprod(matrix(data = c(1, 0.5, 0.5, 1), nrow = 2))

multivar_draws <- rtnorm(n = 1000, mu = c(2, 3), sigmas = list(S1))

multivar_draws[1:5]
```


And now we simulate from the matrix variate normal of size $2 \times 3$. By
default, the covariance matrices will be the identity.

```{r}
matrix_draws <- rtnorm(n = 1e3, mu = matrix(1:6, nrow = 2, ncol = 3))

matrix_draws[[1]]
```


Below is a simulation from a tensor variate normal of size $3 \times 4 \times
2$.The covariance structure is specified with a list of matrices. 

```{r}
mu_true <- array(1:24, dim = c(3, 4, 2))

S1 <- crossprod(matrix(rnorm(9), nrow = 3))
S2 <- crossprod(matrix(rnorm(16), nrow = 4))
S3 <- crossprod(matrix(rnorm(4), nrow = 2))

tvn_draws <- rtnorm(n = 1e3, mu = mu_true,
                    sigmas = list(S1, S2, S3))

tvn_draws[[1]]
```



# Estimation

## MLE estimation for the tensor variate normal 

The package supports MLE estimation. Given an array of draws, it will return a list containing the MLE for the mean and covariance matrices. Earlier, we generated univariate normal draws with $\mu = 2$ and $\sigma^{2} = 4.$ Calling the function  `tensor_mle()`{.R} will return MLEs similar to the true results. 

```{r}
(univarnorm_est <- tensor_mle(draws = univar_draws, model = "normal"))
```


The multivariate normal draws had a mean vector of $\begin{pmatrix} 2 \\ 3 \end{pmatrix}$ and a covariance matrix $\begin{pmatrix} 1.25 & 1 \\ 1 & 1.25 \end{pmatrix}.$ 

```{r}
(multivarnorm_est <- tensor_mle(draws = multivar_draws, model = "normal"))
```


For the matrix variate draws, the mean matrix was $\begin{pmatrix} 1 & 3 & 5 \\ 2 & 4 & 6 \end{pmatrix}$ and the covariance matrices were the identity.


```{r}
matrix_est <- tensor_mle(matrix_draws, model = "normal")

matrix_est$mu
matrix_est$sigmas |> lapply(round, 3)
```


The `tensor_mle()`{.R} function also works for tensor-valued data. The function `frob_norm_diff()`{.R} computes the relative difference in the Frobenius norm between the true and simulated values. It can be interpreted as the percent error in reconstructing the true parameters.  

Also, note that there is nonidentifiability in the scaling of the covariance matrices. Given the tensor variate normal draws, we know the kronecker of all of the covariances, but not the scaling of each one. To compare the true and estimated covariance matrices, we first trace normalization. This removes arbitrary scale differences while preserving the relative covariance structure within each mode.


```{r}
tensor_est <- tensor_mle(draws = tvn_draws, model = "normal")

frob_norm_diff(tensor_est$mu, mu_true)

true_sigmas <- list(S1, S2, S3)

for(i in 1:3) {
  true_scaled <- true_sigmas[[i]] / sum(diag(true_sigmas[[i]]))
  est_scaled <- tensor_est$sigmas[[i]] / sum(diag(tensor_est$sigmas[[i]]))
  
  print(frob_norm_diff(est_scaled, true_scaled))
}
```






# Other models

To generate tensor variate skewed distributions, we will use the normal variance mean mixture model \citep{variancemeanmix} An $r$-dimensional random vector $x$ is a normal variance-mean mixture with mixing distribution $F$ if, for a given $u \ge 0$ that follows a probability distribution $F$ on $[0, \infty)$, $x|u \sim N_r(\mu + u \beta, u \Sigma)$. $\Sigma \mathbb{R}^{r \times r}$ is a constant, positive-definite matrix and $\mu \in \mathbb{R}^{r}$ and $\beta \in \mathbb{R}^{r}$ are constant vectors.

We state that $x$ is a normal variance-mean mixture with position $\mu$, drift $\beta$, structure matrix $\Sigma$, and mixing distribution $F.$ Different mixing distributions lead to different distributions. For example, if $F$, the mixing distribution, is the generalized inverse Gaussian distribution GIG($\lambda, \delta^2, \kappa^2$) then the resulting distribution of $x$ is a generalized hyperbolic distribution.

In the tensor variate case, the normal variance mean mixture model is an efficient way to introduce skewness. A random tensor $\tensor{X}$ can be written as $\tensor{X} = \tensor{M} + \mat{W} \tensor{A} + \sqrt{\mat{W}} \tensor{V}.$

## Tensor variate normal inverse Gaussian 

The density function of the normal inverse Gaussian distribution is

$$
f_{\text{TVNIG}}(\mathcal{X}|\mathbf{V}) =
\frac{
2 \exp\left\{\text{vec}(\mathcal{X} - \mathcal{M})^{T}
\bigotimes_{d=1}^{D} \Sigma_{d}^{-1}
\text{vec}(\mathcal{A} + \kappa) \right\}
}{
(2\pi)^{\frac{n^{*}}{2}}
\prod_{d=1}^{D} |\Sigma_{d}|^{\frac{n^{*}}{2n_{d}}}
}
\left(
\frac{
\delta(\mathcal{X}; \mathcal{M}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + 1
}{
\rho(\mathcal{A}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + \kappa^2
}
\right)^{-\frac{1 + n^{*}}{4}}
$$ 

$$
\quad K_{- \frac{1 + n^{*}}{2}}
\left(
\sqrt{
\left[\rho(\mathcal{A}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + \kappa^{2}\right]
\left[\delta(\mathcal{X}; \mathcal{M}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + 1\right]
}
\right)
$$
where $\Sigma$ is positive definite, $x \in \mathbb{R}^p$, $\mu \in \mathbb{R}^p$ and $\Sigma_{1:k} \in \mathbb{R}^p.$

The function `rtinvgauss()`{.R} simulates random draws from the tensor variate normal inverse Gaussian with a specified mean array mu, a list of covariance matrices, a skew array, and $\kappa$ which describes the shape of the inverse Gaussian distribution


```{r}
mu_true <- array(1:24, dim = c(3, 4, 2))
skew_true <- array(seq(0, 4, length.out = 24), dim = c(3, 4, 2))
kappa_true <- 2

invgauss_draws <- rtinvgauss(n = 1e3, mu = mu_true, skew = skew_true, 
                             sigmas = list(S1, S2, S3), kappa = kappa_true)
```


```{r}
invgauss_est <- tensor_mle(invgauss_draws, model = "invgauss", 
                           quiet = FALSE, tol = 1e-3)
```


For the inverse gaussian distribution, the true mean $E[X] = \tensor{M} + \tensor{A}/\kappa$. We can compare the model's estimation for the mean with the true values. 

```{r}
frob_norm_diff(with(invgauss_est, mu + skew/kappa),
               mu_true + skew_true/kappa_true)
```


```{r}
frob_norm_diff(invgauss_est$mu, mu_true)
frob_norm_diff(invgauss_est$skew, skew_true)
invgauss_est$kappa

for(i in 1:2) {
  true_scaled <- true_sigmas[[i]] / sum(diag(true_sigmas[[i]]))
  est_scaled <- invgauss_est$sigmas[[i]] / sum(diag(invgauss_est$sigmas[[i]]))
  
  print(frob_norm_diff(est_scaled, true_scaled))
}
```




## Tensor variate generalized hyperbolic

The density function of the generalized hyperbolic distribution is
$$f(x) = (2\pi)^{-p*/2} \left(\prod_{i=1}^{k} |\Sigma_i|^{-p*/(2\pi)}\right) \exp\left\{-\frac{1}{2} (x-\mu)^{T} \Sigma_{1:k}^{-1} (x-\mu)\right\}$$
where $\Sigma$ is positive definite, $x \in \mathbb{R}^p$, $\mu \in \mathbb{R}^p$ and $\Sigma_{1:k} \in \mathbb{R}^p.$

The function `rtgenhyper()`{.R} simulates random draws from the tensor variate generalized hyperbolic with a specified mean array mu and a list of covariance matrices called list_sigmas.

Since the univariate normal and matrix variate normal are simpler cases of the
tensor variate normal, this function can simulate from them as well.  Here, we
simulate random draws from a univariate normal with mean $-2$ and variance $4$.

```{r}
lambda_true <- 2
omega_true <- 2

genhyper_draws <- rtgenhyper(n = 1e3, mu = mu_true, skew = skew_true,
                             sigmas = list(S1, S2, S3))
```



For the generalized hyperbolic distribution, the true mean $E[X] = \tensor{M} + \frac{K_{\lambda + 1}(\omega)}{K_{\lambda}(\omega)} \tensor{A}$. We can compare the model's estimation for the mean with the true values. 

```{r}
genhyper_est <- tensor_mle(genhyper_draws, model = "genhyper", quiet = FALSE)

frob_norm_diff(
  with(genhyper_est, mu + besselK(x = omega, nu = lambda + 1)/
                          besselK(x = omega, nu = lambda) * skew),
       mu_true + besselK(x = omega_true, nu = lambda_true + 1)/
                 besselK(x = omega_true, nu = lambda_true) * skew_true)

frob_norm_diff(genhyper_est$mu, mu_true)
frob_norm_diff(genhyper_est$skew, skew_true)

for(i in 1:3) {
  true_scaled <- true_sigmas[[i]] / sum(diag(true_sigmas[[i]]))
  est_scaled <- genhyper_est$sigmas[[i]] / sum(diag(genhyper_est$sigmas[[i]]))
  
  print(frob_norm_diff(est_scaled, true_scaled))
}

genhyper_est$lambda
genhyper_est$omega
```


## Tensor variate variance gamma
```{r}
vargamma_draws <- rtvargamma(n = 1e3, mu = mu_true, skew = skew_true,
                             sigmas = list(S1, S2, S3))
```



```{r}
vargamma_est <- tensor_mle(vargamma_draws, model = "vargamma", quiet = FALSE)

frob_norm_diff(with(vargamma_est, mu + skew),
               mu_true + skew_true)

frob_norm_diff(vargamma_est$mu, mu_true)
frob_norm_diff(vargamma_est$skew, skew_true)

for(i in 1:3) {
  true_scaled <- true_sigmas[[i]] / sum(diag(true_sigmas[[i]]))
  est_scaled <- vargamma_est$sigmas[[i]] / sum(diag(vargamma_est$sigmas[[i]]))
  
  print(frob_norm_diff(est_scaled, true_scaled))
}

vargamma_est$gamma
```

## Tensor variate skewed t

```{r}
nu_true <- 20
skew_true <- array(c(
   0.6,  0.3, -0.2,  0.5,
   0.4, -0.5,  0.2, -0.3,
  -0.4,  0.7,  0.1, -0.6,

   0.5, -0.2,  0.3,  0.4,
  -0.3,  0.6, -0.4,  0.2,
   0.1, -0.5,  0.7, -0.1
), dim = c(3, 4, 2))

skewt_draws <- rtskewt(n = 1e3, mu = mu_true, skew = skew_true,
                       sigmas = list(S1, S2, S3), nu = nu_true)
```


```{r}
skewt_est <- tensor_mle(skewt_draws, model = "skewt",
                        quiet = FALSE, tol = 1e-3)

frob_norm_diff(with(skewt_est, mu + (nu)/(nu-2) * skew),
               mu_true + nu_true / (nu_true - 2) * skew_true)

for(i in 1:3) {
  true_scaled <- true_sigmas[[i]] / sum(diag(true_sigmas[[i]]))
  est_scaled <- skewt_est$sigmas[[i]] / sum(diag(skewt_est$sigmas[[i]]))
  
  print(frob_norm_diff(est_scaled, true_scaled))
}
```



## Assessing tensor variate normality

We can use the Mahalanobis distance to assess the tensor variate normality. If we compute the distances and plot them, they should be distributed as
approximately $\chi_{p}$ where $p$ is the product of all the modes of the draws.

```{r}
A <- matrix(rnorm(6^2), 6, 6)
s_vec <- crossprod(A)   # positive definite, general covariance

vec_draws <- rtnorm(n = 1e3, mu = 1:6, sigmas = list(s_vec))

matrix_nocovar_draws <- lapply(1:1e3, function(i) {
  array(vec_draws[[i]], dim = c(2, 3))
})
```


```{r}
plot_malanobis <- function(draws, title = "Dist") {
  distances <- mahalanobis_dist(draws)
  test <- mahalanobis_test(distances)
  plot_dist(distances)
  title(main = title)
  with(test, mtext(text = sprintf("D = %.3f, p = %.3f", statistic, p.value)), 
       side = 3, line = 1)
}
```


```{r}
par(mfcol = c(1, 4))
plot_malanobis(multivar_draws, title = "Multivar")
plot_malanobis(matrix_draws, title = "Mat norm")
plot_malanobis(matrix_nocovar_draws, title = "Mat no covar norm")
plot_malanobis(tvn_draws, title = "Tensor norm")
```

We are also interested in performing a likelihood ratio test. We are curious if one of the covariance matrices might be the identity matrix. Thus, we run the algorithm and restrict these matrices. 

For a restriction set \(\mathcal{R} \subset \{1,\ldots,D\}\), the restricted model fixes \(\Sigma_r = I_{n_r}\) for every \(r \in \mathcal{R}\), while all other covariance matrices are estimated. The likelihood ratio test compares 

$$H_0: \Sigma_r = I_{n_r} \ \forall r \in \mathcal{R} \quad \text{vs.} \quad H_A: \Sigma_r \neq I_{n_r} \ \text{for at least one } r \in \mathcal{R}.$$

The test statistic is 

$$2\{\ell(\hat\theta_{\mathrm{full}}) - \ell(\hat\theta_{\mathcal{R}})\}
\sim \chi^2_{\nu_{\mathcal{R}}}, \quad \nu_{\mathcal{R}} = \sum_{r \in \mathcal{R}} \left\{\frac{n_r(n_r+1)}{2} - 1\right\}$$

If the $p$-value is large, then we fail to reject the null hypothesis and conclude the restricted model fits the data as well as the unrestricted model. If the $p$-value is small, we reject the null hypothesis, which states the unrestricted model has a significantly better fit than the restricted model.

```{r}
true_str <- kronecker(S3, S2) |> kronecker(S1)
  
test_restrict <- function(draws, restrict) {
  curr_restrict <- tensor_mle(draws, restrict = restrict, model = "normal")
  
  if(is.null(restrict)) restrict <- "None"

  curr_str <- with(curr_restrict, kronecker(sigmas[[3]], sigmas[[2]]) |>
                                  kronecker(sigmas[[1]]))

  tibble(frob_error = frob_norm_diff(curr_str, true_str),
         loglik = curr_restrict$loglik,
         k = curr_restrict$k,
         BIC = curr_restrict$BIC,
         restrict = paste(restrict, collapse = ", "))
}

(res <- map_dfr(list(NULL, c(1), c(2), c(3), c(1, 2), c(1, 3), c(2, 3)),
        draws = tvn_draws, test_restrict))
```


```{r}
lrt_test <- function(draws, restrict) {
  full_model <- tensor_mle(draws, model = "normal")
  restrict_model <- tensor_mle(draws, model = "normal", restrict = restrict)
  
  pval <- pchisq(2 * (full_model$loglik - restrict_model$loglik), 
                 df = full_model$k - restrict_model$k, lower.tail = FALSE)
  
  tibble(restrict = paste(restrict, collapse = ", "), 
         pval = pval)
}

map_dfr(list(NULL, c(1), c(2), c(3), c(1, 2), c(1, 3), c(2, 3)),
        draws = tvn_draws, lrt_test)
```


```{r}
true_str <- kronecker(S3, diag(4)) |> kronecker(S1)

second_identity <- rtnorm(n = 1e3, mu = mu_true, 
                          sigmas = list(S1, diag(4), S3))

map_dfr(list(NULL, c(1), c(2), c(3), c(1, 2), c(1, 3), c(2, 3)), 
        draws = second_identity, test_restrict)

map_dfr(list(NULL, c(1), c(2), c(3), c(1, 2), c(1, 3), c(2, 3)),
        draws = second_identity, lrt_test)
```





## Tensor Variate Skewed T


```{r}
tibble(skew25 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                               skew = matrix(0.25, nrow = 2, ncol = 3), nu = 6)),
       skew50 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                               skew = matrix(0.5, nrow = 2, ncol = 3), nu = 6)),
       skew0 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                              skew = matrix(0, nrow = 2, ncol = 3), nu = 6)),
       skew1 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                              skew = matrix(1, nrow = 2, ncol = 3), nu = 6)),
       skew5 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                              skew = matrix(5, nrow = 2, ncol = 3), nu = 6)),
       skew10 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                               skew = matrix(10, nrow = 2, ncol = 3), nu = 6))) |>
  pivot_longer(cols = everything()) |> 
  mutate(name = factor(name, levels = c("skew0", "skew1", 
                                     "skew5", "skew10", "skew25", "skew50"))) |> 
  ggplot() +
  geom_histogram(aes(x = value), color = "black", fill = "pink") +
  facet_wrap(~name, nrow = 2, scales = "free_x") +
  theme_minimal()

tibble(nu05 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                             skew = matrix(1, nrow = 2, ncol = 3), nu = 0.5)),
       nu1 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                            skew = matrix(1, nrow = 2, ncol = 3), nu = 1)),
       nu5 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                            skew = matrix(1, nrow = 2, ncol = 3), nu = 5)),
       nu10 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                             skew = matrix(1, nrow = 2, ncol = 3), nu = 10)),
       nu20 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                             skew = matrix(1, nrow = 2, ncol = 3), nu = 20)),
       nu50 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                             skew = matrix(1, nrow = 2, ncol = 3), nu = 50))) |>
  pivot_longer(cols = everything()) |>
  mutate(name = factor(name, levels = c("nu05", "nu1", "nu5", "nu10", "nu20", "nu50"))) |> 
  ggplot() +
  geom_histogram(aes(x = value), color = "black", fill = "pink") +
  facet_wrap(~name, nrow = 2, scales = "free_x") +
  theme_minimal()
```








## Installation

You can install the development version of tensormodels from
[GitHub](https://github.com/) with:
    
```{r}
#| eval: false
if (!requireNamespace("remotes")) install.packages("remotes")
remotes::install_github("ankan12/tensormodels")
```


Reset digits

```{r}
options(digits = old_digits)
```




