---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tensor_models
Tensor operations, decompositions, and distributions in R. 


## Installation

You can install the development version of tensormodels from [GitHub](https://github.com/) with:
    
```{r, eval = FALSE}
#install.packages("devtools")
#devtools::install_github("ankan12/tensormodels")
```

```{r, echo = FALSE}
library(tensormodels)
library(tidyverse)
```

# Operations

# n-mode prod

The n-mode product between a tensor and a matrix can be computed with `n_mode_prod`. 

```{r}
a <- array(1:3, dim = c(3, 1, 1))
x <- matrix(4:9, nrow = 2, ncol = 3)
n_prod(a, x, 1)
```

# nm-mode prod

The nm-mode product between a tensor and a tensor can be computed with `nm-mode-prod`.

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)
x <- matrix(c(5, 6), nrow = 2)
nm_prod(A, x, 1, 1)
```

# tensor prod

The tensor product between a tensor and a tensor can be computed with `tensor_prod`. 

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)
x <- matrix(c(5, 6), nrow = 2)
tensor_prod(A, x)
```

# Simulating Tensor Variate Normal Draws

The function `rtnorm` simulates from the tensor variate normal with a specified mean array mu and a list of covariance matrices called list_sigmas. 

Since the univariate normal and matrix variate normal are simpler cases of the tensor variate normal, this function can simulate from them as well.  Here, we simulate from a univariate normal with mean $-2$ and variance $4$. 
```{r}
univar_draws <- rtnorm(n = 10000, mu = -2, sigmas = 4)
mean(univar_draws)
sd(univar_draws)
```


And now we simulate from the matrix variate normal of size $2 \times 3$. By default, the covariance matrices will be the identity. 
```{r}
matrix_draws <- rtnorm(n = 1e3, mu = matrix(1:6, nrow = 2, ncol = 3))

matrix_draws[1, , ]
```


Below is a simulation from a tensor variate normal of size $3 \times 4 \times 2$. 
```{r}
tensor_draws <- rtnorm(n = 1e3, mu = array(1:24, dim = c(3, 4, 2)))

tensor_draws[1, , , ]
```


# MLE Estimation for the Tensor Variate Normal 

The package supports MLE estimation for the mean and list of covariance matrices. 
```{r}
univar_est <- mle_est(univar_draws)

univar_est$mu
univar_est$sigmas
```



Here, the mean matrix from the draws above has values 1 through 6. The covariance matrices are the identity matrices by default. 
```{r}
matrix_est <- mle_est(matrix_draws)

matrix_est$mu
matrix_est$sigmas
```


This function works on array objects of dimensions above 2. 
```{r}
tensor_est <- mle_est(tensor_draws)

tensor_est$mu
tensor_est$sigmas
```


# Tensor Variate Generalized Hyperbolic
```{r}
genhyper_draws <- rtgenhyper(n = 1e3)
```


# Tensor Variate Variance Gamma
```{r}
vargamma_draws <- rtvargamma(n = 1e3)
```


# Tensor Variate Normal Inverse Gaussian 
```{r}
invgauss_draws <- rtinvgauss(n = 1e3)
```


# Assessing Tensor Variate Normality

We can use the Mahalanobis distance to assess the tensor variate normality. If we compute the distances and plot them, they should be distributed as approximately $\chi_{p}$ where $p$ is the product of all the modes of the draws.
```{r}
centered_draws <- matrix_draws - array(rep(matrix_est$mu, each = n), dim = dim(matrix_draws))

inv_sqrt <- function(Sigma, tol = 1e-8) {
  # syemmetric matrix
  Sigma <- (Sigma + t(Sigma)) / 2
  
  eig <- eigen(Sigma, symmetric = TRUE)
  vals <- pmax(eig$values, tol)  # positive eigenvalues
  V <- eig$vectors
  
  Winv <- V %*% diag(1 / sqrt(vals), nrow = length(vals)) %*% t(V)
  Winv
}

D2 <- rep(0, n)

inv_sigmas <- vector("list", length(matrix_est$sigmas))

for(k in 1:2) {
  inv_sigmas[[k]] <- inv_sqrt(matrix_est$sigmas[[k]])
}

for(i in 1:n) {
  Y <- centered_draws[i, , ]
  for(k in 1:2) {
    Y <- n_prod(Y, matrix_est$sigmas[[k]], k)
  }
  D2[i] <- sum(c(Y)^2)
}

sigma2_hat <- mean(D2) / p
D2_std <- D2 / sigma2_hat    

ggplot() +
  geom_histogram(aes(x = D2_std, y = after_stat(density)), color = "black", fill = "pink", bins = 30) + 
  geom_function(fun = dchisq, args = list(df = prod(dim(matrix_est$mu))), color = "red") +
  theme_minimal() + 
  labs(x = "Mahalanobis distance", y = "Density")
```


# Tensor Variate Skewed T


```{r}
tibble(skew25 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 0.25, nu = 6)),
       skew50 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 0.5, nu = 6)),
       skew0 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 0, nu = 6)),
       skew1 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 6)),
       skew5 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 5, nu = 6)),
       skew10 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 10, nu = 6))) |>
  pivot_longer(cols = everything()) |>
  ggplot() +
  geom_histogram(aes(x = value), color = "black", fill = "pink") +
  facet_wrap(~name, nrow = 2, scales = "free_x") +
  theme_minimal()

tibble(nu05 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 0.5)),
       nu1 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 1)),
       nu5 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 5)),
       nu10 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 10)),
       nu20 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 20)),
       nu50 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 50))) |>
  pivot_longer(cols = everything()) |>
  ggplot() +
  geom_histogram(aes(x = value), color = "black", fill = "pink") +
  facet_wrap(~name, nrow = 2, scales = "free_x") +
  theme_minimal()
```
