---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tensor_models
Tensor operations, decompositions, and distributions in R. 


## Installation

You can install the development version of regexcite from [GitHub](https://github.com/) with:
    
```{r, eval = FALSE}
#install.packages("devtools")
devtools::install_github("ankan12/tensormodels")
```

```{r, echo = FALSE}
library(tensormodels)
```

# Operations

# n-mode prod

The n-mode product between a tensor and a matrix can be computed with `n_mode_prod`. 

```{r}
a <- array(1:3, dim = c(3, 1, 1))
b <- matrix(4:9, nrow = 2, ncol = 3)
n_mode_prod(a, b, 1)
```

# nm-mode prod

The nm-mode product between a tensor and a tensor can be computed with `nm-mode-prod`.

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)
b <- matrix(c(5, 6), nrow = 2)
nm_prod(A, b, 1, 1)
```

# tensor prod

The tensor product between a tensor and a tensor can be computed with `tensor_prod`. 

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)
b <- matrix(c(5, 6), nrow = 2)
tensor_prod(A, b)
```

