---
output: github_document
bibliography: Bibliography.bib
csl: nature.csl
header-includes: 
  - |
    \AtBeginDocument{
      \renewcommand{\r}[1]{\textcolor{red}{#1}}
      \renewcommand{\b}[1]{\textcolor{blue}{#1}}
    }
  - \usepackage{xcolor}
  - \usepackage{algorithm}
  - \usepackage{algpseudocode}
  - \usepackage{mathtools}
  - \usepackage{float}
  - \newcommand{\coleq}{\mathrel{\mathop:}=}
  - \newcommand{\ve}[1]{\textbf{#1}}
  - \newcommand{\field}[1]{\mathbb{#1}}
  - \newcommand{\defn}[1]{\emph{#1}}
  - \newcommand{\oneVect}[1]{#1}
  - \newcommand{\mat}[1]{\textbf{#1}}
  - \newcommand{\oneMat}[1]{#1}
  - \newcommand{\tensor}[1]{\mathcal{#1}}
  - \newcommand{\oneTensor}[1]{#1} 
  - \newcommand{\vs}[1]{\mathcal{#1}}
  #- \newcommand{\mn}[1]{{#1} \choose {#2}}
  #- \newcommand{\mn}[2]{\mathrel{\underset{\scriptstyle {#1}\,{#2}}{\times}}}
  #- \newcommand{\mn}[2]{\underset{\scriptsize\color{red}{#1}\,\color{blue}{#2}}{\times}}
  #- \newcommand{\mn}[2]{\mathbb{#1}}
  #- \newcommand{\mn}[2]{\sideset{_{\color{red}{#1}}}{_{\color{blue}{#2}}}{\boldsymbol{\times}}}
  - \newcommand{\mn}[2]{\sideset{_{#1}}{_{#2}}{\mathop{\boldsymbol{\times}}}}
  - \newtheorem{proposition}{Proposition}[section]
  - \newtheorem{proof}{Proof}[section]
---




```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# **tensormodels**

Tensor operations, decompositions, and distributions in R. 

```{r}
#| message: false
library("tensormodels")
library("tidyverse")
old_digits <- getOption("digits")
options(digits = 3)
```







# Operations



## n-mode prod

The $n$-mode matrix product of  a tensor $\tensor{A} \in \field{k}^{I_1 \times I_2 \times ... \times I_n}$ with a matrix $\mat{U} \in \field{k}^{J \times I_n}$  
is
$$(\tensor{A} \times_n \mat{U}) \in \field{k}^{I_1 \times ... \times I_{n-1} \times J \times I_{n+1} \times ... \times I_N}$$ with entries $$(\tensor{A} \times_n \mat{U})_{i_1, ..., i_{n-1}, j, i_{n+1}, ... I_N} = \sum_{i_n = 1}^{I_n} a_{i_1, i_2, \dots, i_N} u_{j, i_n}.$$ The tensor and matrix share one mode in common, denoted here as $I_n$. This operation is also called the tensor times matrix product.

The n-mode product between a tensor and a matrix can be computed with `n_mode_prod()`{.R}.

```{r}
a <- array(1:3, dim = c(3, 1, 1))
x <- matrix(4:9, nrow = 2, ncol = 3)

n_prod(a, x, 1)
```





## nm-mode prod

The nm-mode product between a tensor and a tensor can be computed with
`nm_prod()`{.R}.

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)
x <- matrix(c(5, 6), nrow = 2)

nm_prod(A, x, 1, 1)
```




## [Tensor product](https://en.wikipedia.org/wiki/Tensor_product)

The tensor product between a tensor and a tensor can be computed with
`tensor_prod()`{.R}.

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)
x <- matrix(c(5, 6), nrow = 2)

tensor_prod(A, x)
```






# Simulating tensor variate normal draws

The density function of the multilinear normal distribution is @tensornormprop
$$f(x) = (2\pi)^{-p*/2} \left(\prod_{i=1}^{k} |\Sigma_i|^{-p*/(2\pi)}\right) \exp\left\{-\frac{1}{2} (x-\mu)' \Sigma_{1:k}^{-1} (x-\mu\right\}$$
where $\Sigma$ is positive definite, $x \in \mathbb{R}^p$, $\mu \in \mathbb{R}^p$ and $\Sigma_{1:k} \in \mathbb{R}^p.$

The function `rtnorm()`{.R} simulates from the tensor variate normal with a
specified mean array mu and a list of covariance matrices called list_sigmas.

Since the univariate normal and matrix variate normal are simpler cases of the
tensor variate normal, this function can simulate from them as well.  Here, we
simulate from a univariate normal with mean $-2$ and variance $4$.


```{r}
univar_draws <- rtnorm(n = 1000, mu = -2, sigmas = 4)

mean(simplify2array(univar_draws))

sd(simplify2array(univar_draws))
```


And now we simulate from the matrix variate normal of size $2 \times 3$. By
default, the covariance matrices will be the identity.

```{r}
matrix_draws <- rtnorm(n = 1e3, mu = matrix(1:6, nrow = 2, ncol = 3))

matrix_draws[[1]]
```


Below is a simulation from a tensor variate normal of size $3 \times 4 \times
2$.

```{r}
tvn_draws <- rtnorm(n = 1e3, mu = array(1:24, dim = c(3, 4, 2)))

tvn_draws[[1]]
```



# Estimation

## MLE estimation for the tensor variate normal 

The package supports MLE estimation for the mean and list of covariance
matrices.

```{r}
univarnorm_est <- tensor_mle(draws = univar_draws, model = "normal")

univarnorm_est$mu

univarnorm_est$sigmas
```


Here, the mean matrix from the draws above has values 1 through 6. The
covariance matrices are the identity matrices by default.

```{r}
matrix_est <- tensor_mle(matrix_draws, model = "normal")

matrix_est$mu
matrix_est$sigmas |> lapply(round, 3)
```


This function works on array objects of dimensions above 2. 

```{r}
normal_est <- tensor_mle(draws = tvn_draws, model = "normal")

normal_est$mu
normal_est$sigmas |> lapply(round, 3)
```





# Other models

## Tensor variate generalized hyperbolic

```{r}
genhyper_draws <- rtgenhyper(n = 1e3, mu = array(1:24, dim = c(3, 4, 2)))
```


```{r}
genhyper_est <- tensor_mle(genhyper_draws, model = "genhyper", quiet = FALSE)

genhyper_est$mu
genhyper_est$skew
genhyper_est$sigmas |> lapply(round, 3)
genhyper_est$lambda
genhyper_est$omega
```


## Tensor variate variance gamma
```{r}
vargamma_draws <- rtvargamma(n = 1e3, mu = array(1:24, dim = c(3, 4, 2)))
```



```{r}
vargamma_est <- tensor_mle(vargamma_draws, model = "vargamma", quiet = FALSE)

vargamma_est$mu
vargamma_est$skew
vargamma_est$sigmas |> lapply(round, 3)
```


## Tensor variate normal inverse Gaussian 
```{r}
invgauss_draws <- rtinvgauss(n = 1e3, mu = array(1:24, dim = c(3, 4, 2)))
```


```{r}
invgauss_est <- tensor_mle(invgauss_draws, model = "invgauss", quiet = FALSE)

invgauss_est$mu
invgauss_est$skew |> round(3) 
invgauss_est$sigmas |> lapply(round, 3)
invgauss_est$kappa
```

## Tensor variate skewed t

```{r}
skewt_draws <- rtskewt(n = 1e3, mu = array(1:24, dim = c(3, 4, 2)))
```


```{r}
skewt_est <- tensor_mle(skewt_draws, model = "skewt", 
                        quiet = FALSE, tol = 1e-3)

skewt_est$mu
skewt_est$skew |> round(3) 
skewt_est$sigmas |> lapply(round, 3)
skewt_est$nu

skewt_est$mu + (skewt_est$nu / (skewt_est$nu - 2)) * skewt_est$skew
```


## Assessing tensor variate normality

<!-- We can use the Mahalanobis distance to assess the tensor variate normality. If -->
<!-- we compute the distances and plot them, they should be distributed as -->
<!-- approximately $\chi_{p}$ where $p$ is the product of all the modes of the draws. -->

<!-- ```{r} -->
<!-- #| error: true -->
<!-- centered_draws <- matrix_draws - array(rep(matrix_est$mu, each = n), dim = dim(matrix_draws)) -->

<!-- inv_sqrt <- function(Sigma, tol = 1e-8) { -->
<!--   # syemmetric matrix -->
<!--   Sigma <- (Sigma + t(Sigma)) / 2 -->

<!--   eig <- eigen(Sigma, symmetric = TRUE) -->
<!--   vals <- pmax(eig$values, tol)  # positive eigenvalues -->
<!--   V <- eig$vectors -->

<!--   Winv <- V %*% diag(1 / sqrt(vals), nrow = length(vals)) %*% t(V) -->
<!--   Winv -->
<!-- } -->

<!-- D2 <- rep(0, n) -->

<!-- inv_sigmas <- vector("list", length(matrix_est$sigmas)) -->

<!-- for(k in 1:2) { -->
<!--   inv_sigmas[[k]] <- inv_sqrt(matrix_est$sigmas[[k]]) -->
<!-- } -->

<!-- for(i in 1:n) { -->
<!--   Y <- centered_draws[i, , ] -->
<!--   for(k in 1:2) { -->
<!--     Y <- n_prod(Y, matrix_est$sigmas[[k]], k) -->
<!--   } -->
<!--   D2[i] <- sum(c(Y)^2) -->
<!-- } -->

<!-- sigma2_hat <- mean(D2) / 2 -->
<!-- D2_std <- D2 / sigma2_hat     -->

<!-- ggplot() + -->
<!--   geom_histogram(aes(x = D2_std, y = after_stat(density)), color = "black", fill = "pink", bins = 30) +  -->
<!--   geom_function(fun = dchisq, args = list(df = prod(dim(matrix_est$mu))), color = "red") + -->
<!--   theme_minimal() +  -->
<!--   labs(x = "Mahalanobis distance", y = "Density") -->
<!-- ``` -->


## Tensor Variate Skewed T


```{r}
tibble(skew25 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 0.25, nu = 6)),
       skew50 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 0.5, nu = 6)),
       skew0 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 0, nu = 6)),
       skew1 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 6)),
       skew5 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 5, nu = 6)),
       skew10 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 10, nu = 6))) |>
  pivot_longer(cols = everything()) |> 
  mutate(name = factor(name, levels = c("skew0", "skew1", 
                                     "skew5", "skew10", "skew25", "skew50"))) |> 
  ggplot() +
  geom_histogram(aes(x = value), color = "black", fill = "pink") +
  facet_wrap(~name, nrow = 2, scales = "free_x") +
  theme_minimal()

tibble(nu05 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 0.5)),
       nu1 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 1)),
       nu5 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 5)),
       nu10 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 10)),
       nu20 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 20)),
       nu50 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 50))) |>
  pivot_longer(cols = everything()) |>
  mutate(name = factor(name, levels = c("nu05", "nu1", "nu5", "nu10", "nu20", "nu50"))) |> 
  ggplot() +
  geom_histogram(aes(x = value), color = "black", fill = "pink") +
  facet_wrap(~name, nrow = 2, scales = "free_x") +
  theme_minimal()
```








## Installation

You can install the development version of tensormodels from
[GitHub](https://github.com/) with:
    
```{r}
#| eval: false
if (!requireNamespace("remotes")) install.packages("remotes")
remotes::install_github("ankan12/tensormodels")
```


Reset digits

```{r}
options(digits = old_digits)
```





