---
output: github_document
bibliography: Bibliography.bib
csl: nature.csl
header-includes: 
  - |
    \AtBeginDocument{
      \renewcommand{\r}[1]{\textcolor{red}{#1}}
      \renewcommand{\b}[1]{\textcolor{blue}{#1}}
    }
  - \usepackage{xcolor}
  - \usepackage{algorithm}
  - \usepackage{algpseudocode}
  - \usepackage{mathtools}
  - \usepackage{float}
  - \newcommand{\coleq}{\mathrel{\mathop:}=}
  - \newcommand{\ve}[1]{\textbf{#1}}
  - \newcommand{\field}[1]{\mathbb{#1}}
  - \newcommand{\defn}[1]{\emph{#1}}
  - \newcommand{\oneVect}[1]{#1}
  - \newcommand{\mat}[1]{\textbf{#1}}
  - \newcommand{\oneMat}[1]{#1}
  - \newcommand{\tensor}[1]{\mathcal{#1}}
  - \newcommand{\oneTensor}[1]{#1} 
  - \newcommand{\vs}[1]{\mathcal{#1}}
  #- \newcommand{\mn}[1]{{#1} \choose {#2}}
  #- \newcommand{\mn}[2]{\mathrel{\underset{\scriptstyle {#1}\,{#2}}{\times}}}
  #- \newcommand{\mn}[2]{\underset{\scriptsize\color{red}{#1}\,\color{blue}{#2}}{\times}}
  #- \newcommand{\mn}[2]{\mathbb{#1}}
  #- \newcommand{\mn}[2]{\sideset{_{\color{red}{#1}}}{_{\color{blue}{#2}}}{\boldsymbol{\times}}}
  - \newcommand{\mn}[2]{\sideset{_{#1}}{_{#2}}{\mathop{\boldsymbol{\times}}}}
  - \newtheorem{proposition}{Proposition}[section]
  - \newtheorem{proof}{Proof}[section]
---




```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# **tensormodels**

Tensor operations, decompositions, and distributions in R. 

```{r}
#| message: false
library("tensormodels")
library("tidyverse")
old_digits <- getOption("digits")
options(digits = 3)
```







# Operations



## n-mode prod

The $n$-mode matrix product of  a tensor $\tensor{A} \in \field{k}^{I_1 \times I_2 \times ... \times I_n}$ with a matrix $\mat{U} \in \field{k}^{J \times I_n}$  
is
$$(\tensor{A} \times_n \mat{U}) \in \field{k}^{I_1 \times ... \times I_{n-1} \times J \times I_{n+1} \times ... \times I_N}$$ with entries $$(\tensor{A} \times_n \mat{U})_{i_1, ..., i_{n-1}, j, i_{n+1}, ... I_N} = \sum_{i_n = 1}^{I_n} a_{i_1, i_2, \dots, i_N} u_{j, i_n}.$$ The tensor and matrix share one mode in common, denoted here as $I_n$. This operation is also called the tensor times matrix product.

The n-mode product between a tensor and a matrix can be computed with `n_mode_prod`.

```{r}
a <- array(1:3, dim = c(3, 1, 1))
x <- matrix(4:9, nrow = 2, ncol = 3)

n_prod(a, x, 1)
```





## nm-mode prod

The nm-mode product between a tensor and a tensor can be computed with
`nm_prod()`{.R}.

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)
x <- matrix(c(5, 6), nrow = 2)

nm_prod(A, x, 1, 1)
```




## [Tensor product](https://en.wikipedia.org/wiki/Tensor_product)

The tensor product between a tensor and a tensor can be computed with
`tensor_prod()`{.R}.

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)
x <- matrix(c(5, 6), nrow = 2)

tensor_prod(A, x)
```






# Simulating tensor variate normal draws

The density function of the multilinear normal distribution is @tensornormprop
$$f(x) = (2\pi)^{-p*/2} \left(\prod_{i=1}^{k} |\Sigma_i|^{-p*/(2\pi)}\right) \exp\left\{-\frac{1}{2} (x-\mu)' \Sigma_{1:k}^{-1} (x-\mu\right\}$$
where $\Sigma$ is positive definite, $x \in \mathbb{R}^p$, $\mu \in \mathbb{R}^p$ and $\Sigma_{1:k} \in \mathbb{R}^p.$

The function `rtnorm()`{.R} simulates from the tensor variate normal with a
specified mean array mu and a list of covariance matrices called list_sigmas.

Since the univariate normal and matrix variate normal are simpler cases of the
tensor variate normal, this function can simulate from them as well.  Here, we
simulate from a univariate normal with mean $-2$ and variance $4$.


```{r}
univar_draws <- rtnorm(n = 1000, mu = -2, sigmas = 4)

mean(univar_draws)

sd(univar_draws)
```


And now we simulate from the matrix variate normal of size $2 \times 3$. By
default, the covariance matrices will be the identity.

```{r}
matrix_draws <- rtnorm(n = 1e3, mu = matrix(1:6, nrow = 2, ncol = 3))

matrix_draws[1, , ]
```


Below is a simulation from a tensor variate normal of size $3 \times 4 \times
2$.

```{r}
tensor_draws <- rtnorm(n = 1e3, mu = array(1:24, dim = c(3, 4, 2)))

tensor_draws[1, , , ]
```



# Estimation

## MLE estimation for the tensor variate normal 

The package supports MLE estimation for the mean and list of covariance
matrices.

```{r}
univar_est <- mle_est(univar_draws)

univar_est$mu

univar_est$sigmas
```


Here, the mean matrix from the draws above has values 1 through 6. The
covariance matrices are the identity matrices by default.

```{r}
matrix_est <- mle_est(matrix_draws)

matrix_est$mu

matrix_est$sigmas |> lapply(round, 3)
```


This function works on array objects of dimensions above 2. 

```{r}
tensor_est <- mle_est(tensor_draws)

tensor_est$mu

tensor_est$sigmas |> lapply(round, 3)
```





# Other models

## Tensor variate generalized hyperbolic

```{r}
genhyper_draws <- rtgenhyper(n = 1e3)
```


## Tensor variate variance gamma
```{r}
vargamma_draws <- rtvargamma(n = 1e3)
```


## Tensor variate normal inverse Gaussian 
```{r}
invgauss_draws <- rtinvgauss(n = 1e3)
```





## Assessing tensor variate normality

We can use the Mahalanobis distance to assess the tensor variate normality. If
we compute the distances and plot them, they should be distributed as
approximately $\chi_{p}$ where $p$ is the product of all the modes of the draws.

```{r}
#| error: true
centered_draws <- matrix_draws - array(rep(matrix_est$mu, each = n), dim = dim(matrix_draws))

inv_sqrt <- function(Sigma, tol = 1e-8) {
  # syemmetric matrix
  Sigma <- (Sigma + t(Sigma)) / 2
  
  eig <- eigen(Sigma, symmetric = TRUE)
  vals <- pmax(eig$values, tol)  # positive eigenvalues
  V <- eig$vectors
  
  Winv <- V %*% diag(1 / sqrt(vals), nrow = length(vals)) %*% t(V)
  Winv
}

D2 <- rep(0, n)

inv_sigmas <- vector("list", length(matrix_est$sigmas))

for(k in 1:2) {
  inv_sigmas[[k]] <- inv_sqrt(matrix_est$sigmas[[k]])
}

for(i in 1:n) {
  Y <- centered_draws[i, , ]
  for(k in 1:2) {
    Y <- n_prod(Y, matrix_est$sigmas[[k]], k)
  }
  D2[i] <- sum(c(Y)^2)
}

sigma2_hat <- mean(D2) / p
D2_std <- D2 / sigma2_hat    

ggplot() +
  geom_histogram(aes(x = D2_std, y = after_stat(density)), color = "black", fill = "pink", bins = 30) + 
  geom_function(fun = dchisq, args = list(df = prod(dim(matrix_est$mu))), color = "red") +
  theme_minimal() + 
  labs(x = "Mahalanobis distance", y = "Density")
```


## Tensor Variate Skewed T


```{r}
tibble(skew25 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 0.25, nu = 6)),
       skew50 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 0.5, nu = 6)),
       skew0 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 0, nu = 6)),
       skew1 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 6)),
       skew5 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 5, nu = 6)),
       skew10 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 10, nu = 6))) |>
  pivot_longer(cols = everything()) |>
  ggplot() +
  geom_histogram(aes(x = value), color = "black", fill = "pink") +
  facet_wrap(~name, nrow = 2, scales = "free_x") +
  theme_minimal()

tibble(nu05 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 0.5)),
       nu1 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 1)),
       nu5 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 5)),
       nu10 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 10)),
       nu20 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 20)),
       nu50 = c(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), skew = 1, nu = 50))) |>
  pivot_longer(cols = everything()) |>
  ggplot() +
  geom_histogram(aes(x = value), color = "black", fill = "pink") +
  facet_wrap(~name, nrow = 2, scales = "free_x") +
  theme_minimal()
```








## Installation

You can install the development version of tensormodels from
[GitHub](https://github.com/) with:
    
```{r}
#| eval: false
if (!requireNamespace("remotes")) install.packages("remotes")
remotes::install_github("ankan12/tensormodels")
```


Reset digits

```{r}
options(digits = old_digits)
```





