---
title: "Recommended Tensor Goodness-of-Fit Test"
subtitle: "Score-adjusted Rosenblatt smooth testing"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)

library(tensortools)
```

## Recommended score-adjusted Rosenblatt GOF test

The following example assumes that `skewt_draws` is a `tensor` object
containing the observed draws. Fit the model once and pass that fit to
`tensor_gof_scores()` so the MLE is not calculated twice.

```{r}
sim_res <- function(i) {
  skewt_draws <- rtskewt(1e3, mu = mu_true, sigmas = list(S1, S2, S3),
                       skew = skew_true, nu = 20)

  skewt_fit <- tensor_mle(
    skewt_draws,
    model = "skewt",
    tol = 1e-6,
    max_iter = 10000
  )

  # 2. Run the score-adjusted Rosenblatt smooth GOF test
  skewt_gof <- tensor_gof_scores(
    skewt_draws,
    model = "skewt",
    fit = skewt_fit,
    marginal_degree = 2,
    include_interactions = FALSE,
    max_basis = 48,
    calibration = "multiplier",
    B = 1999
  )

  tibble::tibble(
    simulation = i,
    stat = unname(skewt_gof$statistic),
    p = skewt_gof$p.value,
    asymptotic_stat = unname(skewt_gof$asymptotic.statistic),
    asymptotic_p = skewt_gof$asymptotic.p.value,
    score_stationarity = skewt_gof$score_stationarity,
    nu = skewt_fit$nu,
    loglik = skewt_fit$loglik
  )
}
```


```{r}
future::plan(
  future::multisession,
  workers = 8
)

progressr::handlers("txtprogressbar")

progressr::with_progress({
  progress <- progressr::progressor(steps = 100)

  all_res <- furrr::future_map_dfr(
    1:100,
    function(iteration) {
      result <- sim_res(iteration)

      progress(
        message = sprintf(
          "Completed simulation %d",
          iteration
        )
      )

      result
    },
    .options = furrr::furrr_options(
      seed = TRUE,
      scheduling = Inf
    )
  )
})

ggplot() +
  geom_histogram(aes(x = all_res$p), color = "black", fill = "pink",
                 bins = 100) +
  geom_vline(aes(xintercept = 0.05), color = "red", lty = 2) +
  theme_minimal()

summary(all_res$p)

sum(all_res$p < 0.05)
```


```{r}
# 1. Fit the proposed distribution

```


```{r, eval=FALSE}
```{r recommended-score-gof-example, eval=FALSE}
# 3. Extract the formal test result
skewt_gof$statistic
skewt_gof$p.value

# The former covariance-whitened chi-squared result is diagnostic only
skewt_gof$asymptotic.statistic
skewt_gof$asymptotic.p.value

# 4. Check whether the fit and numerical ranks support the approximation
skewt_gof$score_stationarity
skewt_gof$information_rank
skewt_gof$adjusted_covariance_rank

# At significance level 0.05:
if (skewt_gof$p.value < 0.05) {
  message("Reject the fitted tensor skew-t model.")
} else {
  message("Do not reject the fitted tensor skew-t model.")
}

skewt_gof$uniforms |>
  c() |>
  as_tibble() |>
  ggplot() +
  geom_histogram(aes(x = value), bins = 100)
```

Interpretation: a small p-value is evidence against the fitted distribution.
A large p-value means that the test did not find evidence against it; it does
not prove that the fitted model is correct. Warnings about score stationarity
or rank deficiency should be resolved before interpreting the p-value.

## Earlier unadjusted exploratory implementation

```{r}
tensor_gof <- function(draws, model) {
  mles <- tensor_mle(draws, model = model, tol = 1e-5)

  #true <- list(mu = mu_true, sigmas = list(S1, S2, S3), skew = skew_true, nu = 20)

  ros_unif <- tensor_rosenblatt(draws, model = model,
                                parameters = mles)

  basis <- make_basis(ros_unif, marginal_degree = 2,
                      include_interactions = TRUE, max_basis = 100)

  basis_means <- colMeans(basis)

  ggplot() +
    geom_histogram(aes(x = c(ros_unif)))

  statistic <- n_draws(draws) * sum(basis_means^2)

  pval <- pchisq(statistic, df = ncol(basis), lower.tail = FALSE)

  tibble(model = model, stat = statistic, p = pval)
}

mu_true <- array(1:24, dim = c(2, 3, 4))

S1 <- toeplitz(0.3^(0:1))
S2 <- toeplitz(0.3^(0:2))
S3 <- toeplitz(0.3^(0:3))

norm_draws <- rtnorm(n = 1e3, mu = mu_true,
                     sigmas = list(S1, S2, S3))



sim_setup <- function(draws) {
  norm_res <- tensor_gof(draws, "normal")
  skewt_res <- tensor_gof(draws, "skewt")
  vargamma_res <- tensor_gof(draws, "vargamma")
  invgauss_res <- tensor_gof(draws, "invgauss")
  genhyper_res <- tensor_gof(draws, "genhyper")

  rbind(norm_res, skewt_res, vargamma_res, invgauss_res, genhyper_res)
}
```


```{r}
tic()
res <- sim_res(1)
toc()

res$p

res$stat
```


```{r}
sim_res <- function(i) {
  tryCatch(
    {
      skewt_draws <- rtskewt(
        n = 1e3,
        mu = array(1:24, dim = c(2, 3, 4)),
        sigmas = list(S1, S2, S3),
        skew = skew_true,
        nu = 20
      )

      tensor_gof(skewt_draws, "skewt") |>
        dplyr::mutate(
          simulation = i,
          error = NA_character_,
          .before = 1
        )
    },
    error = function(error) {
      tibble::tibble(
        simulation = i,
        model = "skewt",
        stat = NA_real_,
        p = NA_real_,
        error = conditionMessage(error)
      )
    }
  )
}
```


```{r}
library(furrr)
library(progressr)


```


```{r}
all_res <- tibble(model= rep("skewt", 500), stat = rep(NA, 500), p = rep(NA, 500))

for(i in 1:20) {
  if(i %% 5 == 0) print(i)

  res <- sim_res(i)

  all_res[i, ] <- res
}

sum(all_res$p[1:50] < 0.05)

ggplot() +
  geom_histogram(aes(x = all_res$p[1:50]), bins = 100,
                 color = "black", fill = "pink") +
  geom_vline(aes(xintercept = 0.05), color = "red") +
  theme_minimal()


plan(multisession, workers = 6)

all_res <- future_map_dfr(1:50, sim_res, .progress = TRUE,
                          .options = furrr::furrr_options(seed = TRUE))
```



```{r}
tensor_gof(skewt_draws, "normal")
tensor_gof(skewt_draws, "invgauss")
tensor_gof(skewt_draws, "genhyper")
tensor_gof(skewt_draws, "skewt")
tensor_gof(skewt_draws, "vargamma")

```


```{r}
matricization(ros_norm, 1)


ks_result <- ks.test(
  c(ros_norm),
  "punif",
  min = 0,
  max = 1
)

ros_test <- ad_uniform_test(ros_norm, B = 1e3)

ros_test$p.value
```



```{r}
skew_true <- array(seq(by = 0.1, length = 24), dim = c(2, 3, 4))

skewt_draws <- rtskewt(n = 1e3, mu = array(1:24, dim = c(2, 3, 4)),
                       sigmas = list(S1, S2, S3), skew = skew_true, nu = 20)

mles_skewt <- tensor_mle(skewt_draws, model = "normal",
                         tol = 1e-5, quiet = FALSE, max_iter = 1e4)

res <- tensor_rosenblatt(skewt_draws, model = "normal",
                         parameters = mles_skewt)

ros_test <- ad_uniform_test(res, B = 1e3)

ros_test$statistic
ros_test$p.value

res_true <- tensor_rosenblatt(
  skewt_draws,
  model = "skewt",
  parameters = list(
    mu = array(1:24, dim = c(2, 3, 4)),
    sigmas = list(S1, S2, S3),
    skew = skew_true,
    nu = 20
  )
)

res_2 <- ad_uniform_test(res_true, B = 1000, seed = 1)

res_2$p.value
```

# Finite-basis score-adjusted Rosenblatt test

`tensor_score_gof()` fits the requested model, performs one Rosenblatt
transformation, and adjusts a joint smooth-uniformity statistic for the
estimated parameter scores. First-degree and higher-degree marginal terms
test uniformity, while products between coordinates test residual dependence.

The chi-squared p-value is asymptotic. Check `information_rank` and
`score_stationarity` before interpreting it. The skewed models can take some
time because they require quadrature and finite-difference likelihood scores,
but they do not require a parametric bootstrap. Treat the implementation as
experimental until its size has been verified by simulation for the selected
model, tensor dimensions, and sample size; finite-sample calibration can be
poor for weakly identified skewed-model parameters.

```{r finite-basis-score-gof, eval=FALSE}
skewt_score_gof <-
  tensor_score_gof(draws = norm_draws,
  model = "skewt",
  marginal_degree = 2,
  include_interactions = TRUE,
  max_basis = 100,
  tol = 1e-6,
  max_iter = 10000
)

skewt_score_gof
skewt_score_gof$p.value
skewt_score_gof$information_rank
skewt_score_gof$score_dimension
skewt_score_gof$score_stationarity
```

`make_basis()` exposes the basis matrix directly. Rows correspond to tensor
draws and columns correspond to marginal or interaction terms.

```{r make-uniform-basis, eval=FALSE}
uniform_basis <- make_basis(
  skewt_score_gof$uniforms,
  marginal_degree = 2,
  include_interactions = TRUE,
  max_basis = 100
)

dim(uniform_basis)
colnames(uniform_basis)
colMeans(uniform_basis)

# Unadjusted known-parameter smooth statistic. Its chi-squared p-value is not
# formally calibrated when the Rosenblatt parameters were fitted to these data.
unadjusted_statistic <-
  nrow(uniform_basis) * sum(colMeans(uniform_basis)^2)
unadjusted_p_value <- stats::pchisq(
  unadjusted_statistic,
  df = ncol(uniform_basis),
  lower.tail = FALSE
)
```

# Simulation study for the score-adjusted test

The function below estimates both size and power. The row for the correctly
fitted model estimates the Type I error rate: at `alpha = 0.05`, its rejection
rate should be close to 0.05. Rows for incorrectly fitted models estimate
power, so larger rejection rates are desirable there.

Every fitted model is tested on the same newly generated sample within a
replication. In addition to the rejection rate, the study records numerical
rank problems, score-stationarity problems, errors, and computation time.

```{r finite-basis-score-simulation-function}
score_gof_simulation <- function(
    n_sim = 100L,
    n = 1000L,
    generating_model = "skewt",
    generating_parameters,
    fitted_models = c(
      "normal", "skewt", "vargamma", "invgauss", "genhyper"
    ),
    alpha = 0.05,
    marginal_degree = 2L,
    include_interactions = TRUE,
    max_basis = 100L,
    fd_step = 1e-5,
    eigen_tol = 1e-8,
    mle_tol = 1e-6,
    mle_max_iter = 10000L,
    rel.tol = 1e-7,
    abs.tol = 1e-9,
    subdivisions = 100L,
    seed = NULL,
    show_progress = TRUE) {

  available_models <- c(
    "normal", "skewt", "vargamma", "invgauss", "genhyper"
  )
  generating_model <- match.arg(generating_model, available_models)
  fitted_models <- match.arg(
    fitted_models,
    available_models,
    several.ok = TRUE
  )

  n_sim <- as.integer(n_sim)
  n <- as.integer(n)

  if (length(n_sim) != 1L || is.na(n_sim) || n_sim < 1L) {
    stop("`n_sim` must be one positive integer.")
  }
  if (length(n) != 1L || is.na(n) || n < 3L) {
    stop("`n` must be one integer greater than two.")
  }
  if (!is.list(generating_parameters)) {
    stop("`generating_parameters` must be a parameter list.")
  }
  if (length(alpha) != 1L || !is.finite(alpha) ||
      alpha <= 0 || alpha >= 1) {
    stop("`alpha` must be strictly between zero and one.")
  }
  if (!is.null(seed)) {
    set.seed(seed)
  }

  required_parameter <- function(name, alternative = NULL) {
    value <- generating_parameters[[name]]

    if (is.null(value) && !is.null(alternative)) {
      value <- generating_parameters[[alternative]]
    }
    if (is.null(value)) {
      alternatives <- if (is.null(alternative)) {
        ""
      } else {
        paste0(" (or `", alternative, "`)")
      }
      stop(
        "The generating parameter `", name, "`", alternatives,
        " is missing."
      )
    }

    value
  }

  simulate_sample <- function() {
    mu <- required_parameter("mu")
    sigmas <- required_parameter("sigmas")

    switch(
      generating_model,
      normal = rtnorm(
        n = n,
        mu = mu,
        sigmas = sigmas
      ),
      skewt = rtskewt(
        n = n,
        mu = mu,
        sigmas = sigmas,
        skew = required_parameter("skew"),
        nu = required_parameter("nu")
      ),
      vargamma = rtvargamma(
        n = n,
        mu = mu,
        sigmas = sigmas,
        skew = required_parameter("skew"),
        scale = required_parameter("gamma", "scale")
      ),
      invgauss = rtinvgauss(
        n = n,
        mu = mu,
        sigmas = sigmas,
        skew = required_parameter("skew"),
        kappa = required_parameter("kappa")
      ),
      genhyper = rtgenhyper(
        n = n,
        mu = mu,
        sigmas = sigmas,
        skew = required_parameter("skew"),
        lambda = required_parameter("lambda"),
        omega = required_parameter("omega")
      )
    )
  }

  result_rows <- vector("list", n_sim * length(fitted_models))
  result_index <- 0L

  for (simulation in seq_len(n_sim)) {
    draws <- simulate_sample()

    for (fitted_model in fitted_models) {
      result_index <- result_index + 1L
      start_time <- proc.time()[["elapsed"]]
      warning_messages <- character()

      result_rows[[result_index]] <- tryCatch(
        {
          test <- withCallingHandlers(
            tensor_score_gof(
              draws = draws,
              model = fitted_model,
              marginal_degree = marginal_degree,
              include_interactions = include_interactions,
              max_basis = max_basis,
              fd_step = fd_step,
              eigen_tol = eigen_tol,
              max_iter = mle_max_iter,
              tol = mle_tol,
              quiet = TRUE,
              rel.tol = rel.tol,
              abs.tol = abs.tol,
              subdivisions = subdivisions
            ),
            warning = function(warning) {
              warning_messages <<- c(
                warning_messages,
                conditionMessage(warning)
              )
              invokeRestart("muffleWarning")
            }
          )

          data.frame(
            simulation = simulation,
            generating_model = generating_model,
            fitted_model = fitted_model,
            statistic = unname(test$statistic),
            degrees_freedom = unname(test$parameter),
            p_value = test$p.value,
            reject = test$p.value <= alpha,
            score_dimension = test$score_dimension,
            information_rank = test$information_rank,
            rank_deficient =
              test$information_rank < test$score_dimension,
            adjusted_covariance_rank =
              test$adjusted_covariance_rank,
            score_stationarity = test$score_stationarity,
            elapsed_seconds =
              proc.time()[["elapsed"]] - start_time,
            warning = if (length(warning_messages)) {
              paste(unique(warning_messages), collapse = " | ")
            } else {
              NA_character_
            },
            error = NA_character_,
            stringsAsFactors = FALSE
          )
        },
        error = function(error) {
          data.frame(
            simulation = simulation,
            generating_model = generating_model,
            fitted_model = fitted_model,
            statistic = NA_real_,
            degrees_freedom = NA_integer_,
            p_value = NA_real_,
            reject = NA,
            score_dimension = NA_integer_,
            information_rank = NA_integer_,
            rank_deficient = NA,
            adjusted_covariance_rank = NA_integer_,
            score_stationarity = NA_real_,
            elapsed_seconds =
              proc.time()[["elapsed"]] - start_time,
            warning = if (length(warning_messages)) {
              paste(unique(warning_messages), collapse = " | ")
            } else {
              NA_character_
            },
            error = conditionMessage(error),
            stringsAsFactors = FALSE
          )
        }
      )
    }

    if (show_progress) {
      message("Completed simulation ", simulation, " of ", n_sim)
    }
  }

  results <- do.call(rbind, result_rows)
  rownames(results) <- NULL

  summary_rows <- lapply(fitted_models, function(fitted_model) {
    current <- results[
      results$fitted_model == fitted_model,
      ,
      drop = FALSE
    ]
    completed <- is.na(current$error)
    number_completed <- sum(completed)

    if (number_completed > 0L) {
      rejection_rate <- mean(current$reject[completed])
      rejection_mcse <- sqrt(
        rejection_rate * (1 - rejection_rate) / number_completed
      )
    } else {
      rejection_rate <- NA_real_
      rejection_mcse <- NA_real_
    }

    data.frame(
      generating_model = generating_model,
      fitted_model = fitted_model,
      completed = number_completed,
      failures = sum(!completed),
      rejection_rate = rejection_rate,
      rejection_mcse = rejection_mcse,
      rank_deficiency_rate = if (number_completed > 0L) {
        mean(current$rank_deficient[completed])
      } else {
        NA_real_
      },
      stationarity_warning_rate = if (number_completed > 0L) {
        mean(current$score_stationarity[completed] > 1)
      } else {
        NA_real_
      },
      median_p_value = if (number_completed > 0L) {
        stats::median(current$p_value[completed])
      } else {
        NA_real_
      },
      mean_elapsed_seconds = if (number_completed > 0L) {
        mean(current$elapsed_seconds[completed])
      } else {
        NA_real_
      },
      stringsAsFactors = FALSE
    )
  })

  list(
    summary = do.call(rbind, summary_rows),
    results = results,
    settings = list(
      n_sim = n_sim,
      n = n,
      alpha = alpha,
      generating_model = generating_model,
      generating_parameters = generating_parameters,
      fitted_models = fitted_models,
      marginal_degree = marginal_degree,
      include_interactions = include_interactions,
      max_basis = max_basis,
      seed = seed
    )
  )
}
```

This runs the study requested above: generate from the tensor skew-\(t\), then
fit and test all five candidate distributions.

```{r run-finite-basis-score-simulation, eval=FALSE}
skewt_score_study <- score_gof_simulation(
  n_sim = 10,
  n = 1000,
  generating_model = "skewt",
  generating_parameters = list(
    mu = array(1:24, dim = c(2, 3, 4)),
    sigmas = list(S1, S2, S3),
    skew = skew_true,
    nu = 20
  ),
  fitted_models = c(
    "normal", "skewt"), #, "vargamma", "invgauss", "genhyper"

  alpha = 0.05,
  marginal_degree = 2,
  include_interactions = TRUE,
  max_basis = 100,
  seed = 20260730
)

skewt_score_study$summary
head(skewt_score_study$results)
subset(
  skewt_score_study$results,
  !is.na(error) | rank_deficient | score_stationarity > 1
)

# Under the null, these p-values should be approximately uniform.
skewt_null_results <- subset(
  skewt_score_study$results,
  fitted_model == "skewt" & is.na(error)
)

hist(
  skewt_null_results$p_value,
  breaks = seq(0, 1, by = 0.05),
  main = "Skew-t null p-values",
  xlab = "p-value"
)

stats::binom.test(
  x = sum(skewt_null_results$reject),
  n = nrow(skewt_null_results),
  p = 0.05
)
```

For the correctly specified `"skewt"` row, compare `rejection_rate` with 0.05
and use `rejection_mcse` to judge Monte Carlo uncertainty. With 100
replications, a true 5% rejection rate has Monte Carlo standard error about
0.022, so use at least 500 replications for a serious calibration study. The
other four rejection rates measure power against those fitted distributions,
not Type I error. Begin with a small value such as `n_sim = 10` to estimate
runtime before launching the full study.


#Inverse Transform Norm Draws
```{r}
standard_draws <- norm_draws - mles_norm$mu

for (i in seq_along(mles_norm$sigmas)) {
  curr_sig <- mles_norm$sigmas[[i]]
  lower_chol <- t(chol(curr_sig))
  whitening_matrix <- solve(lower_chol)

  standard_draws <- n_prod(standard_draws, i, whitening_matrix)
}

uniform_draws <- pnorm(standard_draws)

uniform_draws |>
  c() |>
  as_tibble() |>
  ggplot() +
  geom_histogram(aes(x = value), color = "black", fill = "pink",
                   binwidth = 0.01, boundary = 0) +
  theme_minimal()

var(uniform_draws)
mean(uniform_draws)
```


#Inverse Transform Skew-t Draws
```{r}


res |>
  c() |>
  as_tibble() |>
  ggplot() +
  geom_histogram(aes(x = value), bins = 100) +
  theme_minimal()
```


```{r}
ad_uniform_statistic <- function(values) {
  values <- sort(as.numeric(values))
  N <- length(values)

  if (any(!is.finite(values)) ||
      any(values < 0 | values > 1)) {
    stop("Values must be finite and between zero and one.")
  }

  if (any(values == 0 | values == 1)) {
    return(Inf)
  }

  i <- seq_len(N)

  -N - sum(
    (2 * i - 1) *
      (
        log(values) +
        log1p(-rev(values))
      )
  ) / N
}

ad_uniform_test <- function(values, B = 4999, seed = NULL) {
  if (!is.null(seed)) {
    set.seed(seed)
  }

  values <- as.numeric(values)
  observed <- ad_uniform_statistic(values)
  N <- length(values)

  null_statistics <- replicate(
    B,
    ad_uniform_statistic(runif(N))
  )

  p_value <- (
    1 + sum(null_statistics >= observed)
  ) / (B + 1)

  list(
    statistic = observed,
    p.value = p_value,
    B = B,
    null_statistics = null_statistics
  )
}


r <- ad_uniform_test(runif(1e3), B = 1000)

ad_uniform_statistic(unif_true)

result <- ad_uniform_test(
  unif_true,
  B = 1000,
  seed = 1
)

result$statistic
result$p.value
```

# Simulation study: data generated from the tensor skew-t

This function repeats the following steps:

1. Generate a new tensor skew-\(t\) sample.
2. Fit each candidate tensor distribution to the same sample.
3. Apply that fitted model's Rosenblatt transform.
4. calculate the Anderson-Darling statistic and its simple uniform-reference
   p-value.
5. Record whether the model is rejected at the chosen significance level.

The simulated Anderson-Darling null distribution is constructed only once and
reused for every fitted model because every transformed sample has the same
number of entries. Exact zeros and ones are recorded separately because they
make the Anderson-Darling statistic infinite.

```{r}
skewt_rosenblatt_simulation <- function(
    n_sim = 100L,
    n = 1000L,
    mu,
    sigmas,
    skew,
    nu = 20,
    models = c("normal", "skewt", "vargamma", "invgauss", "genhyper"),
    alpha = 0.05,
    B = 1000L,
    mle_tol = 1e-5,
    mle_max_iter = 10000L,
    rel.tol = 1e-7,
    abs.tol = 1e-9,
    subdivisions = 100L,
    seed = NULL,
    show_progress = TRUE) {

  available_models <- c(
    "normal", "skewt", "vargamma", "invgauss", "genhyper"
  )
  models <- match.arg(models, available_models, several.ok = TRUE)

  n_sim <- as.integer(n_sim)
  n <- as.integer(n)
  B <- as.integer(B)

  if (n_sim < 1L || n < 1L || B < 1L) {
    stop("`n_sim`, `n`, and `B` must be positive integers.")
  }
  if (alpha <= 0 || alpha >= 1) {
    stop("`alpha` must be strictly between zero and one.")
  }
  if (!is.null(seed)) {
    set.seed(seed)
  }

  # Every transformed dataset contains n * prod(dim(mu)) uniform values.
  n_uniforms <- n * prod(dim(mu))

  # Reuse this reference distribution for every replication and fitted model.
  ad_null_statistics <- replicate(
    B,
    ad_uniform_statistic(runif(n_uniforms))
  )

  result_rows <- vector("list", n_sim * length(models))
  result_index <- 0L

  for (simulation in seq_len(n_sim)) {
    draws <- rtskewt(
      n = n,
      mu = mu,
      sigmas = sigmas,
      skew = skew,
      nu = nu
    )

    for (fitted_model in models) {
      result_index <- result_index + 1L
      start_time <- proc.time()[["elapsed"]]

      result_rows[[result_index]] <- tryCatch(
        {
          fit <- tensor_mle(
            draws,
            model = fitted_model,
            tol = mle_tol,
            max_iter = mle_max_iter,
            quiet = TRUE
          )

          uniforms <- tensor_rosenblatt(
            draws,
            model = fitted_model,
            parameters = fit,
            rel.tol = rel.tol,
            abs.tol = abs.tol,
            subdivisions = subdivisions
          )

          uniform_values <- as.numeric(uniforms)
          statistic <- ad_uniform_statistic(uniform_values)
          p_value <- (
            1 + sum(ad_null_statistics >= statistic)
          ) / (B + 1)

          data.frame(
            simulation = simulation,
            generating_model = "skewt",
            fitted_model = fitted_model,
            statistic = statistic,
            p_value = p_value,
            reject = p_value <= alpha,
            loglik = fit$loglik,
            exact_zero = sum(uniform_values == 0),
            exact_one = sum(uniform_values == 1),
            elapsed_seconds = proc.time()[["elapsed"]] - start_time,
            error = NA_character_,
            stringsAsFactors = FALSE
          )
        },
        error = function(error) {
          data.frame(
            simulation = simulation,
            generating_model = "skewt",
            fitted_model = fitted_model,
            statistic = NA_real_,
            p_value = NA_real_,
            reject = NA,
            loglik = NA_real_,
            exact_zero = NA_integer_,
            exact_one = NA_integer_,
            elapsed_seconds = proc.time()[["elapsed"]] - start_time,
            error = conditionMessage(error),
            stringsAsFactors = FALSE
          )
        }
      )

      if (show_progress) {
        message(
          "Simulation ", simulation, " of ", n_sim,
          "; fitted model: ", fitted_model
        )
      }
    }
  }

  results <- do.call(rbind, result_rows)

  summary_rows <- lapply(models, function(fitted_model) {
    current <- results[results$fitted_model == fitted_model, , drop = FALSE]
    completed <- is.na(current$error)

    data.frame(
      generating_model = "skewt",
      fitted_model = fitted_model,
      completed = sum(completed),
      failures = sum(!completed),
      rejection_rate = if (any(completed)) {
        mean(current$reject[completed])
      } else {
        NA_real_
      },
      endpoint_rate = if (any(completed)) {
        mean(
          current$exact_zero[completed] > 0 |
            current$exact_one[completed] > 0
        )
      } else {
        NA_real_
      },
      median_statistic = if (any(completed)) {
        median(current$statistic[completed])
      } else {
        NA_real_
      },
      median_p_value = if (any(completed)) {
        median(current$p_value[completed])
      } else {
        NA_real_
      },
      mean_elapsed_seconds = if (any(completed)) {
        mean(current$elapsed_seconds[completed])
      } else {
        NA_real_
      },
      stringsAsFactors = FALSE
    )
  })

  list(
    summary = do.call(rbind, summary_rows),
    results = results,
    settings = list(
      n_sim = n_sim,
      n = n,
      alpha = alpha,
      B = B,
      generating_model = "skewt",
      models = models,
      seed = seed
    ),
    ad_null_statistics = ad_null_statistics
  )
}
```

The complete 100-replication study is intentionally not evaluated while
knitting. It requires four quadrature-based Rosenblatt transformations for
each replication and can therefore take a long time.

```{r skewt-simulation-study, eval=FALSE}
skewt_study <- skewt_rosenblatt_simulation(
  n_sim = 100,
  n = 1000,
  mu = array(1:24, dim = c(2, 3, 4)),
  sigmas = list(S1, S2, S3),
  skew = skew_true,
  nu = 20,
  alpha = 0.05,
  B = 1000,
  mle_tol = 1e-5,
  mle_max_iter = 10000,
  seed = 20260730
)

skewt_study$summary
skewt_study$results
```

For a quick check that the workflow runs before starting all 100 replications:

```{r skewt-simulation-smoke-test, eval=FALSE}
skewt_smoke_test <- skewt_rosenblatt_simulation(
  n_sim = 2,
  n = 1000,
  mu = array(1:24, dim = c(2, 3, 4)),
  sigmas = list(S1, S2, S3),
  skew = array(2 * rnorm(24), dim = c(2, 3, 4)),
  nu = 20,
  B = 199,
  seed = 1
)

skewt_smoke_test$results

skewt_smoke_test$summary |> View()
```


```{r}
unif <- rosenblatt_tskewt(skewt_draws, mles_skewt$mu, mles_skewt$skew,
                          mles_skewt$sigmas, mles_skewt$nu)


unif_true <- rosenblatt_tskewt(
  skewt_draws,
  mu = array(1:24, dim = c(2, 3, 4)),
  skew = skew_true,
  sigmas = list(S1, S2, S3),
  nu = 20
)

unif_true |>
  c() |>
  as_tibble() |>
  ggplot() +
  geom_histogram(aes(x = value, y = after_stat(density)),
                 color = "black", fill = "pink",
                 binwidth = 0.01, boundary = 0) +
  geom_density(aes(x = value), color = "red") +
  theme_minimal()

mles_norm_skew <- tensor_mle(norm_draws, model = "skewt")

unif_false <- rosenblatt_tskewt(
  norm_draws,
  mu = mles_norm_skew$mu,
  skew = mles_norm_skew$skew,
  sigmas = mles_norm_skew$sigmas,
  nu = mles_norm_skew$nu
)

unif_false |>
  c() |>
  as_tibble() |>
  ggplot() +
  geom_histogram(aes(x = value, y = after_stat(density)),
                 color = "black", fill = "pink",
                 binwidth = 0.01, boundary = 0) +
  geom_density(aes(x = value), color = "red") +
  theme_minimal()
```

# Recommended procedure

This file contains only the recommended characteristic-function goodness-of-fit
test. It supports all five distributions fitted by `tensor_mle()`:

- tensor normal;
- tensor skew-\(t\);
- tensor variance-gamma;
- tensor inverse-Gaussian; and
- tensor generalized hyperbolic.

For each fitted model, the test compares the empirical characteristic function
with the fitted theoretical characteristic function at a fixed collection of
multiscale frequencies:

\[
T_n
=
n\frac{1}{L}\sum_{\ell=1}^L
\left|
\widehat\varphi_n(t_\ell)
-
\varphi(t_\ell;\widehat\theta)
\right|^2.
\]

The frequencies include both dense directions and rank-one tensor directions
at several radii. The theoretical characteristic functions are evaluated
analytically, so no reference sample is needed to approximate them.

The null distribution is obtained with a parametric bootstrap. Every bootstrap
sample is refitted before its statistic is calculated. This is the part that
accounts for estimating the model parameters from the same observations used
in the goodness-of-fit test.

# 1. Construct multiscale frequencies

Rank-one directions are included because they can detect discrepancies aligned
with the tensor modes. Dense directions can detect discrepancies that are not
well aligned with a single separable direction.

```{r frequency-directions}
.multiscale_cf_rank_one_direction <- function(dims) {
  mode_vectors <- lapply(dims, function(mode_size) {
    direction <- rnorm(mode_size)
    direction / sqrt(sum(direction^2))
  })

  as.numeric(Reduce(
    function(left, right) as.vector(outer(left, right)),
    mode_vectors
  ))
}
```

```{r frequency-grid}
make_multiscale_cf_frequencies <- function(dims,
                                           radii = c(0.5, 1, 2, 3),
                                           n_per_type = 8L,
                                           seed = NULL) {
  if (!is.null(seed)) {
    set.seed(seed)
  }

  if (any(!is.finite(radii)) || any(radii <= 0)) {
    stop("`radii` must contain positive finite values.", call. = FALSE)
  }
  if (length(n_per_type) != 1L || !is.finite(n_per_type) ||
      n_per_type < 1L || n_per_type != floor(n_per_type)) {
    stop("`n_per_type` must be one positive integer.", call. = FALSE)
  }

  p <- prod(dims)
  n_per_type <- as.integer(n_per_type)
  n_frequencies <- length(radii) * 2L * n_per_type
  frequencies <- matrix(0, nrow = n_frequencies, ncol = p)
  direction_type <- character(n_frequencies)
  frequency_radius <- numeric(n_frequencies)
  row <- 0L

  for (radius in radii) {
    for (j in seq_len(n_per_type)) {
      row <- row + 1L
      dense <- rnorm(p)
      frequencies[row, ] <- radius * dense / sqrt(sum(dense^2))
      direction_type[row] <- "dense"
      frequency_radius[row] <- radius

      row <- row + 1L
      frequencies[row, ] <- radius *
        .multiscale_cf_rank_one_direction(dims)
      direction_type[row] <- "rank-one"
      frequency_radius[row] <- radius
    }
  }

  attr(frequencies, "direction_type") <- direction_type
  attr(frequencies, "radius") <- frequency_radius
  frequencies
}
```

# 2. Center and whiten the tensor observations

Whitening is applied mode by mode. This avoids constructing the full
Kronecker covariance matrix.

```{r whitening-matrices}
.multiscale_cf_whitening <- function(sigmas) {
  lapply(sigmas, function(sigma) {
    sigma <- as.matrix(sigma)
    sigma <- (sigma + t(sigma)) / 2
    solve(t(chol(sigma)))
  })
}
```

```{r standardize-draws}
.multiscale_cf_standardize <- function(draws, fit) {
  n <- n_draws(draws)
  dims <- draw_shape(draws)
  whitening <- .multiscale_cf_whitening(fit$sigmas)
  result <- matrix(0, nrow = n, ncol = prod(dims))

  for (i in seq_len(n)) {
    current <- pull_draw(draws, i) - fit$mu
    for (mode in seq_along(whitening)) {
      current <- n_prod(current, mode, whitening[[mode]])
    }
    result[i, ] <- as.numeric(unclass(current))
  }

  result
}
```

```{r whiten-skew}
.multiscale_cf_whiten_skew <- function(skew, sigmas) {
  whitening <- .multiscale_cf_whitening(sigmas)
  result <- skew

  for (mode in seq_along(whitening)) {
    result <- n_prod(result, mode, whitening[[mode]])
  }

  as.numeric(result)
}
```

# 3. Evaluate the fitted theoretical characteristic function

After whitening, the tensor normal characteristic function is
\(\exp(-\lVert t\rVert^2/2)\).

Each skewed model has the representation

\[
Z = W a + \sqrt{W}\,\varepsilon,
\qquad \varepsilon\sim N(0,I),
\]

where \(a\) is the whitened skew tensor. Its characteristic function is the
Laplace transform of \(W\), evaluated at

\[
s=\frac{\lVert t\rVert^2}{2}-i\,t^\top a.
\]

```{r mixing-laplace-transforms}
.multiscale_cf_mixing_laplace <- function(s, model, fit) {
  if (model == "skewt") {
    shape <- fit$nu / 2
    rate <- fit$nu / 2
    answer <- complex(length(s))
    at_zero <- Mod(s) < sqrt(.Machine$double.eps)
    answer[at_zero] <- 1 + 0i

    if (any(!at_zero)) {
      current_s <- s[!at_zero]
      argument <- 2 * sqrt(rate * current_s)

      # BesselK(..., expon.scaled = TRUE) equals exp(argument) K(argument).
      log_bessel_k <- log(Bessel::BesselK(
        argument,
        nu = shape,
        expon.scaled = TRUE
      )) - argument

      answer[!at_zero] <- exp(
        log(2) +
          (shape / 2) * (log(rate) + log(current_s)) -
          lgamma(shape) +
          log_bessel_k
      )
    }
    return(answer)
  }

  if (model == "vargamma") {
    return((fit$gamma / (fit$gamma + s))^fit$gamma)
  }

  if (model == "invgauss") {
    return(exp(
      fit$kappa * (1 - sqrt(1 + 2 * s / fit$kappa^2))
    ))
  }

  if (model == "genhyper") {
    shifted_psi <- fit$omega + 2 * s
    argument <- sqrt(fit$omega * shifted_psi)

    log_numerator_bessel <- log(Bessel::BesselK(
      argument,
      nu = fit$lambda,
      expon.scaled = TRUE
    )) - argument
    log_denominator_bessel <- log(Bessel::BesselK(
      fit$omega,
      nu = fit$lambda,
      expon.scaled = TRUE
    )) - fit$omega

    return(exp(
      (fit$lambda / 2) * (log(fit$omega) - log(shifted_psi)) +
        log_numerator_bessel - log_denominator_bessel
    ))
  }

  stop(
    "No mixing distribution is required for the normal model.",
    call. = FALSE
  )
}
```

```{r model-characteristic-function}
.multiscale_cf_model <- function(frequencies, model, fit) {
  q <- rowSums(frequencies^2)

  if (model == "normal") {
    return(exp(-q / 2))
  }

  whitened_skew <- .multiscale_cf_whiten_skew(
    fit$skew,
    fit$sigmas
  )
  skew_projection <- drop(frequencies %*% whitened_skew)
  s <- q / 2 - 1i * skew_projection

  .multiscale_cf_mixing_laplace(s, model, fit)
}
```

# 4. Calculate the observed discrepancy

```{r cf-test-statistic}
.multiscale_cf_statistic <- function(draws, model, fit, frequencies) {
  standardized <- .multiscale_cf_standardize(draws, fit)
  phases <- standardized %*% t(frequencies)
  empirical_cf <- colMeans(exp(1i * phases))
  theoretical_cf <- .multiscale_cf_model(frequencies, model, fit)

  if (any(!is.finite(theoretical_cf))) {
    stop(
      "The fitted model produced a non-finite characteristic function.",
      call. = FALSE
    )
  }

  squared_difference <- Mod(empirical_cf - theoretical_cf)^2

  list(
    statistic = n_draws(draws) * mean(squared_difference),
    empirical_cf = empirical_cf,
    theoretical_cf = theoretical_cf,
    squared_difference = squared_difference
  )
}
```

# 5. Simulate from each fitted model

```{r simulate-fitted-model}
.multiscale_cf_simulate <- function(n, model, fit) {
  switch(
    model,
    normal = rtnorm(
      n = n,
      mu = fit$mu,
      sigmas = fit$sigmas
    ),
    skewt = rtskewt(
      n = n,
      mu = fit$mu,
      sigmas = fit$sigmas,
      skew = fit$skew,
      nu = fit$nu
    ),
    vargamma = rtvargamma(
      n = n,
      mu = fit$mu,
      sigmas = fit$sigmas,
      skew = fit$skew,
      scale = fit$gamma
    ),
    invgauss = rtinvgauss(
      n = n,
      mu = fit$mu,
      sigmas = fit$sigmas,
      skew = fit$skew,
      kappa = fit$kappa
    ),
    genhyper = rtgenhyper(
      n = n,
      mu = fit$mu,
      sigmas = fit$sigmas,
      skew = fit$skew,
      lambda = fit$lambda,
      omega = fit$omega
    )
  )
}
```

# 6. Run the complete test

This is the only function users need to call. Bootstrap failures are retained
in the result rather than silently discarded.

```{r multiscale-cf-bootstrap-test}
tensor_cf_multiscale_bootstrap <- function(
    draws,
    model = c("normal", "skewt", "vargamma", "invgauss", "genhyper"),
    B = 999L,
    frequencies = NULL,
    radii = c(0.5, 1, 2, 3),
    n_per_type = 8L,
    max_iter = 1000L,
    tol = 1e-4,
    quiet = TRUE,
    restrict = NULL,
    seed = NULL,
    show_progress = FALSE) {
  model <- match.arg(model)
  data_name <- deparse(substitute(draws))

  if (!inherits(draws, "tensor")) {
    stop("`draws` must be a tensor object.", call. = FALSE)
  }
  if (n_draws(draws) < 2L) {
    stop("At least two tensor draws are required.", call. = FALSE)
  }
  if (length(B) != 1L || !is.finite(B) ||
      B < 1L || B != floor(B)) {
    stop("`B` must be one positive integer.", call. = FALSE)
  }

  B <- as.integer(B)
  if (!is.null(seed)) {
    set.seed(seed)
  }

  p <- prod(draw_shape(draws))
  if (is.null(frequencies)) {
    frequencies <- make_multiscale_cf_frequencies(
      dims = draw_shape(draws),
      radii = radii,
      n_per_type = n_per_type
    )
  } else {
    frequency_attributes <- attributes(frequencies)[
      c("direction_type", "radius")
    ]
    frequencies <- as.matrix(frequencies)
    if (ncol(frequencies) != p) {
      stop("`frequencies` must have ", p, " columns.", call. = FALSE)
    }
    attributes(frequencies)[names(frequency_attributes)] <-
      frequency_attributes
  }

  observed_fit <- tensor_mle(
    draws,
    model = model,
    max_iter = max_iter,
    tol = tol,
    quiet = quiet,
    restrict = restrict
  )
  observed <- .multiscale_cf_statistic(
    draws,
    model,
    observed_fit,
    frequencies
  )

  bootstrap_statistics <- rep(NA_real_, B)
  failure_messages <- character(B)

  for (b in seq_len(B)) {
    if (show_progress) {
      message("Bootstrap replicate ", b, " of ", B, " for ", model)
    }

    bootstrap_result <- tryCatch({
      bootstrap_draws <- .multiscale_cf_simulate(
        n_draws(draws),
        model,
        observed_fit
      )
      bootstrap_fit <- tensor_mle(
        bootstrap_draws,
        model = model,
        max_iter = max_iter,
        tol = tol,
        quiet = quiet,
        restrict = restrict
      )
      .multiscale_cf_statistic(
        bootstrap_draws,
        model,
        bootstrap_fit,
        frequencies
      )
    }, error = function(error) error)

    if (inherits(bootstrap_result, "error")) {
      failure_messages[b] <- conditionMessage(bootstrap_result)
    } else if (is.finite(bootstrap_result$statistic)) {
      bootstrap_statistics[b] <- bootstrap_result$statistic
    } else {
      failure_messages[b] <- "The bootstrap statistic was non-finite."
    }
  }

  successful <- is.finite(bootstrap_statistics)
  if (!any(successful)) {
    stop("All bootstrap replicates failed.", call. = FALSE)
  }

  p_value <- (
    1 + sum(bootstrap_statistics[successful] >= observed$statistic)
  ) / (1 + sum(successful))

  model_label <- c(
    normal = "tensor normal",
    skewt = "tensor skew-t",
    vargamma = "tensor variance-gamma",
    invgauss = "tensor inverse-Gaussian",
    genhyper = "tensor generalized hyperbolic"
  )[[model]]

  structure(
    list(
      statistic = c(T = observed$statistic),
      parameter = c(B = sum(successful)),
      p.value = p_value,
      method = paste0(
        "Parametric-bootstrap multiscale CF test for the ",
        model_label
      ),
      data.name = data_name,
      model = model,
      fit = observed_fit,
      frequencies = frequencies,
      empirical_cf = observed$empirical_cf,
      theoretical_cf = observed$theoretical_cf,
      squared_difference = observed$squared_difference,
      bootstrap_statistics = bootstrap_statistics,
      B_requested = B,
      B_successful = sum(successful),
      B_failed = sum(!successful),
      failure_messages = failure_messages[!successful]
    ),
    class = c("tensor_cf_bootstrap_test", "htest")
  )
}
```

# 7. Use the test

Use a small value of `B` only while checking an analysis. A reported test
should generally use at least `B = 999`, giving a minimum attainable add-one
p-value of \(1/1000\).

```{r quick-example, eval=FALSE}

quick_result <- tensor_cf_multiscale_bootstrap(
  draws = norm_draws,
  model = "skewt",
  B = 49,
  seed = 20260722,
  show_progress = TRUE
)

quick_result
```

```{r reported-example, eval=FALSE}
final_result <- tensor_cf_multiscale_bootstrap(
  draws = observed_tensor_draws,
  model = "skewt",
  B = 999,
  radii = c(0.5, 1, 2, 3),
  n_per_type = 8,
  max_iter = 1000,
  tol = 1e-4,
  seed = 20260722,
  show_progress = TRUE
)

final_result
```

# 8. Compare all five fitted models

Use one frequency matrix for every model so differences are not caused by
different randomly generated directions. Each p-value is a goodness-of-fit
assessment of its own model; the p-values should not be used to rank models.

```{r all-model-example, eval=FALSE}
models <- c("normal", "skewt", "vargamma", "invgauss", "genhyper")

shared_frequencies <- make_multiscale_cf_frequencies(
  dims = draw_shape(observed_tensor_draws),
  radii = c(0.5, 1, 2, 3),
  n_per_type = 8,
  seed = 20260722
)

all_model_tests <- setNames(
  lapply(models, function(current_model) {
    tensor_cf_multiscale_bootstrap(
      draws = observed_tensor_draws,
      model = current_model,
      B = 999,
      frequencies = shared_frequencies,
      max_iter = 1000,
      tol = 1e-4,
      seed = 20260722,
      show_progress = TRUE
    )
  }),
  models
)

all_model_results <- data.frame(
  model = models,
  statistic = vapply(
    all_model_tests,
    function(test) unname(test$statistic),
    numeric(1)
  ),
  p_value = vapply(
    all_model_tests,
    function(test) test$p.value,
    numeric(1)
  ),
  successful_bootstraps = vapply(
    all_model_tests,
    function(test) test$B_successful,
    integer(1)
  ),
  failed_bootstraps = vapply(
    all_model_tests,
    function(test) test$B_failed,
    integer(1)
  )
)

all_model_results
```

# 9. Interpret the result

A small p-value indicates that the empirical characteristic function differs
from the fitted model by more than expected under the fitted null distribution.
Failing to reject does not establish that the distribution is correct; it
means that this set of multiscale CF features did not detect a discrepancy.

Always inspect:

- `p.value`, the add-one bootstrap p-value;
- `B_successful` and `B_failed`, which reveal refitting stability;
- `failure_messages`, if any refits failed; and
- `squared_difference`, which shows which frequencies contributed most.
