---
title: 'test_rtnorm'
output: pdf_document
---

```{r, include = FALSE}
#library(tensormodels)
library(tidyverse)
#library(patchwork)
#library(MVN)
#library(MASS)
```

# Normal Dist Mahalanobis + KS

## Simulate matrix variate data

```{r}
n <- 1e3

s1 <- matrix(rnorm(4), nrow = 2) |> crossprod()
s2 <- matrix(rnorm(9), nrow = 3) |> crossprod()

# univar normal
uvnorm_draws <- rtnorm(n, mu = 1, sigmas = list(4))
univar_distances <- mahalanobis_dist(uvnorm_draws)

# matrix normal
matnorm_draws <- rtnorm(n, mu = matrix(1:6, nrow = 2),
                        sigmas = list(s1, s2))
matnorm_distances <- mahalanobis_dist(matnorm_draws)

# multivariate normal draws
s_multivar <- matrix(rnorm(6*6), nrow = 6) |> crossprod()
multivar_draws <- rtnorm(n, mu = 1:6, sigmas = list(s_multivar))

# make a 2x3 array
multivar_draws_3dim <- 
  lapply(multivar_draws, function(x) { 
    array(x, dim = c(2, 3))
  })

multivar_distances <- mahalanobis_dist(multivar_draws_3dim)

univarnorm_test <- mahalanobis_test(univar_distances)
matnorm_test <- mahalanobis_test(matnorm_distances)
multivarnorm_test <- mahalanobis_test(multivar_distances)
```

## Plot matrix variate results
```{r}
#should this be an object with default plot behavior?
par(mfcol = c(1, 3))

plot_dist(univar_distances)
title(main = "Univar")
with(univarnorm_test, mtext(text = sprintf("D = %.3f, p = %.3f", statistic, p.value)), 
     side = 3, line = 1)

plot_dist(matnorm_distances)
title(main = "Matnorm")
with(matnorm_test, mtext(text = sprintf("D = %.3f, p = %.3f", statistic, p.value)),
     side = 3, line = 1)

matnorm_test$p.value
multivarnorm_test$p.value

plot_dist(multivar_distances)
title(main = "Multivar")
with(multivarnorm_test, mtext(text = sprintf("D = %.3f, p = %.3f", statistic, p.value)),
     side = 3, line = 1)
```


## Simulate tensor variate data
```{r}
s3 <- matrix(rnorm(16), nrow = 4) |> crossprod()

tensorvarnorm_draws <- rtnorm(n, mu = array(1:(2*3*4), dim = c(2, 3, 4)), 
                              sigmas = list(s1, s2, s3))

tensor_distances <- mahalanobis_dist(tensorvarnorm_draws)
tensornorm_test <- mahalanobis_test(tensor_distances)
```


## Plot tensor variate results

```{r}
par(mfcol = c(1, 2))

plot_dist(tensor_distances)
title(main = "Tensor variate")
with(tensornorm_test, mtext(text = sprintf("D = %.3f, p = %.3f", statistic, p.value)),
     side = 3, line = 1)

# multivariate normal draws
s_multivar_12 <- matrix(rnorm(12*12), nrow = 12) |> crossprod()
multivar12_draws <- rtnorm(n, mu = 1:12, sigmas = list(s_multivar_12))

# make a 2x3x4 array
multivar12_draws_3dim <- 
  lapply(multivar12_draws, function(x) { 
    array(x, dim = c(2, 3, 4))
  })

multivar12_distances <- mahalanobis_dist(multivar12_draws_3dim)

tensornorm_test <- mahalanobis_test(multivar12_distances)
plot_dist(multivar12_distances)
title(main = "Multivar")
with(tensornorm_test, mtext(text = sprintf("D = %.3f, p = %.3f", statistic, p.value)),
     side = 3, line = 1)
```


# LRT Test
```{r}
lrt_test <- function(draws, restrict, model) {
  draws_mle <- tensor_mle(draws, model = model)
  draws_restrict_mle <- tensor_mle(draws, restrict = restrict, model = model)
  
  n <- length(draws) 
  
  tot_log <- 0 #log lik
  tot_log_restrict <- 0
  
  for(i in 1:n) { #iterate over all samples
    tot_log <- tot_log + dtnorm(draws[[i]], draws_mle$mu, 
                                draws_mle$sigmas, log = TRUE)
    
    tot_log_restrict <- tot_log_restrict + 
                        dtnorm(draws[[i]], draws_restrict_mle$mu, 
                               draws_restrict_mle$sigmas, log = TRUE)
  }
  
  dim_draws <- dim(draws[[1]])
  
  unrestrict <- dim_draws[-restrict]

  param <- prod(dim_draws) + 
           sum(sapply(dim_draws, function(p) p*(p + 1) / 2)) - length(dim_draws) + 1
  
  param_restrict <- prod(dim_draws) + 
                    sum(sapply(unrestrict, function(p) p*(p + 1) / 2)) - 
                    length(unrestrict) + 1 + length(restrict) # one free param for scalar of identity

  df <- param - param_restrict

  stat <- 2 * (tot_log - tot_log_restrict)
  pval <- pchisq(stat, df = df, lower.tail = FALSE)
  list(stat = stat, df = df, p.value = pval,
       first_loglik = tot_log,
       second_loglik = tot_log_restrict)
}
 
# lrt_test <- function(draws_one, draws_two, model) {
#   draws_one_mle <- tensor_mle(draws_one, model = model)
#   draws_two_mle <- tensor_mle(draws_two, model = model)
#   
#   n1 <- length(draws_one) #num of samples
#   n2 <- length(draws_two) 
#   
#   tot_log_one <- 0 #log lik
#   tot_log_two <- 0
#   
#   for(i in 1:n1) { #iterate over all samples
#     tot_log_one <- 
#       tot_log_one + dtnorm(draws_one[[i]], draws_one_mle$mu, 
#                            draws_one_mle$sigmas, log = TRUE)
#   }
#   
#   for(j in 1:n2) {
#     tot_log_two <- 
#       tot_log_two + dtnorm(draws_two[[j]], draws_two_mle$mu, 
#                            draws_two_mle$sigmas, log = TRUE)
#   }
# 
#   dim_one <- dim(draws_one[[1]])
#   dim_two <- dim(draws_two[[1]])
#   
#   param_one <- prod(dim_one) + sum(sapply(dim_one, function(p) p*(p + 1) / 2)) - length(dim_one) + 1
#   param_two <- prod(dim_two) + sum(sapply(dim_two, function(p) p*(p + 1) / 2)) - length(dim_two) + 1
#   
#   df <- param_one - param_two
# 
#   stat <- 2 * (tot_log_one - tot_log_two)
#   pval <- pchisq(stat, df = df, lower.tail = FALSE)
#   list(stat = stat, df = df, p.value = pval,
#        first_loglik = tot_log_one,
#        second_loglik = tot_log_two)
# }
```


## Restricted model data
```{r}
simulate_null <- function(n = 1000) {

  # generate full Kronecker but with sigma^2 I_4
  s1 <- s1
  s2 <- s2
  s3 <- diag(4) * 2  # scalar identity

  draws <- rtnorm(n, mu = array(0, dim = c(2,3,4)), sigmas = list(s1, s2, s3))

  lrt_test(draws, restrict = 3, model = "normal")$p.value
}

pvals <- replicate(10, simulate_null())

mean(pvals < 0.05)
```


```{r}
univar_draws <- rtnorm(1e3*24, mu = 2) |> unlist()

univar_3dim <- vector("list", 1e3)

for(i in 1:1e3) {
  univar_3dim[[i]] <- array(univar_draws[(24*(i-1)+1):(24*(i))], dim = c(2, 3, 4))
}

lrt_test(univar_3dim, restrict = c(1, 3), model = "normal")

```


```{r}
three_dim <- rtnorm(1e3, mu = array(0, dim = c(2, 3, 4)),
                    sigmas = list(s1, s2, s3))

tic()
lrt_test(three_dim, restrict = c(3), model = "normal")
toc()
```


```{r}
three_dim_identity <- rtnorm(1e3, mu = array(0, dim = c(2, 3, 4)), 
                             sigmas = list(s1, s2, diag(4)))
  
tic()
a <- lrt_test(three_dim_identity, restrict = c(2, 3), model = "normal")
toc()
```


```{r}
simulate_null <- function(rep) {
  # generate full Kronecker but with sigma^2 I_4
  s1 <- s1
  s2 <- s2
  s3 <- diag(4) * 2  # scalar identity

  draws <- rtnorm(1000, mu = array(0, dim = c(2,3,4)), sigmas = list(s1, s2, s3))

  test <- lrt_test(draws, restrict = 3, model = "normal")
  
  with(test, tibble(rep = rep, stat = as.numeric(stat), df = df, p.value = as.numeric(p.value),
                    first_loglik = as.numeric(first_loglik), 
                    second_loglik = as.numeric(second_loglik)))
}

library(furrr)
plan(multisession, workers = 4)

res <- future_map_dfr(1:100, simulate_null, .progress = TRUE,
                      .options = furrr_options(seed = TRUE))

mean(res$p.value < 0.05)
```


```{r}
simulate_alternate <- function(rep) {
  s1 <- s1
  s2 <- s2
  s3 <- s3

  draws <- rtnorm(100, mu = array(0, dim = c(2,3,4)), sigmas = list(s1, s2, s3))

  test <- lrt_test(draws, restrict = 3, model = "normal")
  
  with(test, tibble(rep = rep, stat = as.numeric(stat), df = df, p.value = as.numeric(p.value),
                    first_loglik = as.numeric(first_loglik), 
                    second_loglik = as.numeric(second_loglik)))
}

plan(multisession, workers = 4)

res <- future_map_dfr(1:100, simulate_alternate, .progress = TRUE,
                      .options = furrr_options(seed = TRUE))

```


```{r}
#inverses of all sigma
    inv_sigma <- lapply(est_sigmas, invert_safe)

    #exclude kth mode
    inv_sigma_except_k <- inv_sigma[-k]

    # Kronecker product of inverses except k
    kron_except_k <- Reduce(function(A,B) kronecker(B, A), inv_sigma_except_k)

    # dim for kth mode
    d_k    <- dims[k]
    d_negk <- prod(dims[-k])
    s_k    <- matrix(0, d_k, d_k)

    # accumlate mode-k covar
    for (draw in 1:n) {
      Xi <- data[[draw]] - mean_array

      Xik <- matricization(Xi, k)
      s_k <- s_k + Xik %*% kron_except_k %*% t(Xik)
    }
    
inv_real <- lapply(list(s1, s2, s3), invert_safe)

inv_real_except_k <- inv_real[-k]


matricization(((test |> n_prod(2, inv_real_except_k[[1]])) |> n_prod(3, inv_real_except_k[[2]])), 1) %*%
  t(matricization(test, 1))

  # Kronecker product of inverses except k
kron_real_except_k <- Reduce(function(A,B) kronecker(B, A), inv_real_except_k)

Xik_real <- matricization(test, 1) #2x12


Xik_real %*% kron_real_except_k %*% t(Xik)

a <- n_prod(test, 2, inv_real_except_k[[1]]) |> n_prod(3, inv_real_except_k[[2]])

(Xik_real %*% kron_real_except_k)

matricization(a, 1) #%*% t(matricization(a, 1))

Xik_real %*% kron_real_except_k

matricization(a, 1)


test |> n_prod(2, inv_real_except_k[[1]]) |> n_prod(3, inv_real_except_k[[2]])
```


```{r}
```


```{r}
lrt_test(univar_3dim, restrict = c(3), model = "normal")

lrt_test(univar_3dim, restrict = c(2, 3), model = "normal")

tensor_mle(three_dim, model = "normal")

#1:24
#25:48
#49:...

unlist(univar_draws) |> 


univar_3dim <- lapply(univar_draws, function(x) {
  array(x, dim = c(2, 12))
})





lrt_test(three_dim, restrict = c(3), model = "normal")

lrt_test(multivar_draws_3dim, restrict = c(3), model = "normal")




lrt_test(three_dim, restrict = c(2, 3), model = "normal")


three_dim_mle <- tensor_mle(three_dim, model = "normal")

three_dim_restrict_mle <-
  tensor_mle(three_dim, model = "normal", restrict = c(3))

three_dim_restrict_mle$sigmas

tot_log_one <- 0

for(i in 1:n1) { #iterate over all samples
  tot_log_one <- 
    tot_log_one + dtnorm(three_dim[[i]], three_dim_mle$mu, 
                         three_dim_mle$sigmas, log = TRUE)
}

tot_log_two <- 0  

for(j in 1:n2) {
  tot_log_two <- 
    tot_log_two + dtnorm(three_dim[[j]], three_dim_restrict_mle$mu, 
                         three_dim_restrict_mle$sigmas, log = TRUE)
}

tot_log_one
tot_log_two

two_dim <- lapply(three_dim, function(x) {
  array(x, dim = c(2, 12))
})

one_dim <- lapply(three_dim, function(x) {
  array(x, dim = 24)
})

lrt_test(two_dim, three_dim, model = "normal")
lrt_test(one_dim, two_dim, model = "normal")

Sigma_full_1 <- Reduce(kronecker, three_dim_mle$sigmas)
Sigma_full_2 <- Reduce(kronecker, three_dim_restrict_mle$sigmas)

# compare traces / determinants (use logdet for stability)
tr1 <- sum(diag(Sigma_full_1))
tr2 <- sum(diag(Sigma_full_2))

tr1; tr2; tr2/tr1

```


## Tensor variate normal data

```{r}
tensorvarnorm_draws <- rtnorm(n, mu = array(1:(2*3*4), dim = c(2, 3, 4)), 
                              sigmas = list(s1, s2, s3))
  
tensornorm_twodim <- 
  lapply(tensorvarnorm_draws, function(x) {
  array(x, dim = c(12, 2))
})

tensornorm_onedim <- 
  lapply(tensorvarnorm_draws, function(x) {
  array(x, dim = c(24))
})
  
lrt_test(tensornorm_onedim, tensorvarnorm_draws, model = "normal")

draws_two_mle <- tensor_mle(tensorvarnorm_draws, model = "normal")

emp_cov <- cov(do.call(rbind, lapply(tensorvarnorm_draws, c)))
kroneck_est <- Reduce(kronecker, draws_two_mle$sigmas)
mean(emp_cov / kroneck_est)
```

```{r}
n <- 1e3

# tensor variate random normal draws
s1 <- matrix(rnorm(4), nrow = 2) |> crossprod()
s2 <- matrix(rnorm(9), nrow = 3) |> crossprod()
s3 <- matrix(rnorm(16), nrow = 4) |> crossprod()

rand_mu <- array(rnorm(24, mean = 10), dim = c(2, 3, 4))

rtnorm_draws <- rtnorm(n = n, mu = rand_mu, sigmas = list(s1, s2, s3))

rtnorm_mles <- mle_est(rtnorm_draws)

centered <- (rtnorm_draws - array(rep(rtnorm_mles$mu, each = n), 
                                  dim = c(n, 2, 3, 4))) |> 
  as_tensor()

centered[1, ] %x1% s1 %x2% s2 %x3% s3

n_prod(centered[1,], 1, s1) |> n_prod(2, s2) |> n_prod(3, s3)
```


```{r}
n <- 1e3

# tensor variate random normal draws
s1 <- matrix(rnorm(4), nrow = 2) |> crossprod()
s2 <- matrix(rnorm(9), nrow = 3) |> crossprod()
s3 <- matrix(rnorm(16), nrow = 4) |> crossprod()

rand_mu <- array(rnorm(24, mean = 10), dim = c(2, 3, 4))

rtnorm_draws <- rtnorm(n = n, mu = rand_mu, sigmas = list(s1, s2, s3))

# univariate random normal draws

univar_draws <- rnorm(n = 1e3 * 2 * 3 * 4) |> array(dim = c(1e3, 2, 3, 4))

# multivariate normal draws
Sigma <- matrix(c(rnorm(24*24)), nrow = 24, ncol = 24)
Sigma[lower.tri(Sigma)] = t(Sigma)[lower.tri(Sigma)]
Sigma <- crossprod(Sigma, Sigma)

mu_vec <- rnorm(24, mean = 10, sd = 2)

X_vec <- MASS::mvrnorm(n, mu = mu_vec, Sigma = Sigma)

# reshape into 2x3x4 matrices
multivar_draws <- array(0, dim = c(n, 2, 3, 4))

for(i in 1:n) {
  multivar_draws[i, , ,] <- array(data = X_vec[i, ], dim = c(2, 3, 4))
}

#univar_draws
```


# Test rtskewt draws
```{r}
n <- 1e3
nu <- 4

rand_skew <- array(rnorm(24, mean = 10), dim = c(2, 3, 4))

rtskewt_draws <- rtskewt(n = n, mu = rand_mu, 
                         skew = rand_skew, sigmas = list(s1, s2, s3),
                         nu = nu)

mles <- em_est(rtskewt_draws, quiet = FALSE, 
               max_iter = 5000, model = "skewt", tol = 0.1)

sum(abs(mles$mu - rand_mu))
sum(abs(mles$skew - rand_skew))


orig_data <- 
  kronecker(s3, s2) |> kronecker(s1) |> 
  as_tibble() |> 
  mutate(row = row_number()) |> 
  pivot_longer(cols = V1:V24) |> 
  mutate(col = as.numeric(str_remove(name, "V")),
         origin = "Orig") 

avg_orig <- orig_data$value |> mean()

sim_data <- 
  kronecker(mles$sigmas[[3]], mles$sigmas[[2]]) |> 
  kronecker(mles$sigmas[[1]]) |> 
  as_tibble() |> 
  mutate(row = row_number()) |> 
  pivot_longer(cols = V1:V24) |> 
  mutate(col = as.numeric(str_remove(name, "V")),
         origin = "Sim") |> 
  rename(unscale_value =value)

avg_sim <- sim_data$unscale_value |> mean()

sim_data <- sim_data |> mutate(value = unscale_value * avg_orig/avg_sim)

sim_data |> 
  bind_rows(orig_data) |> 
  ggplot() +
  geom_tile(aes(row, col, fill = value), color = "black") +
  facet_grid(~origin) +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(x = "Row", y = "Col")

# true_mean <- rand_mu + (nu)/(nu-2) * rand_skew
# model_mean <- with(mles, mu + (nu)/(nu-2) * skew)
# sample_mean <- apply(rtskewt_draws, 2:4, mean)
```



```{r}
#mles$nu

mles_invgauss <- em_est(rtskewt_draws, model = "invgauss", 
                        quiet = FALSE, max_iter = 5000, tol = 0.1)

mles$loglik
mles_invgauss$loglik

mles$BIC
mles_invgauss$BIC

centered <- rtskewt_draws

for(i in 1:n) {
  centered[i, , ,] <- 
    (rtskewt_draws[i, , ,] - mles$mu - mles$Ew[i] * mles$skew)/sqrt(mles$Ew[i])
}



mles_invgauss$BIC

centered_invgauss <- rtskewt_draws

for(i in 1:n) {
  centered_invgauss[i, , ,] <- 
    (rtskewt_draws[i, , ,] - mles_invgauss$mu - 
       mles_invgauss$Ew[i] * mles_invgauss$skew)/sqrt(mles_invgauss$Ew[i]) 
}

rtinvgauss_draws <- rtinvgauss(n = n, mu = rand_mu, 
                               skew = rand_skew, sigmas = list(s1, s2, s3))

```


```{r}
whiten_residuals <- function(draws, fit) {
  n <- dim(draws)[1]
  D <- length(dim(draws)) - 1
  
  out <- draws
  for (i in 1:n) {
    V <- (draws[i,,,] - fit$mu - fit$Ew[i] * fit$skew) / sqrt(fit$Ew[i])

    # whiten by Δ_d^{-1/2}
    for (d in rev(1:D)) {
      #R <- chol(fit$sigmas[[d]])          # Δ_d^{1/2}
      V <- n_prod(V, d, chol2inv(chol(fit$sigmas[[d]])))  # multiply by Δ_d^{-1/2}
    }
    out[i,,,] <- V
  }
  out
}

Z_skewt <- whiten_residuals(rtskewt_draws, mles)
Z_invgauss <- whiten_residuals(rtskewt_draws, mles_invgauss)

Z_skewt |> 
  c() |> 
  as_tibble() |>   
  #mutate(value_sq = value^2) |> 
  ggplot() +
  geom_histogram(aes(x = value, y = after_stat(density)), color = "black", fill = "pink") +
  #geom_function(fun = dnorm, color = "red") +
  theme_minimal()
#BIC=−2ℓ(θ^)+klogn

Z_invgauss |> 
  c() |> 
  as_tibble() |> 
  ggplot() +
  geom_histogram(aes(x = value), color = "black", fill = "pink") +
  theme_minimal()


tensor_mahalanobis_plot_normal(Z_skewt) +
  labs(title = "rtskewt") +
tensor_mahalanobis_plot_normal(Z_invgauss) +
  labs(title = "invgauss") 

# for(i in 1:n) {
#   centered[i, , ,] <- 
#     (rtskewt_draws[i, , ,] - mles$mu - mles$Ew[i] * mles$skew)/sqrt(mles$Ew[i]) 
# }

mles_genhyper <- em_est(rtskewt_draws, model = "genhyper", 
                        quiet = FALSE, max_iter = 5000)

centered_genhyper <- rtskewt_draws

for(i in 1:n) {
  centered_genhyper[i, , ,] <- 
    (rtskewt_draws[i, , ,] - mles_genhyper$mu - 
       mles_genhyper$Ew[i] * mles_genhyper$skew)/sqrt(mles_genhyper$Ew[i]) 
}

mles_vargamma <- em_est(rtskewt_draws, model = "vargamma", 
                        quiet = FALSE, max_iter = 5000)

centered_vargamma <- rtskewt_draws

for(i in 1:n) {
  centered_vargamma[i, , ,] <- 
    (rtskewt_draws[i, , ,] - mles_vargamma$mu - 
       mles_vargamma$Ew[i] * mles_vargamma$skew)/sqrt(mles_vargamma$Ew[i]) 
}
```


```{r}
tensor_mahalanobis_plot_normal(centered) +
  labs(title = "rtskewt") +
tensor_mahalanobis_plot_normal(centered_invgauss) +
  labs(title = "invgauss") +
  xlim(0, 300) +
  ylim(0, 1000)
tensor_mahalanobis_plot(centered_genhyper) +
  labs(title = "genhyper") +
tensor_mahalanobis_plot(centered_vargamma) +
  labs(title = "vargamma") 
```


```{r}

mle_covar <- kronecker(mles$sigmas[[3]], mles$sigmas[[2]]) |> kronecker(mles$sigmas[[1]])

real_covar <- kronecker(s3, s2) |> kronecker(s1)

mle_covar[1, ]
real_covar[1, ] / 3
```


# Sphericity
$H_0: \Sigma_{ij} = 0$ for all $1 \le i < j \le k$
$H_a: H_0$ is not true
```{r}
bartlett_diagonal_test <- function(X) {
  X <- as.matrix(X)
  n <- nrow(X); p <- ncol(X)
  #if (n <= p) stop("Need n > p (and MVN assumptions) for the classic Bartlett/LRT to behave well.")

  R <- cor(X, use = "pairwise.complete.obs")
  detR <- det(R)

  # Bartlett-corrected LR chi-square
  cfac <- (n - 1) - (2*p + 5)/6
  stat <- -cfac * log(detR)
  df <- p*(p - 1)/2
  pval <- 1 - pchisq(stat, df)

  list(statistic = stat, df = df, p.value = pval, detR = detR)
}

rtnorm_draws_mle <- mle_est(rtnorm_draws)


bartlett_diagonal_test(rtnorm_draws_mle$sigmas[[1]])


bartlett_diagonal_test(univar_mle_est$sigmas[[1]])
```


```{r}
test_stat <- rep(0, 100)

n_minus <- n - 1

for(iter in 1:100) {
  univar_draws <- rnorm(n = 1e3 * 2 * 3 * 4) |> array(dim = c(1e3, 2, 3, 4))
  
  univar_mle_est <- mle_est(univar_draws)
   
  corr_1 <- cov2cor(univar_mle_est$sigmas[[1]])
  
  m <- nrow(corr_1)
    
  sum_r <- sum(corr_1[lower.tri(corr_1)]^2)
  # 
  # for(i in 2:m) {
  #   for(j in 1:(i-1)) {
  #     sum_r <- sum_r + corr_1[i, j]^2
  #   }
  # }
  
  t_test <- sum_r - ((m * (m-1))/(2*n_minus))
  
  sigma_sq <- (m * (m-1) * (n_minus-1))/(n_minus^2 * (n_minus+2))
  
  z_test <- t_test/sqrt(sigma_sq)
  
  test_stat[iter] <- z_test
  
  if(iter %% 20 == 0) {
    print(iter)
  }
}

test_stat |> 
  as_tibble() |> 
  ggplot() +
  geom_histogram(aes(x = value), bins = 30, 
                 color = "black", fill = "pink") +
  theme_minimal()
```


```{r}
if (z_test < 0) {
  2 * pnorm(q = z_test)
} else {
  2 * pnorm(q = z_test, lower.tail = FALSE)
}
```


```{r}
pnorm(q = abs(z_test))

if(z_test < 0) {
  2 * pnorm(q = z_test)
}


z_test

univar_mle_est$sigmas[[1]]

univar_mle_est$sigmas[[2]]
univar_mle_est$sigmas[[3]]

sig_1 <- univar_mle_est$sigmas[[1]]

W_n <- det(sig_1)^((n+1)/2)/(prod(diag(sig_1))^((n+1)/2)) 

lambda_n <- (n+1)/2 * log(W_n)

rho <- 1 - (2 * (2^3 - 2^3 + 2^3) + 9 * (2^3 - 2^2 + 2^2))/(6 * (n+1) * (2^2 - 2 + 2))

f <- 1/2 * (2^2 - 2^2 + 2^2)

1 - pchisq(-2 * rho * lambda_n, df = f)
```


```{r}

tensor_mahalanobis_plot(X_array)
tensor_mahalanobis_plot(rtnorm_draws_diag)
tensor_mahalanobis_plot(flatten_draws)

test_draws <- rtnorm(1e3, mu = rand_mu, sigmas = list(s1, s2))

test_draws |> dim()
```


```{r}




# tensor variate random normal draws
s1 <- matrix(rnorm(4), nrow = 2); s2 <- diag(3); s3 <- matrix(rnorm(16), nrow = 4)
s1 <- crossprod(s1, s1); s3 <- crossprod(s3, s3)

rand_mu <- array(rnorm(24, mean = 10), dim = c(2, 3, 4))

rtnorm_draws <- rtnorm(n = n, mu = rand_mu, sigmas = list(s1, s2, s3))

flatten_draws <- rtnorm_draws_diag |> array(dim = c(1000, 6, 4))

(tensor_mahalanobis_plot(rtnorm_draws) + labs(title = "Tensor")) +
(tensor_mahalanobis_plot(rtnorm_draws_diag) + labs(title = "Matrix")) +
(tensor_mahalanobis_plot(X_array) + labs(title = "Multivar")) +
theme_minimal()


tensor_mahalanobis_plot(flatten_draws) 

rtnorm_draws_diag

tensor_mahalanobis_plot(rtnorm_draws_diag)
```


```{r}
n <- 1e3

s1 <- matrix(rnorm(4), nrow = 2); s2 <- matrix(rnorm(9), nrow = 3); s3 <- matrix(rnorm(16), nrow = 4)
s1 <- crossprod(s1, s1); s2 <- crossprod(s2, s2); s3 <- crossprod(s3, s3)
rand_mu <- array(rnorm(24, mean = 10), dim = c(2, 3, 4))

# random normal draws
rtnorm_draws <- rtnorm(n = n, mu = rand_mu, sigmas = list(s1, s2, s3))
```


```{r}
sample_mean <- apply(rtnorm_draws, 2:4, mean)

dims <- dim(sample_mean)

sample_stats <- mle_est(rtnorm_draws)

sample_s1 <- sample_stats$sigmas[[1]]
sample_s2 <- sample_stats$sigmas[[2]]
sample_s3 <- sample_stats$sigmas[[3]]

all_dist <- rep(0, n)

for(i in 1:n) {
  centered <- rtnorm_draws[i, , , ] - sample_mean # center data
  
  centered_multiply <- centered
  
  for(j in 1:length(dims)) { # standardize with sigma estimates
    centered_multiply <- n_prod(centered_multiply, 
                                j, solve(sample_stats$sigmas[[j]]))
  }
  
  all_dist[i] <- c(centered) %*% c(centered_multiply)
}

draws <- matrix(0, nrow = n, ncol = 24)

for(i in 1:n) {
  draws[i, ] <- c(rtnorm_draws[i, , ,]) # vectorize data
}

mu_vec <- apply(draws, MARGIN = 2, FUN = mean)

vec_dist <- mahalanobis(draws, center = mu_vec, cov = cov(draws))

c_hat <- sum(diag(cov(draws)))/24

all_dist <- all_dist / c_hat

tibble("vec" = (vec_dist),
       "mat" = (all_dist)) |> 
  ggplot() +
  geom_point(aes(x = vec, y = mat)) +
  geom_abline(slope = 1, color = "red") +
  labs(x = "Multivariate", y = "Tensor Variate", 
       title = "Tensor Variate Data") +
  theme_minimal()
```


```{r}
# multivariate normal draws
Sigma <- matrix(c(rnorm(24*24)), nrow = 24, ncol = 24)

Sigma[lower.tri(Sigma)] = t(Sigma)[lower.tri(Sigma)]

Sigma <- crossprod(Sigma, Sigma)

mu_vec <- rnorm(24, mean = 10, sd = 2)

X_vec <- mvrnorm(n, mu = mu_vec, Sigma = Sigma)

# reshape into 2x3x4 matrices
X_array <- array(0, dim = c(n, 2, 3, 4))

for(i in 1:n) {
  X_array[i, , ,] <- array(data = X_vec[i, ], dim = c(2, 3, 4))
}

#rtnorm_draws <- X_array
```



```{r}
# generate vectorized matrices
n <- 1000
X_vec <- mvrnorm(n, mu = mu_vec, Sigma = Sigma)

# reshape into 2x3 matrices
X_array <- array(0, dim = c(n, 2, 3, 4))
for(i in 1:n) {
  X_array[i,,] <- array(X_vec[i,], dim = c(2, 3, 4))
}

rtnorm_draws <- X_array

draws <- matrix(0, nrow = n, ncol = 6)

for(i in 1:n) {
  draws[i, ] <- c(rtnorm_draws[i, , ])
}

mu_vec <- apply(draws, MARGIN = 2, FUN = mean)

vec_dist <- mahalanobis(draws, center = mu_vec, cov = cov(draws))

est_norm <- mle_est(rtnorm_draws)

mean_mat <- est_norm$mu

U <- est_norm$sigmas[[1]]
V <- est_norm$sigmas[[2]]

mat_dist <- rep(0, 1000)

c_hat <- sum(diag(cov(draws)))/sum(diag(kronecker(U, V)))
 
for(i in 1:1000) {
  Xi <- rtnorm_draws[i, , ] - mean_mat
  mat_dist[i] <- sum(diag(solve(U) %*% Xi %*% solve(V) %*% t(Xi)))
}

mat_dist <- mat_dist / c_hat

tibble("vec" = (vec_dist),
       "mat" = (mat_dist)) |> 
  ggplot() +
  geom_point(aes(x = vec, y = mat)) +
  geom_abline(slope = 1, color = "red") +
  labs(x = "Multivariate", y = "Matrix Variate", title = "Not Matrix Variate") +
  theme_minimal()
```


```{r}
ggplot() + 
  geom_histogram(aes(x = all_dist,
                     y = after_stat(density)),
                 fill = "pink", color = "black", bins = 60) +
  geom_function(fun = dchisq, args = list(df = 2), color = "red") +
  labs(x = "Distances", y = "Count") + 
  theme_minimal()
```


```{r}
draws <- matrix(0, nrow = n, ncol = 6)

for(i in 1:n) {
  draws[i, ] <- c(rtnorm_draws[i, , ])
}

mu_vec <- apply(draws, MARGIN = 2, FUN = mean)

vec_dist <- mahalanobis(draws, center = mu_vec, cov = cov(draws))

est_norm <- mle_est(rtnorm_draws)

mean_mat <- est_norm$mu

U <- est_norm$sigmas[[1]]
V <- est_norm$sigmas[[2]]

mat_dist <- rep(0, 1000)

c_hat <- sum(diag(cov(draws)))/sum(diag(kronecker(U, V)))
 
for(i in 1:1000) {
  Xi <- rtnorm_draws[i, , ] - mean_mat
  mat_dist[i] <- sum(diag(solve(U) %*% Xi %*% solve(V) %*% t(Xi)))
}

mat_dist <- mat_dist / c_hat

tibble("vec" = (vec_dist),
       "mat" = (mat_dist)) |> 
  ggplot() +
  geom_point(aes(x = vec, y = mat)) +
  geom_abline(slope = 1, color = "red") +
  labs(x = "Multivariate", y = "Matrix Variate", 
       title = "Matrix Variate") +
  theme_minimal()
```


# Multivariate Normal Tests

## Mardia's test

This test is based on skewness and kurtosis. Given we observe $k$ samples of $X_{1}, X_{2}, \dots, X_{k}$ we compute 

\begin{align*}
\hat{\Sigma} &= \frac{1}{n} \sum_{j=1}^{n} (x_{j} - \bar{x})(x_{j} - \bar{x})^{T} \\
A &= \frac{1}{6n} \sum_{i=1}^{n} \sum_{j=1}^{n} \left[(x_{i} - \bar{x}) \hat{\Sigma}^{-1} (x_{j} - \bar{x})^{T}\right]^{3} \\
B &= \sqrt{\frac{n}{8k(k+2)}} \left\{\frac{1}{n} \sum_{i=1}^{n} \left[(x_{i} - \bar{x})^{T} \hat{\Sigma}^{-1} (x_{i} - \bar{x})\right]^{2}\right\}
\end{align*}

Under $H_0: X \sim N_{p}(\mu, \sigma), A \sim \chi^2_{1/6 k (k+1) (k+2)}$ and $B \sim N(0, 1)$. 

```{r}
MVN::mardia(iris[1:50, 1:4])
```


## Malahanobis Distance QQ Plot

The variable $d^{2} = (\boldsymbol{x} - \boldsymbol{\mu})^{T} \boldsymbol{\Sigma}^{-1} (\boldsymbol{x} - \boldsymbol{\mu}) \sim \chi^2_{p}$. We can plot the ordered Malahanobis distances versus the esimated quantiles on a QQ plot. 

```{r}
MVN::multivariate_diagnostic_plot(iris[1:50, 1:4])
```

## Eigenvalue 

Assume we observe $n$ samples of $X_{1}, X_{2}, \dots, X_{n}$ and that the eigenvalues of $\Sigma$ are distinct and positive such that $\lambda_{1} > \lambda_{2} > \dots > \lambda_{p} > 0$. Denote the vector of eigenvalues as $\mathbf{\lambda}$ and the corresponding eigenvectors as $\hat{\mathbf{e}}_{1}, \dots \hat{\mathbf{e}}_{1}$.

Let $\mathbf{\Lambda}$ is a diagonal matrix of the eigenvalues. Then $\sqrt{n}(\hat{\mathbf{\lambda}} - \mathbf{\lambda}) \simeq N_{p}(0, 2 \mathbf{\Lambda}^2)$ 

Let $$E_{i} = \lambda_{i} \sum_{k=1, k \neq i}^{p} \frac{\lambda_{k}}{(\lambda_{k} - \lambda_{i})^2} \mathbf{e}_{k} \mathbf{e}_{k}^{T}$$. Then $\sqrt{n}(\hat{\mathbf{e_{i}}} - \mathbf{e_{i}}) \simeq N_{p}(0, E_{i})$.

Each $\hat{\lambda}_{i}$ is distributed independently of elements of associated $\hat{\mathbf{e}_{i}}.$

However, in practice, this large sample approximation converges very slowly. The asymptitc distributions depend on the true eigenstructure of $\Sigma$ too. 

# Matrix Variate Normal Tests

## Malahanobis Distance QQ Plot

We can calculate the Malahanobis Distance for a multivariate normal model as $$D(\boldsymbol{X}_{i}, \hat{\boldsymbol{\mu}}, \hat{\boldsymbol{\Sigma}}) = (\text{vec}(\boldsymbol{X}_{i}) - \boldsymbol{\mu})^{T} \boldsymbol{\Sigma}^{-1} (\text{vec}(\boldsymbol{X}_{i}) - \boldsymbol{\mu})$$. 


If a Kronecker product structure exists for $\Sigma$, then $$D(\boldsymbol{X}_{i}, \hat{\boldsymbol{\mu}}, \hat{\boldsymbol{\Sigma}}) = D_{M}(\boldsymbol{X}_{i}, \boldsymbol{M}, \boldsymbol{V}, \boldsymbol{U})$$.

I was wondering how this would be generalizable for the tensor variate normal. For example, if each $X_{i}$ draw is a $2 \times 3 \times 4$. Then we would have three covariance estimates: $\boldsymbol{\Sigma}_{1}, \boldsymbol{\Sigma}_{2}, \boldsymbol{\Sigma}_{3}$ Instead of using the matrix product, we would use the $n$-mode product between the draws and the covariance matrices.

Since we take the trace of the matrix product, we only care about the diagonals of the resulting object. I am unsure what the generalization of that in the tensor case would be. The structure reminds me of the hyperdiagonal tensor used in the PARAFAC decomposition though. 

Would we calculate the Malahanobis Distance for a tensor variate normal model as something like the following? $$D_{T}(\mathcal{X}_{i}, \mathcal{M}, \boldsymbol{V}, \boldsymbol{U}) = \text{tr}\{(\mathcal{X}_{i} - \mathcal{M}) \times_{1} \boldsymbol{\Sigma}_{1}^{-1} \times_{2} \boldsymbol{\Sigma}_{2}^{-2} \times_{3} \boldsymbol{\Sigma}_{3}^{-1} (\mathcal{X}_{i} - \mathcal{M})^{T}\}.$$

```{r}
n <- 1e3

s1 <- matrix(rnorm(4), nrow = 2); s2 <- matrix(rnorm(9), nrow = 3); 
s1 <- crossprod(s1, s1); 
s2 <- crossprod(s2, s2); 
list_sigmas <- list(s1, s2)
rand_mu <- array(rnorm(6, mean = 10), dim = c(2, 3))
rtnorm_draws <- rtnorm(n = n, mu = rand_mu, sigmas = list(s1, s2))

draws <- matrix(0, nrow = n, ncol = 6)

for(i in 1:n) {
  draws[i, ] <- c(rtnorm_draws[i, , ])
}

mu_vec <- apply(draws, MARGIN = 2, FUN = mean)

vec_dist <- mahalanobis(draws, center = mu_vec, cov = cov(draws))

est_norm <- mle_est(rtnorm_draws)

mean_mat <- est_norm$mu

U <- est_norm$sigmas[[1]]
V <- est_norm$sigmas[[2]]

mat_dist <- rep(0, 1000)

c_hat <- sum(diag(cov(draws)))/sum(diag(kronecker(U, V)))
 
for(i in 1:1000) {
  Xi <- rtnorm_draws[i, , ] - mean_mat
  mat_dist[i] <- sum(diag(solve(U) %*% Xi %*% solve(V) %*% t(Xi)))
}

mat_dist <- mat_dist / c_hat

tibble("vec" = (vec_dist),
       "mat" = (mat_dist)) |> 
  ggplot() +
  geom_point(aes(x = vec, y = mat)) +
  geom_abline(slope = 1, color = "red") +
  labs(x = "Multivariate", y = "Matrix Variate", 
       title = "Matrix Variate") +
  theme_minimal()
```


```{r}
Sigma <- matrix(c(
  2, 1, 0.5, 0.1, 0,   0.5,
  1, 3, 1,   0.3, 0.2, 0.4,
  0.5, 1, 4, 0.5, 0.1, 0.7,
  0.1, 0.3, 0.5, 2,   0.8, 0.2,
  0,   0.2, 0.1, 0.8, 3,   1.2,
  0.5, 0.4, 0.7, 0.2, 1.2, 5
), nrow = 6, byrow = TRUE)

# mean for vec form
mu_vec <- rep(0, 6)

# generate vectorized matrices
n <- 1000
X_vec <- mvrnorm(n, mu = mu_vec, Sigma = Sigma)

# reshape into 2x3 matrices
X_array <- array(0, dim = c(n, 2, 3))
for(i in 1:n) {
  X_array[i,,] <- matrix(X_vec[i,], nrow = 2)
}

rtnorm_draws <- X_array

draws <- matrix(0, nrow = n, ncol = 6)

for(i in 1:n) {
  draws[i, ] <- c(rtnorm_draws[i, , ])
}

mu_vec <- apply(draws, MARGIN = 2, FUN = mean)

vec_dist <- mahalanobis(draws, center = mu_vec, cov = cov(draws))

est_norm <- mle_est(rtnorm_draws)

mean_mat <- est_norm$mu

U <- est_norm$sigmas[[1]]
V <- est_norm$sigmas[[2]]

mat_dist <- rep(0, 1000)

c_hat <- sum(diag(cov(draws)))/sum(diag(kronecker(U, V)))
 
for(i in 1:1000) {
  Xi <- rtnorm_draws[i, , ] - mean_mat
  mat_dist[i] <- sum(diag(solve(U) %*% Xi %*% solve(V) %*% t(Xi)))
}

mat_dist <- mat_dist / c_hat

tibble("vec" = (vec_dist),
       "mat" = (mat_dist)) |> 
  ggplot() +
  geom_point(aes(x = vec, y = mat)) +
  geom_abline(slope = 1, color = "red") +
  labs(x = "Multivariate", y = "Matrix Variate", title = "Not Matrix Variate") +
  theme_minimal()
```


```{r}

```


## Likelihood Ratio Tests for Sigma

Using the paper Likelihood Ratio Tests for High-Dimensional Normal Distributions by Jiang and Qi (2015), they suggest a spherical test. For a multivariate normal distribution $N_{p}(\mathbf{\mu}, \mathbf{\Sigma})$ where $\mathbf{\mu} \in \mathbb{R}^{p}$ is a mean vector and $\mathbf{\Sigma}$ is a $p \times p$ covariance matrix, we consider this spherical test. 

$$H_{0}: \mathbf{\Sigma} = \lambda \textbf{I}_{p} \text{ vs } H_{a}: \mathbf{\Sigma} \neq \lambda \textbf{I}_{p}$$

with unspecified $\lambda$. 

Suppose $\textbf{x}_{1}, \dots, \textbf{x}_{n}$ are iid $\mathbb{R}^{p}$ from the normal distribution. As usual, set

$$\bar{\textbf{x}} = \frac{1}{n} \sum_{i=1}^n x_i \text{ and } \textbf{S} = \frac{1}{n} \sum_{i=1}^{n} (\textbf{x}_{i} - \bar{\textbf{x}})(\textbf{x}_{i} - \bar{\textbf{x}})^{T}.$$ 

Then the LRT statistic for this test is 

$$V_{n} = |\textbf{S}| \left(\frac{\text{tr}(\textbf{S})}{p}\right)^{-p}.$$

Then $-(n-1) \rho \log{V_{n}}$ converges to $\chi^{2}_{f}$ as $n \to \infty$ where

$$\rho = 1 - \frac{2p^2 + p + 2}{6(n-1) * p} \text{ and } f = \frac{1}{2} (p-1)(p+2).$$

```{r}
test_sigmas <- function(est_sig) {
  p_vals <- rep(0, length(est_sig))
  
  for(d in 1:length(est_sig)) {
    curr_sig <- est_sig[[d]]
    p <- dim(curr_sig)[[1]]
    
    rho <- 1 - (2 * p^2 + p + 2)/(6 * (n-1) * p)
    f <- 1/2 * (p - 1) * (p + 2)
    
    Vn <- det(curr_sig) * (sum(diag(curr_sig))/p)^(-p)
    
    test_stat <- -(n - 1) * rho * log(Vn)
  
    p_vals[d] <- pchisq(q = test_stat, df = f, lower.tail = FALSE)
  }
  
  p_vals
}

n <- 1e3
```


Test with sigma 1 being diagonal. Since the null is that the sigma is diagonal, we expect to fail to reject for the first sigma. 
```{r}
s1 <- diag(2); s2 <- matrix(rnorm(9), nrow = 3); s3 <- matrix(rnorm(16), nrow = 4)
s1 <- 4 * s1
s2 <- crossprod(s2, s2); s3 <- crossprod(s3, s3)
list_sigmas <- list(s1, s2, s3)
rand_mu <- array(rnorm(24, mean = 10), dim = c(2, 3, 4))
rtnorm_draws <- rtnorm(n = n, mu = rand_mu, sigmas = list(s1, s2, s3))

est_mle <- mle_est(rtnorm_draws)

test_sigmas(est_mle$sigmas)
```


Test with sigma 1, 2, 3 being diagonal
```{r}
s1 <- diag(2); s2 <- diag(3); s3 <- diag(4)
s1 <- 4 * s1
s2 <- 2 * s2
s3 <- 8 * s3
rand_mu <- array(rnorm(24, mean = 10), dim = c(2, 3, 4))
rtnorm_draws <- rtnorm(n = n, mu = rand_mu, 
                       sigmas = list(s1, s2, s3))

est_mle <- mle_est(rtnorm_draws)
test_sigmas(est_mle$sigmas)
```


Test with sigma 2 being diagonal with some error
```{r}
s1 <- diag(2); s2 <- diag(3); s3 <- diag(4)
s1 <- 4 * s1
s2 <- 2 * s2 + rnorm(9, mean = 0.4, sd = 0.2)
s3 <- 8 * s3
rand_mu <- array(rnorm(24, mean = 10), dim = c(2, 3, 4))
rtnorm_draws <- rtnorm(n = n, mu = rand_mu, 
                       sigmas = list(s1, s2, s3))

est_mle <- mle_est(rtnorm_draws)
test_sigmas(est_mle$sigmas)
```

Test with sigma 1 and sigma 3 being diagonal
```{r}
s1 <- diag(2); s2 <- diag(3); s3 <- 3 * diag(4)
s1 <- 4 * s1
s2 <- 2 * s2 + rnorm(9, mean = 0.4, sd = 0.2)
s3 <- 8 * s3
rand_mu <- array(rnorm(24, mean = 10), dim = c(2, 3, 4))
rtnorm_draws <- rtnorm(n = n, mu = rand_mu, 
                       sigmas = list(s1, s2, s3))

est_mle <- mle_est(rtnorm_draws)
test_sigmas(est_mle$sigmas)
```

Test the power when sigma 1 and sigma 3 are diagonal

```{r}
rt_power <- tibble("p1" = rep(0, 1000), "p2" = rep(0, 1000), "p3" = rep(0, 1000))

list_mles <- vector("list", 3)

for(i in 1:100) {
  rtnorm_draws <- rtnorm(n = n, mu = rand_mu, 
                         sigmas = list(s1, s2, s3))
  
  est_mle <- mle_est(rtnorm_draws)
  
  list_mles[[i]] <- est_mle$sigmas
  
  curr_test <- test_sigmas(est_mle$sigmas)
  
  rt_power[i, "p1"] <- curr_test[1]
  rt_power[i, "p2"] <- curr_test[2]
  rt_power[i, "p3"] <- curr_test[3]
}

rt_power$p2[1:100] |> 
  as_tibble() |> 
  ggplot() +
  geom_histogram(aes(x = value))
```



```{r, eval = FALSE, include = FALSE}
est_mle <- mle_est(rtnorm_draws)

est_mle$sigmas
```


```{r, eval = FALSE, include = FALSE}
sig1_mle <- est_mle$sigmas[[2]]

rho <- dim(sig1_mle)[[1]]

f <- 1/2 * (rho - 1) * (rho + 2)

Vn <- det(sig1_mle) * (sum(diag(sig1_mle))/rho)^(-rho)

test_stat <- -(n - 1) * 2 * log(Vn)

pchisq(q = test_stat, df = f, lower.tail = FALSE)

xbar <- apply(rtnorm_draws, 2:4, mean)

d <- 1

tot_sum <- 0

for(i in 1:1e3) {
  centered <- matricization(rtnorm_draws[[i]] - xbar, k = d)
  
  tot_sum <- tot_sum + centered %*% t(centered)
}

tot_sum/(1e3)

tot_sum/sum(diag(tot_sum))
```




```{r, eval = FALSE, include = FALSE}
rand_skew <- array(rnorm(24, mean = 2), dim = c(2, 3, 4))

full_covar <- kronecker(s3, s2) |> kronecker(s1)
inv_covar <- chol2inv(chol(full_covar))

res <- rep(0, 1e3)

for(i in 1:1e3) {
  curr_draw <- rtnorm_draws[i, , ,]
  
  diff <- c(curr_draw - rand_mu)
  
  res[i] <- t(diff) %*% inv_covar %*% diff
}

cutoff <- qchisq(p = 0.95, df = 24)
sum(res > cutoff)

ggplot() +
  geom_histogram(aes(x = res, y = after_stat(density)), 
                 fill = "pink", color = "black") +
  geom_function(fun = dchisq, args = list(df = 24), color = "red") +
  geom_vline(aes(xintercept = cutoff), color = "purple") +
  labs(x = "Vectorized Draws", y = "Density") +
  theme_minimal()
```


```{r, eval = FALSE, include = FALSE}
#rtskewt_draws <- rtskewt(n = 1e3, mu = rand_mu, sigmas = list(s1, s2, s3),
                         #skew = rand_skew, nu = 4)

# full_covar <- kronecker(mle_estimates$sigmas[[3]],
#                         mle_estimates$sigmas[[2]]) |>
#               kronecker(mle_estimates$sigmas[[1]])


# full_covar[1:10]
# full_covar_true[1:10]

#full_covar <- full_covar/sum(diag(full_covar))

# test_norm <- function(draws) {
#   mle_estimates <- mle_est(draws)
#   
#   sample_size <- dim(draws)[1]
#   num_dim <- length(dim(draws)) - 1
#   
#   resids <- rep(0, sample_size)
#   
#   for(i in 1:sample_size) {
#     centered <- draws[i, , ,] - mle_estimates$mu
#     
#       for(d in 1:3) {
#         S_half_inv <- chol2inv(chol(mle_estimates$sigmas[[d]]))
#         centered <- n_prod(centered, n = d, mat = S_half_inv)
#       }
#     resids[i] <- sum(centered^2)
#   }
# }

# 
# for(i in 1:sample_size) {
#   centered <- draws[i, , ,] - mu
# 
#   # Apply whitening: multiply by Sigma_d^{-1/2}
#   for(d in 1:3) {
#     S_half_inv <- chol2inv(chol(sigmas[[d]]))   # Σ_d^{-1/2} Σ_d^{-1/2}ᵀ = Σ_d^{-1}
#     centered <- n_prod(centered, n = d, mat = S_half_inv)
#   }
# 
#   res[i] <- sum(centered^2)   # equivalent to diffᵀ Σ^{-1} diff
# }





res2 <- rep(0, 1e3)

for(i in 1:1e3) {
  curr_draw <- rtnorm_draws[i, , ,]
  
  diff <- c(curr_draw - rand_mu)
  
  res2[i] <- t(diff) %*% inv_covar %*% diff
}

ggplot() +
  geom_histogram(aes(x = res2, y = after_stat(density)), 
                 fill = "pink", color = "black") +
  geom_function(fun = dchisq, args = list(df = 24), color = "red") +
  labs(x = "Vectorized Draws", y = "Density") +
  theme_minimal() 

vec <- c(rtnorm_draws)
```
