---
title: 'test_rtnorm'
output: pdf_document
---

# Multivariate Normal Tests

## Mardia's test

This test is based on skewness and kurtosis. Given we observe $k$ samples of $X_{1}, X_{2}, \dots, X_{k}$ we compute 

\begin{align*}
\hat{\Sigma} &= \frac{1}{n} \sum_{j=1}^{n} (x_{j} - \bar{x})(x_{j} - \bar{x})^{T} \\
A &= \frac{1}{6n} \sum_{i=1}^{n} \sum_{j=1}^{n} \left[(x_{i} - \bar{x}) \hat{\Sigma}^{-1} (x_{j} - \bar{x})^{T}\right]^{3} \\
B &= \sqrt{\frac{n}{8k(k+2)}} \left\{\frac{1}{n} \sum_{i=1}^{n} \left[(x_{i} - \bar{x})^{T} \hat{\Sigma}^{-1} (x_{i} - \bar{x})\right]^{2}\right\}
\end{align*}

Under $H_0: X \sim N_{p}(\mu, \sigma), A \sim \chi^2_{1/6 k (k+1) (k+2)}$ and $B \sim N(0, 1)$. 

```{r}
MVN::mardia(iris[1:50, 1:4])

multivariate_diagnostic_plot(iris[1:50, 1:4])
```



# Matrix Variate Normal Tests

## Malahanobis Distance

```{r}
s1 <- diag(2); s2 <- matrix(rnorm(9), nrow = 3); s3 <- matrix(rnorm(16), nrow = 4)
s1 <- 4 * s1
s2 <- crossprod(s2, s2); s3 <- crossprod(s3, s3)
list_sigmas <- list(s1, s2, s3)
rand_mu <- array(rnorm(24, mean = 10), dim = c(2, 3, 4))
rtnorm_draws <- rtnorm(n = 1e3, mu = rand_mu, sigmas = list(s1, s2, s3))

flat_dim <- prod(dim(rtnorm_draws)[-1])

draws_vec <- 
  matrix(rtnorm_draws, 
         nrow = dim(rtnorm_draws)[1],
         ncol = flat_dim)

mu_vec <- apply(draws_vec, MARGIN = 2, FUN = mean)

sig_mat <- matrix(0, nrow = flat_dim, ncol = flat_dim)

for(i in 1:flat_dim) {
  sig_mat[i, i] <- sum((draws_vec[i, ] - mu_vec) %*% t(draws_vec[i, ] - mu_vec))
}

sig_mat <- sig_mat * 1/(n-1)


```


```{r}
dim(sig_mat)

#24 x 1000

all <- (draws_vec - mu_vec) %*% chol(sig_mat) %*% t(draws_vec - mu_vec)

all
```


## Likelihood Ratio Tests for Sigma

Using the paper Likelihood Ratio Tests for High-Dimensional Normal Distributions by Jiang and Qi (2015), they suggest a spherical test. For a multivariate normal distribution $N_{p}(\mathbf{\mu}, \mathbf{\Sigma})$ where $\mathbf{\mu} \in \mathbb{R}^{p}$ is a mean vector and $\mathbf{\Sigma}$ is a $p \times p$ covariance matrix, we consider this spherical test. 

$$H_{0}: \mathbf{\Sigma} = \lambda \textbf{I}_{p} \text{ vs } H_{a}: \mathbf{\Sigma} \neq \lambda \textbf{I}_{p}$$

with unspecified $\lambda$. 

Suppose $\textbf{x}_{1}, \dots, \textbf{x}_{n}$ are iid $\mathbb{R}^{p}$ from the normal distribution. As usual, set

$$\bar{\textbf{x}} = \frac{1}{n} \sum_{i=1}^n x_i \text{ and } \textbf{S} = \frac{1}{n} \sum_{i=1}^{n} (\textbf{x}_{i} - \bar{\textbf{x}})(\textbf{x}_{i} - \bar{\textbf{x}})^{T}.$$ 

Then the LRT statistic for this test is 

$$V_{n} = |\textbf{S}| \left(\frac{\text{tr}(\textbf{S})}{p}\right)^{-p}.$$

Then $-(n-1) \rho \log{V_{n}}$ converges to $\chi^{2}_{f}$ as $n \to \infty$ where

$$\rho = 1 - \frac{2p^2 + p + 2}{6(n-1) * p} \text{ and } f = \frac{1}{2} (p-1)(p+2).$$

```{r}
test_sigmas <- function(est_sig) {
  p_vals <- rep(0, length(est_sig))
  
  for(d in 1:length(est_sig)) {
    curr_sig <- est_sig[[d]]
    p <- dim(curr_sig)[[1]]
    
    rho <- 1 - (2 * p^2 + p + 2)/(6 * (n-1) * p)
    f <- 1/2 * (p - 1) * (p + 2)
    
    Vn <- det(curr_sig) * (sum(diag(curr_sig))/p)^(-p)
    
    test_stat <- -(n - 1) * rho * log(Vn)
  
    p_vals[d] <- pchisq(q = test_stat, df = f, lower.tail = FALSE)
  }
  
  p_vals
}

n <- 1e3
```


Test with sigma 1 being diagonal. Since the null is that the sigma is diagonal, we expect to fail to reject for the first sigma. 
```{r}
s1 <- diag(2); s2 <- matrix(rnorm(9), nrow = 3); s3 <- matrix(rnorm(16), nrow = 4)
s1 <- 4 * s1
s2 <- crossprod(s2, s2); s3 <- crossprod(s3, s3)
list_sigmas <- list(s1, s2, s3)
rand_mu <- array(rnorm(24, mean = 10), dim = c(2, 3, 4))
rtnorm_draws <- rtnorm(n = n, mu = rand_mu, sigmas = list(s1, s2, s3))

est_mle <- mle_est(rtnorm_draws)

test_sigmas(est_mle$sigmas)
```


Test with sigma 1, 2, 3 being diagonal
```{r}
s1 <- diag(2); s2 <- diag(3); s3 <- diag(4)
s1 <- 4 * s1
s2 <- 2 * s2
s3 <- 8 * s3
rand_mu <- array(rnorm(24, mean = 10), dim = c(2, 3, 4))
rtnorm_draws <- rtnorm(n = n, mu = rand_mu, 
                       sigmas = list(s1, s2, s3))

est_mle <- mle_est(rtnorm_draws)
test_sigmas(est_mle$sigmas)
```


Test with sigma 2 being diagonal with some error
```{r}
s1 <- diag(2); s2 <- diag(3); s3 <- diag(4)
s1 <- 4 * s1
s2 <- 2 * s2 + rnorm(9, mean = 0.4, sd = 0.2)
s3 <- 8 * s3
rand_mu <- array(rnorm(24, mean = 10), dim = c(2, 3, 4))
rtnorm_draws <- rtnorm(n = n, mu = rand_mu, 
                       sigmas = list(s1, s2, s3))

est_mle <- mle_est(rtnorm_draws)
test_sigmas(est_mle$sigmas)
```

Test with sigma 1 and sigma 3 being diagonal
```{r}
s1 <- diag(2); s2 <- diag(3); s3 <- 3 * diag(4)
s1 <- 4 * s1
s2 <- 2 * s2 + rnorm(9, mean = 0.4, sd = 0.2)
s3 <- 8 * s3
rand_mu <- array(rnorm(24, mean = 10), dim = c(2, 3, 4))
rtnorm_draws <- rtnorm(n = n, mu = rand_mu, 
                       sigmas = list(s1, s2, s3))

est_mle <- mle_est(rtnorm_draws)
test_sigmas(est_mle$sigmas)
```

Test the power when sigma 1 and sigma 3 are diagonal

```{r}
rt_power <- tibble("p1" = rep(0, 1000), "p2" = rep(0, 1000), "p3" = rep(0, 1000))

for(i in 1:100) {
  rtnorm_draws <- rtnorm(n = n, mu = rand_mu, 
                         sigmas = list(s1, s2, s3))
  
  est_mle <- mle_est(rtnorm_draws)
  
  list_mles[[i]] <- est_mle$sigmas
  
  curr_test <- test_sigmas(est_mle$sigmas)
  
  rt_power[i, "p1"] <- curr_test[1]
  rt_power[i, "p2"] <- curr_test[2]
  rt_power[i, "p3"] <- curr_test[3]
}

rt_power$p2[1:100] |> 
  as_tibble() |> 
  ggplot() +
  geom_histogram(aes(x = value))
```



```{r}
est_mle <- mle_est(rtnorm_draws)

est_mle$sigmas
```


```{r}
sig1_mle <- est_mle$sigmas[[2]]

n <- 1e3

rho <- dim(sig1_mle)[[1]]

f <- 1/2 * (rho - 1) * (rho + 2)

Vn <- det(sig1_mle) * (sum(diag(sig1_mle))/rho)^(-rho)

test_stat <- -(n - 1) * 2 * log(Vn)

pchisq(q = test_stat, df = f, lower.tail = FALSE)

xbar <- apply(rtnorm_draws, 2:4, mean)

d <- 1

tot_sum <- 0

for(i in 1:1e3) {
  centered <- matricization(rtnorm_draws[[i]] - xbar, k = d)
  
  tot_sum <- tot_sum + centered %*% t(centered)
}

tot_sum/(1e3)

tot_sum/sum(diag(tot_sum))
```




```{r}
rand_skew <- array(rnorm(24, mean = 2), dim = c(2, 3, 4))

full_covar <- kronecker(s3, s2) |> kronecker(s1)
inv_covar <- chol2inv(chol(full_covar))

res <- rep(0, 1e3)

for(i in 1:1e3) {
  curr_draw <- rtnorm_draws[i, , ,]
  
  diff <- c(curr_draw - rand_mu)
  
  res[i] <- t(diff) %*% inv_covar %*% diff
}

cutoff <- qchisq(p = 0.95, df = 24)
sum(res > cutoff)

ggplot() +
  geom_histogram(aes(x = res, y = after_stat(density)), 
                 fill = "pink", color = "black") +
  geom_function(fun = dchisq, args = list(df = 24), color = "red") +
  geom_vline(aes(xintercept = cutoff), color = "purple") +
  labs(x = "Vectorized Draws", y = "Density") +
  theme_minimal()
```


```{r}
#rtskewt_draws <- rtskewt(n = 1e3, mu = rand_mu, sigmas = list(s1, s2, s3),
                         #skew = rand_skew, nu = 4)

# full_covar <- kronecker(mle_estimates$sigmas[[3]],
#                         mle_estimates$sigmas[[2]]) |>
#               kronecker(mle_estimates$sigmas[[1]])


# full_covar[1:10]
# full_covar_true[1:10]

#full_covar <- full_covar/sum(diag(full_covar))

# test_norm <- function(draws) {
#   mle_estimates <- mle_est(draws)
#   
#   sample_size <- dim(draws)[1]
#   num_dim <- length(dim(draws)) - 1
#   
#   resids <- rep(0, sample_size)
#   
#   for(i in 1:sample_size) {
#     centered <- draws[i, , ,] - mle_estimates$mu
#     
#       for(d in 1:3) {
#         S_half_inv <- chol2inv(chol(mle_estimates$sigmas[[d]]))
#         centered <- n_prod(centered, S_half_inv, n = d)
#       }
#     resids[i] <- sum(centered^2)
#   }
# }

# 
# for(i in 1:sample_size) {
#   centered <- draws[i, , ,] - mu
# 
#   # Apply whitening: multiply by Sigma_d^{-1/2}
#   for(d in 1:3) {
#     S_half_inv <- chol2inv(chol(sigmas[[d]]))   # Σ_d^{-1/2} Σ_d^{-1/2}ᵀ = Σ_d^{-1}
#     centered <- n_prod(centered, S_half_inv, n = d)
#   }
# 
#   res[i] <- sum(centered^2)   # equivalent to diffᵀ Σ^{-1} diff
# }





res2 <- rep(0, 1e3)

for(i in 1:1e3) {
  curr_draw <- rtnorm_draws[i, , ,]
  
  diff <- c(curr_draw - rand_mu)
  
  res2[i] <- t(diff) %*% inv_covar %*% diff
}

ggplot() +
  geom_histogram(aes(x = res2, y = after_stat(density)), 
                 fill = "pink", color = "black") +
  geom_function(fun = dchisq, args = list(df = 24), color = "red") +
  labs(x = "Vectorized Draws", y = "Density") +
  theme_minimal() 

vec <- c(rtnorm_draws)
```

