---
title: "tensormodels: Statistical Models for Tensor-Valued Data"
date: "2026-06-18"
abstract: >
  Replace this text with a plain-text abstract of no more than 250 words.
  Briefly describe the problem addressed by tensormodels, the package's main
  contributions, and the practical benefit to R users.
draft: true
author:
  - name: "Your Name"
    affiliation: "Your Institution"
    address:
      - "Department or unit"
      - "City, State, Country"
    email: "you@example.org"
    orcid: "0000-0000-0000-0000"
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: tensormodels.bib
packages:
  cran: []
  bioc: []
CTV: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4
)

library(tensormodels)
```

# Introduction



Explain the statistical or computational problem that motivates
\pkg{tensormodels}. State who the package is for and summarize its main
contributions. End this section with a short roadmap of the paper.

# Background and related work

Introduce the tensor notation and modeling ideas needed by the reader. Compare
the package with the closest existing R software and cite relevant literature
using Pandoc citations, for example [@r-base].

# Package design

The main goal of tensortools is to provide functions for random number generation, density computation, and MLE estimation for common tensor variate distributions. Additionally, the package provides diagnostic tests for the structure of input data. 


One intentional decision was to store independent, identically distributed draws from a tensor variate distribution as a list. For example, if we generate $n = 100$ draws from a tensor variate normal distribution of dimension $2 \times 3 \ times 4$, the resulting R object will be a list of length 100. Each element in the list is a base R array object of dimension $2 \times 3 \times 4$. Other packages in this space created unique S3 and S4 class objects to store these types of draws. However, these choices are difficult for the average R user, as they have unique interactions and functions. 

Another option would be to output a tensor of size $100 \times 2 \times 3 \times 4$; that is, create a tensor of one higher dimension where the first mode represents the index of the draw. The choice to have the first mode rpresent the draw is arbitrary and could potentially confuse users, especially if they permute the dimensions. 

```{r}
three_dim_iid <- rtnorm(n = 100, mu = array(1:12, dim = c(2, 3, 4)))

length(three_dim_iid)
str(three_dim_iid[1:2], max.level = 1)
```


Describe the package's data structures, model classes, and computational
approach. Emphasize design decisions that matter to an R user.

## Core workflow

Show the smallest useful end-to-end example. Keep examples reproducible and
fast enough that the full article can be rebuilt in under ten minutes.

```{r basic-workflow, eval=FALSE}
library(tensormodels)

# Replace with a compact, reproducible example:
# fit <- model_function(data, ...)
# summary(fit)
```

# Model fitting and inference

Explain estimation, initialization, convergence checks, uncertainty
quantification, and any constraints or identifiability conventions.

# Examples

Use one or two examples that demonstrate what the package makes possible.
Prefer datasets that can be included with the submission or downloaded from a
stable public archive.

```{r example-figure, eval=FALSE, fig.cap="Replace with an informative caption.", fig.alt="Replace with a concise description of the figure for readers using assistive technology."}
# Plotting code goes here.
```

# Performance and validation

Report numerical validation, simulation results, benchmarks, or comparisons
with established implementations. Clearly distinguish package features from
claims supported by experiments.

# Example

Now that we have set up our tensor distributions, we can use them on a real-world example. The CIFAR 100 dataset contains images of a set of 100 common objects. In this case, we will only be using the maple trees, which have an ID value of 48.

```{r}
library(torchvision)
library(patchwork)
library(tidyverse)

plot_image <- function(img) {
  rgb_image <-
    expand_grid(x = seq_len(dim(img)[1]),
                y = seq_len(dim(img)[2])) |>
    mutate(r = as.vector(img[, , 1]), 
           g = as.vector(img[, , 2]),
           b = as.vector(img[, , 3]),
           fill = rgb(r, g, b))
  
  ggplot(rgb_image, aes(x = x, y = y, fill = fill)) +
    geom_raster(interpolate = TRUE) +
    scale_fill_identity() +
    coord_fixed(expand = FALSE) +
    scale_y_reverse() +
    theme_void()
}

plot_matrix <- function(input) { # function to plot matrices
  input |> 
    as_tibble(.name_repair = ~ paste0("V", seq_along(.))) |> 
    mutate(row = row_number()) |> 
    pivot_longer(-row, names_to = "col", values_to = "value") |> 
    mutate(col = as.integer(str_remove(col, "^V"))) |> 
    ggplot(aes(x = col, y = row, fill = value)) +
    geom_tile(color = "black", lwd = 0.1) +
    scale_y_reverse() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) 
}
```


Once we have the CIFAR dataset loaded in from torchvision, we can plot 20 example images. The images are stored as an array of size 32 x 32 x 3. This is an image that is 32 x 32 pixels, and the last mode represents the red, green, and blue color values. 

his is an image that is 32 x 32 pixels stored as red, green, and blue values. The values vary from 0 to 255 and represent the intensity of that color. 
```{r}
cifar100 <- cifar100_dataset(download = TRUE)

maple_idx <- which(cifar100$y == 48)

maple_trees <- vector("list", length(maple_idx))

# store as a list of images
for (i in seq_along(maple_idx)) {
  maple_trees[[i]] <- cifar100[[maple_idx[i]]]$x/255
}

wrap_plots(lapply(maple_trees[1:20], plot_image), ncol = 5)
```


We can fit the best model to this set of images. The function `model_compare()`{.R} takes in a dataset and fits all of the possible tensor models. It returns the result as a data frame with columns containing the name of the model, log likelihood, number of parameters denoted as k, AIC, BIC, and the model object itself. We can then select the model with the lowest BIC to proceed. 
```{r}
model_compare <- function(draws) {
  res_normal <- tensor_mle(draws, model = "normal", quiet = FALSE)
  res_skewt <- tensor_mle(draws, model = "skewt", quiet = FALSE)
  res_vargam <- tensor_mle(draws, model = "vargamma", quiet = FALSE)
  res_invgauss <- tensor_mle(draws, model = "invgauss", quiet = FALSE)
  res_genhyper <- tensor_mle(draws, model = "genhyper", quiet = FALSE)

  all_mod <- 
    tibble(model = c("Normal", "Skewt", "Vargamma", "Invgauss", "Genhyper"),
    loglik = c(res_normal$loglik, res_skewt$loglik,
               res_vargam$loglik, res_invgauss$loglik, res_genhyper$loglik),
    k = c(res_normal$k, res_skewt$k,
          res_vargam$k, res_invgauss$k, res_genhyper$k),
    AIC = c(res_normal$AIC, res_skewt$AIC,
            res_vargam$AIC, res_invgauss$AIC, res_genhyper$AIC),
    BIC = c(res_normal$BIC, res_skewt$BIC,
            res_vargam$BIC, res_invgauss$BIC, res_genhyper$BIC), 
    mod_obj = list(res_normal, res_skewt, res_vargam, res_invgauss, res_genhyper)
  )
  
  all_mod
}

res <- model_compare(maple_trees)

res |> select(model, loglik, BIC)

best_mod <- res$mod_obj[[which.min(res$BIC)]]
```


In this case, the best model was the normal inverse Gaussian. The expected value of its draws is $\mathcal{M} + \mathcal{A}/\kappa$, where $\mathcal{M}$ is the location parameter and $\mathcal{A}$ is the skew parameter. The plot below contains three images: the first is the empirical average of all of the maple trees without a model assumption. The second is the location parameter $\mathcal{M}$ from the model. Lastly, we have $E[X]$, which is the expected average from the model. 

$\mathcal{M}$ is closer to a traditional green maple tree. This seems to be a good starting point to model the variety of colors of the maples, from green to red. Once we add the skew parameter, the $E[X]$ appears similar to the empirical average.. 
```{r}
plot_image(apply(simplify2array(maple_trees), 1:3, mean)) + 
  labs(title = "Avg")+
plot_image(best_mod$mu) +
  labs(title = "M") +
plot_image(with(best_mod, mu + skew/kappa)) +
  labs(title = "E[X]")
```


We can also create summary plots of the skew $\mathcal{A}$ and the covariance matrices $\Sigma$ from the model. We converted the covariance matrices to correlation matrices to make it easier to interpret each matrix, since this places them on the same scale. 
```{r}
plot_matrix(best_mod$skew[, , 1]) +
  plot_matrix(best_mod$skew[, , 2]) +
  plot_matrix(best_mod$skew[, , 3]) + 
  plot_layout(guides = 'collect') &
  scale_fill_gradient2(#low = "blue", mid = "white", high = "red",
    midpoint = 0,
    limits = c(min(best_mod$skew), max(best_mod$skew)))

sig_corr <- lapply(best_mod$sigmas, cov2cor)

plot_matrix(sig_corr[[1]]) +
  plot_matrix(sig_corr[[2]]) +
  plot_matrix(sig_corr[[3]]) +
  plot_layout(guides = "collect") &
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-1, 1))
```



Now that we have the best model for the maple trees, we can obtain the likelihood of each image. 
```{r}
all_lik <- vector(mode = "numeric", length = 500)

for(i in 1:500) {
  all_lik[i] <- dtinvgauss(x = maple_trees[[i]], mu = best_mod$mu,
                           skew = best_mod$skew, sigmas = best_mod$sigmas, 
                           kappa = best_mod$kappa, log = TRUE)
}

all_lik |> 
  as_tibble() |> 
  ggplot() +
  geom_histogram(aes(x = value), color = "black", fill = "lightpink",
                 bins = 40) +
  theme_minimal() +
  labs(x = "Likelihood", y = "Count")
```


There are clearly some outliers, both at much smaller and much larger likelihoods. Out of curiosity, I plotted the lowest likelihoods. All of these maples are predominantly orange and red, which differ from the average tensor that the model had. It seems that the data contains many more green maples, so the model struggles to account for the possible variation in tree color. 
```{r}
wrap_plots(lapply(maple_trees[which(all_lik <= 4500)], plot_image), 
           ncol = 5)
```


As expected, all of the maples that had a large likelihood follow the pattern of being a forest green and mainly in the center of the image. 
```{r}
wrap_plots(lapply(maple_trees[which(all_lik >= 9000)], plot_image), 
           ncol = 5)
```



```{r, cache=TRUE}
library(ggplot2)
library(dplyr)
library(tibble)
library(ggimage)

# sort by likelihood and select 30 evenly spaced observations
lik_order <- order(all_lik)
selected_ranks <- unique(round(seq(from = 1, to = length(lik_order), length.out = 30)))

selected_ids <- lik_order[selected_ranks]

maple_plot_data <- 
  tibble(id = selected_ids, log_likelihood = all_lik[selected_ids]) |>
  arrange(log_likelihood) |>
  mutate(row = rep(1:3, length.out = n()))

thumbnail_dir <- "paper/figures/maple-thumbnails"
dir.create(thumbnail_dir, recursive = TRUE, showWarnings = FALSE)

maple_plot_data$image <- file.path(
  thumbnail_dir,
  paste0("maple-", maple_plot_data$id, ".png")
)

for (i in seq_len(nrow(maple_plot_data))) {
  tree_id <- maple_plot_data$id[i]

  ggsave(filename = maple_plot_data$image[i],
         plot = plot_image(maple_trees[[tree_id]]),
         width = 1, height = 1,
         units = "in", dpi = 150)
}

ggplot(maple_plot_data, 
       aes(x = log_likelihood, y = row,
       image = image)) +
  geom_image(size = 0.09, by = "width") +
  scale_y_continuous(limits = c(0.5, 3.5), breaks = NULL) +
  labs(x = "Log-likelihood",
       y = NULL) +
  theme_minimal() 
```


# Discussion

Summarize the package's strengths, limitations, and intended scope. Mention
important future work without turning this section into a feature wish list.

# Reproducibility

List any external data and scripts required to reproduce the article. Place
supporting code in `scripts/`, data in `data/`, and figure assets in `figures/`.

```{r session-info, echo=FALSE}
cat("R version:", as.character(getRversion()), "\n")
cat("Platform:", R.version$platform, "\n")
cat("rjtools version:", as.character(packageVersion("rjtools")), "\n")
```

# Acknowledgments

Add funding, contributor, and institutional acknowledgments here.
